---
title: "Questions → Data"
subtitle: "Teaching a computer what we mean"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis-fonts", "css/ds1000.css"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
.hero-image {
  display: flex;
  align-items: center;
  justify-content: center;
  height: 74vh;
}
.hero-image img {
  max-width: 92%;
  max-height: 74vh;
  object-fit: contain;
}

</style>

```{css,echo=FALSE}
.small .remark-code { font-size: 85% !important; }
.tiny .remark-code { font-size: 60% !important; }
.big { font-size: 150%; }
.huge { font-size: 200%; }
```

```{r,include=FALSE}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=FALSE,warning=FALSE)
library(tidyverse)

df <- tryCatch(
  read_rds("../data/sc_debt.Rds"),
  error = function(e) tryCatch(
    read_rds("data/sc_debt.Rds"),
    error = function(e2) NULL
  )
)
```

# Last time

We started with a claim:

> **Someone tells you something is true. What evidence would let you decide?**

--

And we broke research problems into pieces:

**Question → Concepts → Variables → Unit of analysis → Understand variables → Put variables together → Evaluate → Communicate**

---

# Today

.big[How do the things we care about become **data**?]

--

By the end of class, you should be able to:

1. identify the **unit of analysis** in a dataset;
2. connect substantive **concepts** to measured **variables**;
3. read a simple tabular dataset;
4. read simple R/tidyverse instructions;
5. decide whether code actually implements the analytical task we intended.

<!-- INSTRUCTOR NOTE — LO2 is primary today; LO4 and LO5 begin at a basic level. -->

---

# Our running claim

> **More selective colleges produce graduates who earn more.**

--

Before touching R:

- What concepts are in this claim?
- What would we need to measure?
- What would one observation represent?

--

Homework 1 asked you to begin answering these questions.

---

# Concept ≠ variable

### Concept

An abstract idea we care about.

Examples:

- college selectivity
- graduate earnings
- political ideology
- health

--

### Variable

A recorded measurement intended to capture some aspect of a concept.

---

# The measurement problem

Suppose we care about:

.big[**College selectivity**]

--

How might we measure it?

- admissions rate?
- average SAT score?
- applicant qualifications?
- institutional ranking?

--

These are related ideas.

They are **not the same thing**.

---

# First principle

.center[.huge[
**The computer only sees the variable.**
]]

--

It does not know what you *meant* by the variable.

--

That judgment belongs to you.

<!-- INSTRUCTOR NOTE — Early construct-validity preview. Lecture 3 will develop this much more fully. -->

---

# From the world to a table

A dataset is a structured representation of observations.

--

In the kind of tabular data we will use most often:

> **Rows are observations.**

> **Columns are variables.**

---

# The most important question in a dataset

.center[.huge[
**What does one row represent?**
]]

--

A person?

A college?

A country?

A tweet?

A person on a particular day?

--

This is the **unit of analysis**.

---

# Why the unit of analysis matters

Imagine a dataset with:

- `student_id`
- `college`
- `earnings`

--

If one row is a **student**, we can ask about differences among people.

--

If one row is a **college**, we can ask about differences among institutions.

--

Same topic. Different evidence.

---

# College Scorecard data

For our running example, we will use a subset of the U.S. College Scorecard.

--

.big[What should one row represent?]

--

**One college / institution.**

---

# Now open RStudio

Today R is not the subject.

.big[R is how we inspect and manipulate the evidence.]

--

You need enough R to:

- understand what an instruction is doing;
- modify simple instructions;
- recognize when an instruction does the wrong thing;
- ask AI for help without surrendering judgment.

---

# The minimum R vocabulary

**Object:** Something we have given a name.

**Function:** An instruction that does something.

**Argument:** Information we give the function.

**Pipe (`%>%`):** Take what we have and do something next.

---

# Objects: giving something a name

```{r}
claim <- "More selective colleges produce graduates who earn more"
claim
```

--

`<-` means:

> **Store the thing on the right under the name on the left.**

--

You do not need to love the syntax.

You need to be able to read the instruction.

---

# Functions: giving an instruction

```{r}
mean(c(2, 4, 6))
```

--

Read it as:

> Use the function `mean()` on the values 2, 4, and 6.

--

Functions are verbs.

Arguments tell the verb what to act on or how to act.

---

# Load the data

```{r,eval=FALSE}
df <- read_rds("data/sc_debt.Rds")
```

--

Read this in English:

> Read the College Scorecard file and store it as an object called `df`.

--

That is enough for now.

---

# Inspect before manipulating

```{r,eval=!is.null(df)}
df
```

---

# Inspect before manipulating

Do not immediately start changing the data.

First ask:

1. What does one row represent?
2. What variables are available?
3. Which variables might bear on our question?

---

# What variables are available?

```{r,eval=!is.null(df)}
names(df)
```

--

Names are not meanings.

--

We need a **data dictionary**.

---

# A few College Scorecard variables

| Variable | Recorded meaning |
|---|---|
| `instnm` | Institution name |
| `adm_rate` | Proportion of applicants admitted |
| `sat_avg` | Average SAT score |
| `md_earn_wne_p6` | Earnings of recent graduates |
| `grad_debt_mdn` | Median graduate debt |
| `ugds` | Number of undergraduates |
| `stabbr` | State |

--

.big[Which variables bear most directly on our claim?]

---

# Operationalizing the claim

> **More selective colleges produce graduates who earn more.**

--

Possible operationalization:

**Selectivity** → `adm_rate`

**Graduate earnings** → `md_earn_wne_p6`

--

But notice what just happened.

We made **substantive choices** before writing analytical code.

---

# Does `adm_rate` measure selectivity?

It measures:

> the proportion of applicants who are admitted.

--

That seems useful.

But ask:

- Who applies in the first place?
- Are application pools comparable across colleges?
- Could two schools with the same admissions rate differ in meaningful ways?

--

.big[Variables are evidence about concepts, not the concepts themselves.]

---

# Does the earnings variable mean "college value"?

Not necessarily.

Graduate earnings may reflect:

- what the college did;
- who enrolled;
- field of study;
- geography;
- family background;
- labor-market conditions;
- many other things.

--

This is why good analysis begins **before** the model.

---

# Which observations belong in the analysis?

Suppose we want to look only at Vanderbilt.

Substantive instruction:

> Keep the observation whose institution name is Vanderbilt University.

--

R instruction:

```{r,eval=!is.null(df)}
df %>%
  filter(instnm == "Vanderbilt University")
```

---

# `filter()` answers a substantive question

.center[.huge[
**Which observations do I want?**
]]

--

`filter()` keeps **rows** that satisfy a condition.

--

The important decision is not remembering the word `filter`.

The important decision is knowing **which observations belong in the analysis and why**.

---

# Which variables belong in the analysis?

Suppose we want only the pieces most relevant to the claim.

`select()` keeps **columns**.


```{r,eval=!is.null(df)}
df %>%
  select(instnm, adm_rate, md_earn_wne_p6)
```


---

# `select()` answers another substantive question

.center[.huge[
**Which variables do I need?**
]]

--

Again, the syntax is secondary.

The analytical judgment comes first.

---

# The pipe

```{r,eval=!is.null(df)}
df %>%
  filter(stabbr == "TN") %>%
  select(instnm, adm_rate, md_earn_wne_p6)
```

---

# The pipe

Read `%>%` as:

> **and then...**

--

Start with `df`

**and then** keep Tennessee colleges

**and then** keep these variables.

---

# Computational thinking

The code is just a translation of a sequence of decisions:

1. Start with the College Scorecard data.
2. Decide which observations matter.
3. Decide which variables matter.
4. Keep those pieces.

--

.big[The hard part is specifying the task clearly.]

---

# AI can write this code

Suppose you tell an AI assistant:

> "Using `df`, keep only Tennessee colleges and retain institution name, admissions rate, and graduate earnings. Use tidyverse."

--

It might produce:

```{r,eval=FALSE}
df %>%
  filter(stabbr == "TN") %>%
  select(instnm, adm_rate, md_earn_wne_p6)
```

--

That is useful.

But who decided what `TN`, `adm_rate`, and `md_earn_wne_p6` meant?

---

# AI can also be confidently wrong

Imagine AI returns:

```{r,eval=FALSE}
df %>%
  filter(instnm == "TN") %>%
  select(stabbr, sat_avg, grad_debt_mdn)
```

--

.big[What is wrong?]

--

The code is valid R.

It is a bad implementation of the analytical task.

<!-- INSTRUCTOR NOTE — LO5. Distinguish syntactic validity from substantive validity. -->

---

# The standard for AI-generated code

Do not ask only:

> **Does the code run?**

--

Ask:

> **Does the code do what I intended?**

> **Do the variables mean what I think they mean?**

> **Did it keep the observations I intended to study?**

--

A program can run perfectly and answer the wrong question.

---

# How do we verify code?

Use simple checks.

If we asked for Tennessee colleges:

```{r,eval=!is.null(df)}
df %>%
  filter(stabbr == "TN") %>%
  select(instnm, stabbr) %>%
  head()
```

--

Verification means comparing the output against the task we specified.

---

# A quick practice problem

Claim:

> **Students at larger colleges graduate with more debt.**

Available variables include:

- `ugds`: number of undergraduates
- `grad_debt_mdn`: median graduate debt
- `adm_rate`: admissions rate
- `stabbr`: state

--

Before writing code:

1. What are the concepts?
2. What is the unit of analysis?
3. Which variables would you retain?

---

# Now evaluate the instruction

An AI assistant proposes:

```{r,eval=FALSE}
df %>%
  select(instnm, ugds, grad_debt_mdn)
```

--

Is this a reasonable **first step**?

--

Yes.

But it has not answered the research question.

It has only prepared relevant evidence for inspection.

---

# What should happen next?

We now have candidate variables:

- `ugds`
- `grad_debt_mdn`

--

Should we immediately put them together?

--

.center[.huge[
**No.**
]]

---

# Before asking how variables relate...

.center[.huge[
**Understand each variable.**
]]

--

For each one, ask:

- What does it measure?
- What type of variable is it?
- What values does it take?
- What does its distribution look like?
- Is anything missing or strange?

--

That is where we go next.

---

# The course roadmap

**Question**

↓

**Concepts → Variables → Unit of analysis**  ← *today*

↓

**Understand each variable**  ← *next*

↓

**Put variables together**

↓

**Evaluate → Communicate**

---

# What you should take away

1. A dataset is a representation of the world, not the world itself.
2. One row has to represent something specific.
3. Variables operationalize concepts imperfectly.
4. `filter()` means choosing observations; `select()` means choosing variables.
5. AI can translate clear analytical instructions into code.
6. **You remain responsible for whether those instructions and that code make sense.**

---

# Before next class

Work through `ds1000_hw_2_f2026.Rmd`.

--

You will return to the College Scorecard claim with the actual data.

--

The goal is not to memorize R syntax.

.big[The goal is to learn how questions become data.]
