---
title: "How Can Data and AI Systems Produce Misleading Claims?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

gms <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/game_summary.Rds")

# Biased-training-data demo: fit only on home games, evaluate on both.
train_home <- gms %>% filter(locationGame == "H")
test_home <- gms %>% filter(locationGame == "H")
test_away <- gms %>% filter(locationGame == "A")

model_home_only <- lm(isWin ~ tov + pts + treb + pctFG, data = train_home)

acc_on <- function(model, newdata) {
  preds <- predict(model, newdata = newdata)
  pred_class <- ifelse(preds >= 0.5, TRUE, FALSE)
  mean(pred_class == newdata$isWin, na.rm = TRUE)
}

acc_home <- acc_on(model_home_only, test_home)
acc_away <- acc_on(model_home_only, test_away)

# A tiny spurious-correlation demo, generated live rather than borrowed.
set.seed(42)
spurious_dat <- tibble(
  year = 2001:2020,
  series_a = cumsum(rnorm(20)),
  series_b = cumsum(rnorm(20))
)
spurious_cor <- cor(spurious_dat$series_a, spurious_dat$series_b)
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 19 makes explicit and systematic a habit the course has practiced
implicitly all semester: auditing how misleading claims get produced. Four
mechanisms, each demonstrated concretely rather than via case-study images
(keeping this lecture consistent with the course's text/code style):
(1) unrepresentative training data, demonstrated by retraining the
Classification unit's win-predictor on home games only; (2) confounded/
spurious relationships, via a live-generated spurious correlation plus a
callback to HW6's Simpson's paradox; (3) false precision from a measure
that sounds exact but isn't (callback to Lecture 18's sentiment score);
(4) AI hallucination -- fluency without verification.

INSTRUCTOR NOTE:
Avoid citing specific real-world statistics/studies that can't be verified
in this environment; keep real-world framing general and qualitative
(the mechanisms themselves are well-documented in algorithmic-bias and
misinformation research, without needing a specific unverified figure).
-->

# How Can Data and AI Systems Produce Misleading Claims?

.big[
All semester:

> **Audit the AI's claim. What's missing? What's overstated?**

Today:

> **Zoom out — how do misleading claims get produced in the first place?**
]

---

# Where We Are

.big[
**Question → ... → measure meaning, not just topic → recognize how misleading claims get produced**
]

Today:

**Recognize how misleading claims get produced**

---

# A Reframe

.big[
Misinformation does not require a villain.
]

Every mechanism today can happen with **no one intending to deceive anyone** — just by skipping a step this course has spent 18 lectures teaching you not to skip.

---

# Mechanism 1: Unrepresentative Data

.huge[
> **A model trained on one slice of the world, applied to the whole world.**
]

Let's demonstrate this directly, with a classifier we already built.

---

# Retraining on a Biased Slice

Recall Classification I's win-predictor. This time, train it **only on home games**.

```{r, eval=FALSE}
train_home <- gms %>% filter(locationGame == "H")
model_home_only <- lm(isWin ~ tov + pts + treb + pctFG, data = train_home)
```

Nothing about this looks malicious. It's just a modeling choice — maybe made for convenience, or without thinking about it.

---

# Evaluate on Both Kinds of Games

```{r}
round(acc_home, 3)
round(acc_away, 3)
```

.big[
Trained only on home games, does the model perform the same on away games as it does on the games like the ones it was trained on?
]

---

# Why This Matters Beyond Basketball

.big[
The mechanism here — a model trained on one group performs differently for a group it rarely saw — is the same mechanism behind well-documented problems in hiring tools, lending models, and facial recognition systems trained on unrepresentative data.
]

No one has to intend harm. The training data just has to be unrepresentative of who the model will be used on.

<!-- Deliberately general/qualitative framing -- no specific unverified statistic cited. -->

---

# The Output Still Looks Neutral

.big[
This is what makes the mechanism dangerous: the model's predictions don't come with a warning label.
]

A number is a number. Nothing about `predict()`'s output announces "this estimate is less trustworthy for this group."

---

# Mechanism 2: Confounded and Spurious Relationships

.big[
Recall HW6: ice-cream sales and drownings were associated in the pooled data, purely because both were driven by temperature.
]

Reported without the Unit 2 audit habit, that pooled relationship is a ready-made piece of misinformation.

---

# A Relationship That's Pure Coincidence

```{r}
spurious_dat %>%
  ggplot(aes(x = series_a, y = series_b)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

---

# The Correlation Is Real

```{r}
round(spurious_cor, 2)
```

.big[
`series_a` and `series_b` are two **independently generated random sequences**. They share no cause. The correlation is not a coding error — with enough random series, some pairs will look related by chance.
]

---

# The Lesson

.big[
A relationship being statistically real in the data is not the same as it being meaningful.
]

Misinformation frequently isn't a fabricated number — it's a real number, reported without the "what else could explain this?" step from Unit 2.

---

# Mechanism 3: False Precision

.big[
Recall Lecture 18: a "net sentiment score" sounds like a precise measurement of feeling. It is actually a count of dictionary-matched words.
]

A single clean number invites more trust than the underlying measurement deserves.

---

# Numbers That Sound More Certain Than They Are

.big[
"Toxicity score: 87%." "Risk score: 6.2." "Approval likelihood: 73%."
]

Each of these is a specific-sounding number built from an imperfect proxy, exactly like the sentiment score. The precision of the number does not imply precision in what it measures.

<!-- LO1: Reinforce construct validity from Unit 1 in a new register. -->

---

# Mechanism 4: AI Fluency Without Verification

.big[
A language model is trained to produce **plausible-sounding** text — not verified-true text.
]

It can generate a specific statistic, a citation, or a quote that **sounds exactly like the real thing** and is entirely fabricated. This is often called **hallucination**.

---

# Why This Happens

.big[
The model is optimizing for "what would a fluent answer look like," not "is this claim actually true."
]

Confidence in the writing and accuracy of the claim are two completely different things — and a fluent, confident tone provides no evidence about which one you're getting.

---

# Putting It Together: Audit a Real-Looking Claim

An AI-generated summary states:

> "A recent study found that 73% of remote workers report higher productivity, based on a nationally representative survey."

.big[
**Mechanism check — could this involve:**

Unrepresentative data? False precision? A confound the summary skipped? Simple fabrication?
]

You may not be able to tell which, from the sentence alone — which is itself the point.

---

# The Only Real Defense

.big[
Not "trust it" or "distrust it" — **verify it.**
]

Find the actual study. Check the sample. Check what "productivity" was operationalized as. Check whether the number even appears in the source.

---

# A Calibrated Habit

Instead of:

> "The AI said it, so it must be true (or must be fake)."

Try:

.big[
> **This claim has a specific, checkable structure — a source, a sample, and a measurement. Before repeating it, I should verify each piece rather than reacting to how confidently it's stated.**
]

---

# Before You Leave

When you see a data-driven or AI-generated claim, ask:

1. What data was the underlying model or statistic trained or calculated on — and who might be missing from it?
2. Could this relationship be confounded or coincidental?
3. Does a precise-looking number come from a proxy that isn't actually precise?
4. Could this specific claim be a fabrication dressed up as fluent, confident text?
5. Have I verified this, or just found it plausible?

---

# The Rule to Remember

.huge[
**Misinformation is rarely a single lie.**

**It is usually a skipped step.**
]

---

# Homework 15

You will practice:

- identifying unrepresentative training data in a real scenario;
- distinguishing a real-but-spurious relationship from a meaningful one;
- spotting false precision in a reported measure;
- auditing an AI-generated claim for fabrication, and explaining how you would verify it.

Next:

.big[
**AI, Judgment & Responsibility**
]
