---
title: "What Else Could Explain It?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

df <- read_rds("https://github.com/jbisbee1/DS1000_F2024/raw/main/data/sc_debt.Rds")

mi_ep <- read_rds(
  "https://github.com/jbisbee1/DS1000_F2024/raw/main/data/MI2020_ExitPoll_small.rds"
)

mi_final <- mi_ep %>%
  filter(
    preschoice == "Donald Trump, the Republican" |
      preschoice == "Joe Biden, the Democrat"
  ) %>%
  mutate(
    BidenVoter = ifelse(preschoice == "Joe Biden, the Democrat", 1, 0),
    TrumpVoter = 1 - BidenVoter,
    AGE10 = ifelse(AGE10 == 99, NA, AGE10)
  )

# Synthetic Simpson's-paradox example.
# Within each preparation group, more practice is associated with higher scores,
# while the pooled relationship goes in the opposite direction because the groups
# occupy different parts of the X/Y space.
simpson <- bind_rows(
  tibble(
    group = "Advanced course",
    practice = seq(1, 5, length.out = 35),
    score = 82 + 2.2 * practice + rnorm(35, 0, 1.8)
  ),
  tibble(
    group = "Intro course",
    practice = seq(5, 9, length.out = 35),
    score = 60 + 2.2 * practice + rnorm(35, 0, 1.8)
  )
)
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 7 completes the core instructional content for Unit 2.

Primary Test 2 competencies:
- identify plausible alternative explanations / confounders;
- reason about subgroup patterns;
- distinguish association from causation;
- communicate a calibrated conclusion.

INSTRUCTOR NOTE:
Keep historical/source-development notes invisible to students. The student-facing
deck should read as one coherent argument built from concepts they have already
encountered in this semester.
-->

# What Else Could Explain It?

.big[
Last class:

> **What relationship do I see?**

Today:

> **Why might I see that relationship?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → interrogate explanations → communicate**
]

Today:

**Interrogate explanations**

---

# Our Running Claim

.huge[
> **More selective colleges produce graduates who earn more.**
]

We have spent weeks earning the ability to evaluate this sentence.

---

# What We Have Established

Using admissions rate and graduate earnings, we can ask whether:

.big[
**selectivity and earnings are associated**
]

in the College Scorecard data.

Suppose they are.

What word in the original claim still causes trouble?

---

# The Dangerous Word

.huge[
**produce**
]

"Selective colleges **produce** graduates who earn more."

That is a causal claim.

.big[
An association is not yet a causal explanation.
]

<!-- LO1: Make the inferential gap explicit. -->

---

# Three Stories

Suppose X and Y move together.

At least three broad stories are possible:

.big[
**1. X → Y**

**2. Y → X**

**3. Z → X and Z → Y**
]

The observed X–Y relationship alone does not tell us which story generated it.

---

# Story 1: X Causes Y

For our college claim:

.big[
**Selectivity → graduate earnings**
]

Possible story:

Selective colleges provide resources, networks, instruction, or credentials that increase later earnings.

This is the story implied by the word **produce**.

---

# Story 2: Reverse Direction

Sometimes the direction we tell ourselves is backwards.

.big[
**Y → X**
]

This is called **reverse causality**.

Putting X on the horizontal axis does not make X the cause.

<!-- Instructor note: reinforce the distinction between axis placement and causal direction. -->

---

# Reverse Causality Example

Suppose we observe:

> People who exercise more report better health.

Possible story:

**Exercise → health**

But also:

.big[
**Health → ability to exercise**
]

The observed relationship alone does not separate these stories.

---

# Story 3: Something Else

Return to colleges.

Suppose selective colleges have higher-earning graduates.

What else might differ across colleges?

.big[
**Who enrolls?**

**What students study?**

**Where the college is located?**

**Which labor markets graduates enter?**
]

---

# We Already Saw This

Earlier in the course, we noted that graduate earnings may reflect:

- what the college did;
- who enrolled;
- field of study;
- geography;
- family background;
- labor-market conditions.

.big[
Those were not side notes.

They are now central to the causal question.
]

<!-- Instructor note: reconnect explicitly to the measurement cautions introduced earlier in the course. -->

---

# A Third Variable

Call another relevant variable:

.huge[
**Z**
]

Suppose Z is related to both:

**X: selectivity**

and

**Y: graduate earnings**

Then some of the X–Y relationship may reflect Z.

---

# Confounding

.big[
A **confounder** is a variable related to both the explanatory variable and the outcome that can provide an alternative explanation for their observed relationship.
]

For example:

**Z:** characteristics of students before college

Those characteristics may affect:

- which colleges students attend;
- how much they later earn.

<!-- LO1: Introduce confounding intuitively rather than formally. -->

---

# Another Familiar Pattern

Suppose we observe:

.big[
**More education → higher wages**
]

It is tempting to conclude that education itself produced the entire difference.

But education and wages may both reflect:

> family resources, prior preparation, ability, geography, or other opportunities.

.big[
The observed relationship may combine several processes.
]

---

# "Spurious" Does Not Mean Fake

A spurious relationship can be:

- real in the dataset;
- strong;
- reproducible;

and still fail to represent the causal story we first imagined.

.big[
The pattern can be real while the explanation is wrong.
]

---

# A Skeptic's Question

Whenever X and Y move together, ask:

.huge[
> **What else moves with both?**
]

This is one of the most useful habits in empirical reasoning.

---

# How Can We Investigate Z?

One simple strategy:

.big[
**Look at the X–Y relationship within relevant groups of Z.**
]

This is **conditional analysis**:

> How does the relationship look **depending on** another variable?

.big[
Instead of asking only "Are X and Y related?", we ask whether that relationship changes when Z changes.
]

---

# Add a Third Variable

Consider a different relationship:

> **Did men and women differ in support for Trump in Michigan in 2020?**

We can first compare men and women overall.

Then ask:

.big[
> **Does that gender difference look the same at different ages?**
]

Age becomes our third variable, Z.

---

# First: The Overall Gender Gap

```{r}
mi_final %>%
  filter(!is.na(SEX)) %>%
  group_by(SEX) %>%
  summarise(
    TrumpSupport = mean(TrumpVoter),
    .groups = "drop"
  )
```

.big[
This gives us an overall relationship between gender and vote choice.
]

---

# Then Condition on Age

```{r}
mi_final %>%
  filter(!is.na(SEX), !is.na(AGE10)) %>%
  group_by(SEX, AGE10) %>%
  summarise(
    TrumpSupport = mean(TrumpVoter),
    .groups = "drop"
  )
```

Now ask:

.big[
> **Is the gender gap similar within each age group?**
]

<!-- Instructor note: use this as a concrete transition from an overall comparison to a conditional comparison. -->

---

# Why Look Within Groups?

Because an aggregate relationship may combine:

- differences **within** groups;
- differences **between** groups;
- unequal group composition.

.big[
The pooled pattern can hide important structure.
]

---

# A More Dramatic Example

Sometimes the aggregate relationship can even point in a different direction from the relationships within groups.

This is known as:

.huge[
**Simpson's paradox**
]

---

# The Pooled Data

```{r, echo=FALSE}
simpson %>%
  ggplot(aes(x = practice, y = score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

.big[
Looking only at the pooled data, what relationship do you see?
]

---

# Now Reveal the Groups

```{r, echo=FALSE}
simpson %>%
  ggplot(aes(x = practice, y = score, shape = group)) +
  geom_point() +
  geom_smooth(aes(group = group), method = "lm", se = FALSE)
```

.big[
Within each course, more practice accompanies higher scores.

The pooled relationship tells a different story.
]

---

# What Happened?

Students in the two courses differ in:

- typical practice time;
- typical scores.

The group variable is related to both X and Y.

.big[
Ignoring the group changed the apparent relationship.
]

---

# The Lesson Is Not "Always Control"

A third variable can:

- explain part of a relationship;
- reveal heterogeneity;
- clarify a comparison;

but conditioning on variables is itself a substantive decision.

.big[
Do not add variables mechanically.

Ask why Z belongs in the explanation.
]

<!-- Important guardrail before later regression instruction. -->

---

# Not Every Z Is a Confounder

Suppose X affects Z, and Z then affects Y.

.big[
**X → Z → Y**
]

Z is now part of the pathway through which X may affect Y.

Conditioning on it answers a different question.

We will return to these distinctions later.

---

# The Core Habit

Before interpreting an association causally:

.big[
1. Name the causal story you are tempted to tell.

2. Generate plausible alternative stories.

3. Ask what evidence would distinguish them.
]

That is principled skepticism.

---

# AI Will Happily Explain

Suppose you tell an AI assistant:

> "College selectivity and graduate earnings have a strong negative correlation because lower admissions rates lead to better career outcomes."

It may produce an eloquent explanation involving:

- networks;
- signaling;
- resources;
- peers.

.huge[
Fluency is not evidence.
]

<!-- LO5: AI can elaborate a causal story without evidence that identifies it. -->

---

# Audit the AI's Story

Ask:

.big[
**What evidence supports this mechanism?**

**What alternative explanations were ignored?**

**Could student selection explain the pattern?**

**Could geography or field of study matter?**

**What evidence would distinguish these stories?**
]

AI can generate hypotheses.

It cannot turn an association into identified causation by explaining it confidently.

---

# Association Language

When the evidence is observational and bivariate:

Good:

> "Selectivity and earnings are associated."

> "More selective colleges tend to have higher-earning graduates."

> "The pattern is consistent with the claim."

---

# Causal Language

Stronger statements require stronger evidence:

> "Selectivity increases earnings."

> "Attending a selective college causes higher earnings."

> "Selective colleges produce higher-earning graduates."

.big[
Those statements require us to rule out credible alternative explanations.
]

---

# "Consistent With" Is Useful

.big[
**Results can be consistent with a theory without proving the theory.**
]

This is useful language when the evidence fits an explanation but does not uniquely establish it.

Evidence can increase or decrease our confidence without uniquely identifying one explanation.

---

# What Would Stronger Evidence Look Like?

To make a stronger causal claim, we would want evidence that helps separate:

.big[
**what the college does**

from

**who attends the college**
]

Possible strategies might compare more similar students or exploit research designs that make exposure to different colleges less confounded.

We are not solving that problem today.

---

# Our Goal Today

You do not need to master causal inference.

You need to develop the reflex:

.huge[
**Association observed. Explanation not yet earned.**
]

---

# Practice

Claim:

> **Students who use AI to study earn higher grades.**

Suppose the association is real.

Name a plausible Z.

--

Examples might include:

- prior academic preparation;
- motivation;
- course difficulty;
- access to resources.

.big[
Which variables could affect both AI use and grades?
]

---

# Practice

Claim:

> **Neighborhoods with more police are safer.**

Possible stories:

.big[
Police → safety

Safety → policing decisions

Third variables → both policing and safety
]

What third variables might matter?

---

# Practice

Claim:

> **Older NBA players score more points.**

What else might affect both:

- player age / experience;
- total points?

.big[
Playing time?

Career selection?

Player quality?

Role on the team?
]

---

# A Better Conclusion

Return to our original claim.

Instead of:

> "Selective colleges produce graduates who earn more."

A calibrated conclusion is:

.big[
> **More selective colleges tend to have higher graduate earnings in these data, but this association alone does not establish that selectivity causes higher earnings.**
]

---

# Why This Matters

The world is full of claims that move silently from:

.big[
**X and Y are related**
]

to:

.huge[
**X causes Y**
]

Politicians do it.

News stories do it.

Peers do it.

AI does it.

Data scientists should notice the jump.

---

# The Unit 2 Workflow

.big[
**Choose the comparison**

↓

**See the relationship**

↓

**Describe the relationship**

↓

**Ask what else could explain it**

↓

**Calibrate the conclusion**
]

That is the reasoning Test 2 asks you to transfer.

---

# Before You Leave

When you see an association, ask:

1. What is the observed X–Y relationship?
2. What causal story am I tempted to tell?
3. Could Y affect X?
4. What Z might affect both?
5. Does the relationship differ across relevant groups?
6. What evidence would distinguish the explanations?
7. What language does the evidence actually justify?

---

# The Rule to Remember

.huge[
**Association is evidence.**

**Explanation must be earned.**
]

---

# Next

Homework 7 and Quiz 6 will practice:

- generating alternative explanations;
- identifying plausible confounders;
- reasoning about subgroup patterns;
- recognizing Simpson's paradox;
- auditing causal interpretations from AI;
- writing calibrated conclusions.

Then:

.huge[
**Evidence Debate 1**
]
