---
title: "Variables Together"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
library(tidyverse)
df <- read_rds("https://github.com/jbisbee1/DS1000_F2024/raw/main/data/sc_debt.Rds")
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO2: Translate questions into data problems.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 5 begins Unit 2. It primarily advances LO3 and LO4 while reinforcing LO1, LO2, LO5, and LO6.

INSTRUCTOR NOTE:
Keep the conceptual focus on choosing representations from the variable types and substantive question.
Do not turn this into a catalogue of ggplot geoms.
Correlation, strength/form, confounding, and causal interpretation belong primarily to Lectures 6–7.
-->

# Variables Together

.big[
For four classes, we have deliberately avoided answering our original question.

Now we finally put the variables together.
]

---

# Our Question

.big[
> **More selective colleges produce graduates who earn more.**
]

We have already asked:

- What does "selectivity" mean?
- How will we measure it?
- What does one row represent?
- What does admissions rate look like?
- What does earnings look like?
- Which observations belong in our analysis?

Now:

.huge[
**How do X and Y relate?**
]

<!-- LO2: The multivariate step follows the earlier decomposition rather than replacing it. -->

---

# Where We Are

.big[
**Question → Concepts → Variables → Unit of analysis → Understand each variable → Prepare the evidence → Put variables together → Evaluate → Communicate**
]

Today:

**Put variables together**

---

# One Variable vs. Two

In Unit 1:

> What is **X** like?

or

> What is **Y** like?

Now:

> What happens to **Y** as **X** changes?

.big[
That is a **multivariate** question.
]

---

# Start With the Question

Before choosing a graph, identify:

.big[
**What is X?**

**What is Y?**

**What type is each variable?**
]

The types of X and Y determine which representations make sense.

<!-- LO3: Method selection follows the data structure. -->

---

# Two Continuous Variables

Suppose:

**X:** admissions rate

**Y:** graduate earnings

Both are continuous.

.big[
**Continuous + continuous → scatterplot**
]

---

# A Scatterplot

```{r}
df %>%
  ggplot(aes(x = adm_rate, y = md_earn_wne_p6)) +
  geom_point()
```

Each point represents one observation.

In these data:

> **one point = one college**

<!-- LO4: Connect visual mark to unit of analysis. -->

---

# Read the Axes First

Before interpreting the cloud of points:

.big[
**What does the x-axis measure?**

**What does the y-axis measure?**
]

A graph cannot rescue a bad operationalization.

The measurement questions from Unit 1 still apply.

<!-- LO1: Multivariate analysis does not erase construct-validity concerns. -->

---

# What Can We Ask?

From this plot we can begin asking:

- Do higher values of X tend to accompany higher or lower values of Y?
- Is there an obvious pattern?
- Are some observations unusual?
- Does the relationship look simple or complicated?

.big[
Today we learn to **see the relationship**.

Next class we learn to describe it more precisely.
]

---

# Direction

Look again at the College Scorecard plot.

.big[
As admissions rate increases, what appears to happen to earnings?
]

Remember:

Lower admissions rate = more selective.

So interpretation requires understanding what the variable means, not merely whether the dots slope up or down.

<!-- LO4 + LO6: Translate graphical direction back into substantive meaning. -->

---

# X and Y Have Roles

A useful convention:

**X** = explanatory / predictor variable

**Y** = outcome / response variable

For our question:

**X:** admissions rate

**Y:** earnings

.big[
But putting X on the horizontal axis does **not** prove X causes Y.
]

<!-- LO1: Early guardrail against causal interpretation. -->

---

# The Plot Is Evidence

A scatterplot is not decoration.

It is one way to ask:

.big[
> **What pattern is present in these observations?**
]

But it does not automatically tell us:

> Why does that pattern exist?

That distinction will become central later in this unit.

---

# Continuous + Categorical

Now consider a different question:

> **Do public and private colleges have different graduate earnings?**

Suppose:

**X:** college type — categorical

**Y:** earnings — continuous

.big[
**Categorical + continuous → compare distributions across groups**
]

---

# Compare Groups

A boxplot is one useful representation.

```{r}
df %>%
  filter(!is.na(control), !is.na(md_earn_wne_p6)) %>%
  ggplot(aes(x = factor(control), y = md_earn_wne_p6)) +
  geom_boxplot()
```

<!-- Instructor note: If `control` is not available in the current Scorecard file, substitute a categorical variable that is. -->

---

# What Is the Comparison?

The question is no longer:

> What does earnings look like?

It is:

.big[
> **How does the distribution of earnings differ across categories of X?**
]

Look for differences in:

- typical values
- spread
- unusual observations

---

# Same Variable, New Question

Notice what changed.

In Lecture 3:

> What does `md_earn_wne_p6` look like?

Today:

> Does `md_earn_wne_p6` look different across groups?

.big[
The visualization changes because the **question changed**.
]

<!-- LO3: Reinforces that visualization choice is question-dependent. -->

---

# Categorical + Categorical

Consider:

> **Are private colleges more likely to be highly selective?**

Suppose both variables are categorical.

.big[
**Categorical + categorical → compare counts or proportions across groups**
]

This is different from comparing two continuous quantities.

---

# Counts or Proportions?

Suppose one group contains 5,000 colleges and another contains 500.

Raw counts may be difficult to compare.

.big[
Sometimes the relevant question is not:

**How many?**

but:

**What proportion?**
]

<!-- LO3: Choice of summary depends on substantive comparison. -->

---

# The Decision Rule

.huge[
**Continuous + continuous → scatterplot**

**Categorical + continuous → compare distributions**

**Categorical + categorical → compare counts/proportions**
]

This is a starting rule, not a law of nature.

---

# What About X and Y?

The variable types tell us which displays are possible.

The substantive question tells us which variables play which roles.

.big[
Do not start with:

> "I want to make a scatterplot."

Start with:

> **"What relationship am I trying to see?"**
]

<!-- LO3: Analytical method follows question. -->

---

# A Common Mistake

Suppose someone asks:

> "Do public and private colleges differ in earnings?"

An AI assistant generates a histogram of earnings for all colleges.

Does the graph show earnings?

Yes.

Does it answer the question?

.huge[
**No.**
]

---

# Correct Code, Wrong Analysis

.big[
A graph can be:

- valid R
- attractive
- correctly labeled

and still fail to answer the research question.
]

This is the multivariate version of a lesson from Unit 1:

> **Code that runs is not necessarily analysis that answers the question.**

<!-- LO5: Evaluate AI-generated implementation against analytical intent. -->

---

# AI Can Choose Too

You can ask an AI assistant:

> "I have a continuous outcome Y and a categorical explanatory variable X. What visualization should I use?"

It will usually give you a plausible answer.

Your responsibility is to ask:

.big[
**Does that representation actually match my variables and my question?**
]

<!-- LO5: AI as proposer; student as evaluator. -->

---

# Audit the AI

When AI proposes a visualization, check:

1. What is X?
2. What is Y?
3. What type is each?
4. What does one mark on the graph represent?
5. Does the graph display the comparison the question asks about?

If you cannot answer those questions, the visualization is not yet evidence you understand.

---

# Back to Selectivity

Our running question asks about:

**X:** admissions rate

**Y:** graduate earnings

Both continuous.

So the appropriate first representation is:

.big[
**a scatterplot**
]

Now we can finally see whether the two variables move together.

---

# But Read Carefully

Suppose the points generally slope downward.

Because admissions rate is our X:

> Higher admissions rates accompany lower earnings.

But substantively:

> **More selective colleges tend to have higher earnings.**

.big[
Statistical interpretation must return to the meaning of the variables.
]

<!-- LO6: Translate statistical representation into substantive language. -->

---

# An Observed Pattern

Suppose we see a relationship.

What can we say today?

.big[
> **These variables appear to be associated in these data.**
]

What can we **not** yet say?

> Why the relationship exists.

> Whether it is causal.

> How much confidence we should place in it beyond these observations.

---

# Keep the Units Straight

A point on the plot represents a **college**.

The variables describe characteristics of colleges.

Therefore the relationship is:

.big[
**between colleges**
]

It does not automatically tell us what happens to an individual student who switches colleges.

<!-- LO1: Guards against ecological/level-of-analysis slippage without requiring formal terminology yet. -->

---

# Three Questions, Three Displays

.big[
**Are admissions rates related to earnings?**

→ continuous + continuous

→ scatterplot
]

--

.big[
**Do college types differ in earnings?**

→ categorical + continuous

→ compare distributions
]

--

.big[
**Does selectivity category differ by college type?**

→ categorical + categorical

→ compare counts/proportions
]

---

# Practice: Choose the Display

Claim:

> **Older NBA players score more points.**

Variables:

- age: continuous
- points: continuous

What should we use first?

.huge[
**Scatterplot**
]

<!-- LO3: Transfer beyond College Scorecard. -->

---

# Practice: Choose the Display

Claim:

> **Rookies score fewer points than veteran players.**

Variables:

- rookie status: categorical
- points: continuous

What should we use first?

.huge[
**Compare the distributions across groups**
]

---

# Practice: Choose the Display

Claim:

> **Rookies are more likely to play for Western Conference teams.**

Variables:

- rookie status: categorical
- conference: categorical

What should we use first?

.huge[
**Compare counts or proportions**
]

---

# A Visualization Is an Argument

Choosing a graph means choosing what comparison to make visible.

.big[
A good visualization makes the relevant comparison easy to see.

A bad visualization can obscure the comparison even when every number is correct.
]

<!-- LO6: Visualization as communication of evidence. -->

---

# What We Have Added

Unit 1 taught us to ask:

> **Can I trust these variables as evidence?**

Today adds:

> **How should I put the variables together so I can see the relationship I care about?**

---

# The Next Question

Once we have the right visualization, we still need to ask:

.huge[
**How should we describe the relationship?**
]

Direction?

Strength?

Form?

Unusual observations?

Can one number summarize it?

That is Lecture 6.

---

# Before You Leave

Given any two-variable question, you should be able to answer:

1. What is X?
2. What is Y?
3. What type is each variable?
4. What does one observation represent?
5. What comparison does the question require?
6. What visualization makes that comparison visible?
7. What pattern can you describe without overclaiming?

<!-- LO3 + LO4 + LO6: Reusable multivariate checklist. -->

---

# The Rule to Remember

.huge[
**Do not choose the graph first.**
]

.big[
Choose the comparison first.

Then let the question and variable types determine the graph.
]

---

# Homework 5

In Homework 5, you will practice:

- identifying X and Y;
- classifying their variable types;
- choosing an appropriate multivariate representation;
- using AI to generate visualization code;
- checking whether the visualization actually answers the question;
- describing the visible pattern without explaining it away.

Next class:

.big[
**Describe the Relationship**
]
