---
title: "How Do We Interpret and Evaluate Regression Evidence?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

debt <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/sc_debt.Rds") %>%
  filter(adm_rate > 0, md_earn_wne_p6 > 0, sat_avg > 0)

model_adm <- lm(md_earn_wne_p6 ~ adm_rate, data = debt)
debt$pred_adm <- predict(model_adm)
debt$resid_adm <- debt$md_earn_wne_p6 - debt$pred_adm

rmse_in_sample <- sqrt(mean(debt$resid_adm^2))

# Single train/test split, for the walkthrough.
set.seed(1021)
inds <- sample(1:nrow(debt), size = round(nrow(debt) / 2), replace = FALSE)
train <- debt %>% slice(inds)
test <- debt %>% slice(-inds)

model_train <- lm(md_earn_wne_p6 ~ adm_rate, data = train)
test$pred <- predict(model_train, newdata = test)
rmse_holdout_once <- sqrt(mean((test$md_earn_wne_p6 - test$pred)^2, na.rm = TRUE))

# Repeated cross-validation, for adm_rate.
set.seed(123)
cv_adm <- NULL
for (i in 1:100) {
  inds_i <- sample(1:nrow(debt), size = round(nrow(debt) / 2), replace = FALSE)
  train_i <- debt %>% slice(inds_i)
  test_i <- debt %>% slice(-inds_i)
  m_i <- lm(md_earn_wne_p6 ~ adm_rate, data = train_i)
  test_i$pred <- predict(m_i, newdata = test_i)
  cv_adm <- c(cv_adm, sqrt(mean((test_i$md_earn_wne_p6 - test_i$pred)^2, na.rm = TRUE)))
}

# Repeated cross-validation, for sat_avg, to compare models.
set.seed(123)
cv_sat <- NULL
for (i in 1:100) {
  inds_i <- sample(1:nrow(debt), size = round(nrow(debt) / 2), replace = FALSE)
  train_i <- debt %>% slice(inds_i)
  test_i <- debt %>% slice(-inds_i)
  m_i <- lm(md_earn_wne_p6 ~ sat_avg, data = train_i)
  test_i$pred <- predict(m_i, newdata = test_i)
  cv_sat <- c(cv_sat, sqrt(mean((test_i$md_earn_wne_p6 - test_i$pred)^2, na.rm = TRUE)))
}
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 12 completes Unit 3's core instructional content. It extends the
adm_rate model built in Lecture 11 with residual diagnostics, RMSE, and
cross-validation -- deliberately framed as "the bootstrap idea, applied to
model evaluation" to tie the Regression sub-unit back to Uncertainty I/II.

INSTRUCTOR NOTE:
This is the last instructional lecture before the Unit 3 Assessment Day
(Evidence Debate 3 + Unit Test 3). Per the established course pattern, it
closes with a Daily Quiz callout and an assessment-day tease, not a
"Homework N" slide -- the next homework (HW10) belongs to Lecture 13
(Prediction vs. Explanation, the Unit 4 opener).
-->

# How Do We Interpret and Evaluate Regression Evidence?

.big[
Last class:

> **How can a model summarize a relationship?**

Today:

> **How good is that summary, and how do we know?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → interrogate explanations → quantify uncertainty → summarize with a model → evaluate the model → communicate**
]

Today:

**Evaluate the model**

---

# Picking Up Where We Left Off

Yesterday's model:

```{r}
model_adm <- lm(md_earn_wne_p6 ~ adm_rate, data = debt)
```

.big[
We interpreted the intercept and slope. We never asked: **is this a good model?**
]

---

# Every Model Makes Mistakes

.big[
If a line fit every observation perfectly, it wouldn't be summarizing anything — it would just be tracing the data.
]

The question is not *whether* the model makes mistakes. It's **how big, and how patterned**.

---

# Residuals, Again

For every college, the model predicts earnings. The residual is:

.huge[
$\text{residual} = Y - \hat{Y}$
]

```{r}
debt$pred_adm <- predict(model_adm)
debt$resid_adm <- debt$md_earn_wne_p6 - debt$pred_adm

summary(debt$resid_adm)
```

---

# Look at the Residuals: Univariate

```{r}
debt %>%
  ggplot(aes(x = resid_adm)) +
  geom_histogram(bins = 40) +
  labs(x = "Residual (actual − predicted earnings)")
```

.big[
Residuals average to zero by construction. Look instead at their **shape**.
]

---

# Look at the Residuals: Multivariate

```{r}
debt %>%
  ggplot(aes(x = adm_rate, y = resid_adm)) +
  geom_point(alpha = .4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth() +
  labs(x = "Admission rate", y = "Residual")
```

---

# What Would We Want to See?

.big[
**Ideal:** residuals scattered evenly around zero across the whole range of $X$ — a formless cloud.

**Warning sign:** a visible pattern — a curve, a funnel, a cluster — meaning the model systematically over- or under-predicts in some region of $X$.
]

A pattern in the residuals means there is structure left in the data that a straight line isn't capturing.

<!-- LO4: Let the rendered plot drive the discussion; don't presume its exact shape here. -->

---

# From a Picture to a Number

.big[
Residual plots are essential, but we also want a **single number** to summarize how good the model is — especially for comparing models.
]

We already have the tool: **RMSE**.

---

# RMSE, Again

```{r}
rmse_in_sample <- sqrt(mean(debt$resid_adm^2))
round(rmse_in_sample, 0)
```

.big[
On average, this model's predictions are off by about \$`r scales::comma(round(rmse_in_sample,0))`.
]

Is that good? Good compared to **what**?

---

# A Hidden Problem

.big[
We calculated RMSE using the **same data** we used to fit the model.
]

The model was built to minimize error on exactly this data. Its RMSE here is close to the *best-case* performance.

.big[
That is not the same as asking: **how well would this model do on data it hasn't seen?**
]

---

# This Should Sound Familiar

.big[
We faced a version of this problem in Unit 3 already.
]

A single sample can make a comparison look more solid than it really is. **Bootstrapping** asked: *would this hold up under a different sample?*

Today's version of that question: **would this model's accuracy hold up on data it wasn't fit to?**

---

# Train and Test

.big[
**Split the data in two:**

- Fit the model on a **training** set.
- Evaluate its RMSE on a separate **test** set the model never saw.
]

This is called **cross-validation**.

---

# One Split

```{r, eval=FALSE}
set.seed(1021)
inds <- sample(1:nrow(debt), size = round(nrow(debt) / 2), replace = FALSE)
train <- debt %>% slice(inds)
test <- debt %>% slice(-inds)

model_train <- lm(md_earn_wne_p6 ~ adm_rate, data = train)
test$pred <- predict(model_train, newdata = test)
```

```{r}
rmse_holdout_once <- sqrt(mean((test$md_earn_wne_p6 - test$pred)^2, na.rm = TRUE))
round(rmse_holdout_once, 0)
```

---

# In-Sample vs. Held-Out RMSE

.big[
In-sample RMSE: \$`r scales::comma(round(rmse_in_sample,0))`

One held-out RMSE: \$`r scales::comma(round(rmse_holdout_once,0))`
]

A single train/test split depends on which colleges happened to land in the test set — the same fragility we saw with a single bootstrap resample.

---

# Repeat It, Like a Bootstrap

```{r, eval=FALSE}
cv_adm <- NULL
for (i in 1:100) {
  inds_i <- sample(1:nrow(debt), size = round(nrow(debt) / 2), replace = FALSE)
  train_i <- debt %>% slice(inds_i)
  test_i <- debt %>% slice(-inds_i)
  m_i <- lm(md_earn_wne_p6 ~ adm_rate, data = train_i)
  test_i$pred <- predict(m_i, newdata = test_i)
  cv_adm <- c(cv_adm, sqrt(mean((test_i$md_earn_wne_p6 - test_i$pred)^2, na.rm = TRUE)))
}
```

.big[
100 different train/test splits, each producing its own held-out RMSE.
]

---

# The Distribution of Held-Out RMSE

```{r}
data.frame(rmse = cv_adm) %>%
  ggplot(aes(x = rmse)) +
  geom_density(fill = "grey70") +
  geom_vline(xintercept = mean(cv_adm), linetype = "dashed") +
  labs(x = "Held-out RMSE across 100 splits")
```

---

# A More Honest Estimate

.big[
Average held-out RMSE: \$`r scales::comma(round(mean(cv_adm),0))`

In-sample RMSE: \$`r scales::comma(round(rmse_in_sample,0))`
]

Cross-validation gives a more honest sense of how the model would perform on colleges it wasn't built from.

---

# Using RMSE to Compare Models

Recall Lecture 11's second model, using `sat_avg` instead of `adm_rate`.

.big[
Which predictor summarizes graduate earnings more reliably?
]

Cross-validate **both** and compare their average held-out RMSE.

---

# Comparing Two Models

```{r}
tibble(
  model = c("md_earn_wne_p6 ~ adm_rate", "md_earn_wne_p6 ~ sat_avg"),
  mean_cv_rmse = c(mean(cv_adm), mean(cv_sat))
)
```

.big[
The model with the **lower** average held-out RMSE makes smaller predictive mistakes on data it wasn't fit to.
]

---

# What a Lower RMSE Does *Not* Mean

.big[
A lower RMSE means one model's predictions are, on average, closer to the truth.
]

It does **not** mean:

- that predictor is the *causal* driver of earnings;
- the winning model is correct, only less wrong;
- either model would generalize to colleges very different from these.

<!-- LO1: Keep separating predictive performance from causal explanation. -->

---

# Audit an AI Claim

Suppose an AI assistant compares the two RMSE values and says:

> "This proves that admission rate is the true cause of earnings differences, since its model has lower error."

.big[
**What does a lower cross-validated RMSE actually measure?**

**Does better prediction imply a causal mechanism?**

**What Unit 2 habits still apply here?**
]

Predictive accuracy and causal explanation are different questions.

---

# A Calibrated Conclusion

Instead of:

> "This model proves what drives earnings."

Try:

.big[
> **Cross-validated on 100 random splits, the admission-rate model's predictions are off by about \$`r scales::comma(round(mean(cv_adm),0))`, on average, on colleges it wasn't fit to — a measure of predictive reliability, not proof of what causes earnings differences.**
]

---

# Before You Leave

When someone hands you a regression's RMSE, ask:

1. Was it calculated on the same data used to fit the model?
2. Would it hold up on data the model hasn't seen?
3. Does the residual plot show a formless cloud, or a pattern?
4. If comparing models, which has the lower held-out error?
5. Does a lower error imply anything about *why* the relationship exists?

---

# The Rule to Remember

.huge[
**A good-looking fit on the data you trained on is the easy version of the question.**

**Cross-validation asks the harder, more honest one.**
]

---

# The Unit 3 Workflow, Complete

.big[
**Compare in a sample**

↓

**Resample to measure confidence**

↓

**Ask what that confidence justifies**

↓

**Summarize with a model**

↓

**Evaluate the model out of sample**
]

This is the same reasoning pattern the Evidence Debate and Unit Test will ask you to transfer.

---

# Daily Quiz 12

.huge[

Now test yourself independently.

]

.big[

Open **Daily Quiz 12** in Brightspace.

]

.small[

No AI or internet access during the quiz.

]
