---
title: "How Do We Predict Categories?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)
library(scales)

gms <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/game_summary.Rds")

set.seed(123)
inds <- sample(1:nrow(gms), size = round(nrow(gms) / 2), replace = FALSE)
train <- gms %>% slice(inds)
test <- gms %>% slice(-inds)

model_class <- lm(isWin ~ tov + pts + treb + pctFG, data = train)
test$pred_prob <- predict(model_class, newdata = test)
test$pred_class <- ifelse(test$pred_prob >= 0.5, TRUE, FALSE)

conf_mat <- test %>%
  count(isWin, pred_class)

get_n <- function(actual, predicted) {
  v <- conf_mat %>% filter(isWin == actual, pred_class == predicted) %>% pull(n)
  if (length(v) == 0) 0 else v
}

tp <- get_n(TRUE, TRUE)
tn <- get_n(FALSE, FALSE)
fp <- get_n(FALSE, TRUE)
fn_ <- get_n(TRUE, FALSE)

accuracy <- (tp + tn) / (tp + tn + fp + fn_)
baseline_acc <- max(mean(test$isWin), 1 - mean(test$isWin))
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 14 turns Lecture 13's "broken" linear probability model into a real
classification workflow: threshold the predicted values, build a confusion
matrix, calculate accuracy, and compare against a baseline. Sensitivity /
specificity are introduced only by name, as a teaser -- full treatment of
evaluation tradeoffs (thresholds, ROC/AUC) is reserved for Classification II.

INSTRUCTOR NOTE:
Continues the game_summary.Rds / isWin thread from Lecture 13 rather than
introducing a new dataset, to keep the unit's cognitive load manageable.
-->

# How Do We Predict Categories?

.big[
Last class:

> **What happens when I use a continuous-outcome tool on a categorical outcome?**

Today:

> **How do I turn that into an actual classifier, and know if it's any good?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → interrogate explanations → quantify uncertainty → summarize with a model → evaluate the model → choose the right goal → build a classifier**
]

Today:

**Build a classifier**

---

# Picking Up Where We Left Off

Lecture 13 fit this on the training half of the data:

```{r, eval=FALSE}
model_class <- lm(isWin ~ tov + pts + treb + pctFG, data = train)
test$pred_prob <- predict(model_class, newdata = test)
```

.big[
The predictions weren't real probabilities — but we can still use them to make a **decision**.
]

---

# From a Number to a Decision

.big[
**Classification:** turning a predicted score into a predicted category.
]

```{r}
test$pred_class <- ifelse(test$pred_prob >= 0.5, TRUE, FALSE)

test %>%
  select(idGame, isWin, pred_prob, pred_class) %>%
  head(5)
```

---

# Every Prediction Is One of Four Things

For a binary outcome, compare the prediction to the truth:

.big[
- Predicted **win**, actually won → correct
- Predicted **loss**, actually lost → correct
- Predicted **win**, actually lost → wrong
- Predicted **loss**, actually won → wrong
]

---

# The Confusion Matrix

```{r}
test %>%
  count(isWin, pred_class)
```

.big[
A **confusion matrix**: every combination of truth and prediction, counted.
]

---

# Naming the Four Cells

.big[
**True positive:** predicted win, actually a win.

**True negative:** predicted loss, actually a loss.

**False positive:** predicted win, actually a loss.

**False negative:** predicted loss, actually a win.
]

"Positive" here just means the class we're tracking — a win.

---

# From Counts to Accuracy

.huge[
$$\text{Accuracy} = \frac{\text{correct predictions}}{\text{all predictions}}$$
]

```{r}
accuracy <- (tp + tn) / (tp + tn + fp + fn_)
round(accuracy * 100, 1)
```

---

# Is That Good?

.big[
Accuracy alone means nothing without a **baseline** to compare it to.
]

The simplest possible classifier: always guess the more common outcome.

```{r}
round(baseline_acc * 100, 1)
```

.big[
Our model should beat this "dumbest guess" baseline — if it doesn't, it isn't adding anything.
]

---

# A Trap: Accuracy Can Hide Unevenness

.big[
A model can post a great overall accuracy while doing **very differently** on the two classes.
]

Imagine a classifier that is excellent at predicting losses but bad at predicting wins. Overall accuracy could still look fine, because losses may be more common.

---

# Looking Inside the Matrix

```{r}
conf_mat %>%
  group_by(isWin) %>%
  mutate(total = sum(n), pct = percent(n / total))
```

.big[
Same accuracy number can come from very different confusion matrices. Always look inside it.
]

---

# Two Names Worth Knowing

.big[
**Sensitivity:** of the games actually won, what share did we correctly predict as wins?

**Specificity:** of the games actually lost, what share did we correctly predict as losses?
]

We will unpack these — and the tradeoff between them — in the next class.

---

# The Threshold Was a Choice

.big[
We predicted "win" whenever `pred_prob >= 0.5`. That cutoff was arbitrary.
]

```{r}
test$pred_class_40 <- ifelse(test$pred_prob >= 0.4, TRUE, FALSE)

mean(test$pred_class_40 == test$isWin)
```

Changing the threshold changes the confusion matrix — and the accuracy.

---

# Same Held-Out Discipline

.big[
Notice: everything today still respected the train/test split from Lecture 12.
]

We never evaluated the classifier on the same games used to fit it. That habit doesn't change just because the outcome is now categorical.

---

# Do It Together

.blue[Claim]: **A simpler model — using only turnovers — classifies about as well as the full model.**

1. Fit `isWin ~ tov` on `train`.
2. Predict on `test`, threshold at 0.5.
3. Build the confusion matrix.
4. Compare its accuracy to the full model's.

---

# Audit an AI Claim

Suppose an AI assistant is shown only the specificity of a classifier (95%) and says:

> "This model is excellent — it correctly predicts the outcome 95% of the time."

.big[
**What does specificity alone measure?**

**What is missing from this claim?**

**What could make this misleading even if it's technically true?**
]

Quoting one favorable number from a confusion matrix is not the same as reporting overall performance.

---

# A Calibrated Conclusion

Instead of:

> "The model predicts game outcomes."

Try:

.big[
> **On held-out games, the classifier correctly predicted the outcome `r round(accuracy*100,1)`% of the time, compared to `r round(baseline_acc*100,1)`% for always guessing the more common outcome — but its performance was not identical across wins and losses, which the next class will examine more closely.**
]

---

# Before You Leave

When you see a classification result, ask:

1. What threshold turned predicted scores into predicted categories?
2. What does the confusion matrix look like, not just the accuracy number?
3. How does accuracy compare to the simplest possible baseline?
4. Is performance even across both classes, or lopsided?
5. Was this evaluated on data the model wasn't fit to?

---

# The Rule to Remember

.huge[
**A single accuracy number can hide a lot.**

**Always look inside the confusion matrix.**
]

---

# Homework 11

You will practice:

- turning predicted scores into classified predictions using a threshold;
- building and reading a confusion matrix;
- comparing a model's accuracy to a baseline guess;
- recognizing when overall accuracy hides uneven performance;
- auditing an AI claim that cites one favorable number in isolation.

Next:

.big[
**Classification II**
]
