---
title: "Relationships in Practice: NBA Scoring"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis-fonts", "css/ds1000.css"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
.hero-image {
  display: flex;
  align-items: center;
  justify-content: center;
  height: 74vh;
}
.hero-image img {
  max-width: 92%;
  max-height: 74vh;
  object-fit: contain;
}

.dag-wrap {
  display: flex;
  justify-content: center;
  align-items: center;
  margin-top: 1.5rem;
}
.dag-wrap svg {
  width: min(640px, 92%);
  height: auto;
}

.plot-wrap {
  display: flex;
  justify-content: center;
  align-items: center;
  height: 70vh;
}
.plot-wrap img {
  max-width: 96%;
  max-height: 68vh;
  object-fit: contain;
}

.callout {
  border-left: 5px solid #b89b5e;
  padding: 0.8rem 1rem;
  background: #f3eee4;
  margin: 1rem 0;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 11,
  fig.height = 7
)

library(tidyverse)
library(scales)

nba <- readr::read_rds(
  "https://github.com/jbisbee1/DS1000_F2024/raw/main/data/nba_players_2018.Rds"
)

# Keep the variables used in the case and remove incomplete player-seasons.
nba_case <- nba %>%
  select(fg3a, pts, minutes) %>%
  drop_na()

# Quartiles used for the conditional comparison in Evidence C.
nba_case <- nba_case %>%
  mutate(
    minutes_group = ntile(minutes, 4),
    minutes_group = factor(
      minutes_group,
      levels = 1:4,
      labels = c("Lowest minutes", "Low-mid minutes", "High-mid minutes", "Highest minutes")
    )
  )
```

# Relationships in Practice

<div class="huge">NBA scoring</div>

<div class="callout huge">
Today is not about learning a new technique.

It is about putting the Unit 2 reasoning workflow together.
</div>

---

# The claim

.hero-image[
.huge[

> **Players who attempt more three-pointers score more points because taking more threes makes them more effective scorers.**

]
]

---

# Before we see the evidence

Work through Section 1 of the worksheet.

.big[

What are the **unit of analysis**, **X**, and **Y**?

What would we expect to see if the claim were correct?

]

---

# Meet the data

.big[

**Unit of analysis:** one player-season

**X:** `fg3a` — three-point attempts

**Y:** `pts` — total points scored

**Z:** `minutes` — minutes played during the season

]

<div class="callout">Before we analyze the relationship, notice the measurement issue: <strong>total points is not the same thing as scoring effectiveness.</strong></div>

---

# Evidence A

```{r evidence-a}
ggplot(nba_case, aes(x = fg3a, y = pts)) +
  geom_point(alpha = 0.55, size = 2.4) +
  labs(
    title = "Points scored and three-point attempts",
    subtitle = "Each point represents one player-season",
    x = "Three-point attempts",
    y = "Points scored"
  ) +
  theme_minimal(base_size = 18)
```

---

# What does Evidence A show?

Work through Section 2 of the worksheet.

.big[

Describe the relationship:

**Direction → Form → Strength → Variation**

]

---

# What does Evidence A warrant?

.big[

The data show a **positive association** between three-point attempts and total points scored.

But the graph alone does **not** establish that taking more threes causes players to score more points.

]

---

# What else could explain it?

Work through Section 3 of the worksheet **before** we reveal more evidence.

.huge[

What variable **Z** might help explain why X and Y are related?

]

---

# A candidate explanation

<div class="dag-wrap">
<svg viewBox="0 0 520 260" role="img" aria-label="Minutes played points to three-point attempts and points scored, while three-point attempts also point to points scored">
  <defs>
    <marker id="arrowhead" markerWidth="8" markerHeight="8" refX="7" refY="4" orient="auto">
      <path d="M0,0 L8,4 L0,8 Z" fill="#333"/>
    </marker>
  </defs>

  <rect x="210" y="20" width="100" height="55" rx="8" fill="#fafafa" stroke="#777" stroke-width="2"/>
  <text x="260" y="55" text-anchor="middle" font-size="22" font-weight="700">Minutes</text>

  <rect x="80" y="170" width="100" height="55" rx="8" fill="#fafafa" stroke="#777" stroke-width="2"/>
  <text x="130" y="205" text-anchor="middle" font-size="22" font-weight="700">3PA</text>

  <rect x="340" y="170" width="100" height="55" rx="8" fill="#fafafa" stroke="#777" stroke-width="2"/>
  <text x="390" y="205" text-anchor="middle" font-size="22" font-weight="700">Points</text>

  <line x1="225" y1="75" x2="165" y2="165" stroke="#333" stroke-width="2.5" marker-end="url(#arrowhead)"/>
  <line x1="295" y1="75" x2="355" y2="165" stroke="#333" stroke-width="2.5" marker-end="url(#arrowhead)"/>
  <line x1="180" y1="197" x2="335" y2="197" stroke="#333" stroke-width="2.5" marker-end="url(#arrowhead)"/>
</svg>
</div>

---

# What would make minutes a plausible confounder?

.big[

We need evidence that minutes played is related to:

1. **three-point attempts**, and
2. **points scored**.

]

---

# Evidence B1

```{r evidence-b1}
ggplot(nba_case, aes(x = minutes, y = fg3a)) +
  geom_point(alpha = 0.55, size = 2.4) +
  labs(
    title = "Three-point attempts and minutes played",
    subtitle = "Each point represents one player-season",
    x = "Minutes played",
    y = "Three-point attempts"
  ) +
  theme_minimal(base_size = 18)
```

---

# Evidence B2

```{r evidence-b2}
ggplot(nba_case, aes(x = minutes, y = pts)) +
  geom_point(alpha = 0.55, size = 2.4) +
  labs(
    title = "Points scored and minutes played",
    subtitle = "Each point represents one player-season",
    x = "Minutes played",
    y = "Points scored"
  ) +
  theme_minimal(base_size = 18)
```

---

# What does Evidence B tell us?

Work through Section 4 of the worksheet.

.big[

Minutes played is strongly related to **both** three-point attempts and total points.

So minutes played is a **plausible alternative explanation** for at least part of the original relationship.

]

---

# But is that the whole story?

.huge[

We need another comparison.

]

.big[

**What should happen to the 3PA → points relationship if minutes played explains the original association?**

]

---

# Compare like with like

Work through Section 5 of the worksheet.

.big[

Instead of comparing every player to every other player, compare players with **similar amounts of playing time**.

]

---

# Evidence C

```{r evidence-c}
ggplot(nba_case, aes(x = fg3a, y = pts)) +
  geom_point(alpha = 0.55, size = 2.1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  facet_wrap(~ minutes_group, ncol = 2, scales = "free") +
  labs(
    title = "Three-point attempts and points, within minutes-played groups",
    subtitle = "Players are grouped by quartile of minutes played",
    x = "Three-point attempts",
    y = "Points scored"
  ) +
  theme_minimal(base_size = 17)
```

---

# What happened?

.big[

Within **every** minutes-played group, the relationship remains positive.

]

<div class="callout"><strong>Minutes played explains <em>some</em> of the pooled relationship, but it does <em>not</em> explain all of it.</strong></div>

---

# Return to the claim

.huge[

> **Players who attempt more three-pointers score more points because taking more threes makes them more effective scorers.**

]

Work through Section 6 of the worksheet **before** we discuss the conclusion.

---

# What can we say now?

.big[

**The evidence supports:**

- a positive association between three-point attempts and total points;
- the idea that playing time explains part of that association;
- a remaining positive relationship among players with similar playing time.

]

---

# What can we *not* say?

.big[

The evidence does **not** establish that taking more three-pointers makes players better scorers.

]

Possible remaining explanations include:

- player role or usage;
- shooting ability;
- position;
- team strategy;
- overall player quality;
- other variables we have not measured.

---

# The workflow

.huge[

**Claim → X/Y → Compare → Describe → Propose Z → Test Z → Compare like with like → Revise the conclusion**

]

This is the same reasoning pattern we will use in the Evidence Debate and Unit Test.

---

# Daily Quiz 8

.huge[

Now test yourself independently.

]

.big[

Open **Daily Quiz 8** in Brightspace.

]

.small[

No AI or internet access during the quiz.

]
