---
title: "How Can Unstructured Text Become Evidence?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

tweet_words <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/Trump_tweet_words.Rds") %>%
  mutate(PostPresident = Tweeting.date > as.Date("2016-11-06"))

odds <- tweet_words %>%
  count(word, PostPresident) %>%
  group_by(word) %>%
  filter(sum(n) >= 5) %>%
  ungroup() %>%
  pivot_wider(names_from = PostPresident, values_from = n, values_fill = 0,
              names_prefix = "post_") %>%
  mutate(
    tot_false = sum(post_FALSE),
    tot_true = sum(post_TRUE),
    prop_false = (post_FALSE + 1) / (tot_false + 1),
    prop_true = (post_TRUE + 1) / (tot_true + 1),
    odds_ratio = prop_true / prop_false,
    log_odds = log(odds_ratio)
  )

n_tweets <- tweet_words %>% distinct(document, PostPresident) %>% count(PostPresident)
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 17 opens Unit 5. It introduces the "bag of words" idea: a tokenized
document-word dataset can be counted, compared, and visualized using the
exact same tools (counts, proportions, ratios, logs for skew) already
established for numeric variables. Uses Trump_tweet_words.Rds, comparing
pre- and post-presidency tweets. Sentiment dictionaries are deliberately
NOT introduced here -- that is Lecture 18's job. This lecture stops at
"which words distinguish the two groups," a measurement/description task.

INSTRUCTOR NOTE:
No new modeling machinery (lm/glm) is used today. The core move is treating
word-group association as a variable, computed and audited the same way
every other variable has been all semester.
-->

# How Can Unstructured Text Become Evidence?

.big[
All semester:

> **How do we treat a number or a category as evidence?**

Today:

> **What about a paragraph, a tweet, or a transcript?**
]

---

# Where We Are

.big[
**Question → ... → find structure without labels → treat text as evidence**
]

Today:

**Treat text as evidence**

---

# A Familiar Kind of Question

.huge[
> **Did what this person talks about change after a specific event?**
]

Same structure as every claim this semester: a comparison, across a specific grouping variable.

---

# The Data: Before and After

```{r}
tweet_words %>%
  select(document, word, Tweeting.date, PostPresident) %>%
  head(6)
```

Every row is one **word**, from one **tweet**, tagged with whether it was posted before or after January 20, 2017.

---

# This Didn't Start as a Table

.big[
Raw text starts as a wall of characters:
]

> "Big day for Israel. Congratulations!"

.big[
To get the table on the last slide, someone had to **tokenize** it — split it into individual words, one row per word per document.
]

`tidytext::unnest_tokens()` does this automatically; we're starting from the already-tokenized result.

---

# The "Bag of Words"

.big[
Tokenizing throws away word order, grammar, and punctuation.
]

What's left is just: which words appeared, and how often.

.big[
That sounds like a huge loss — and it is. But it turns text into something we already know how to analyze: **counts**.
]

---

# Counting Words by Group

```{r}
tweet_words %>%
  filter(word %in% c("great", "sad", "fake", "border")) %>%
  count(word, PostPresident)
```

.big[
Same move as every group comparison this semester — just counting words instead of measuring a continuous variable.
]

---

# A Naive Comparison

.big[
Could we just compare **raw counts** of a word before vs. after?
]

```{r}
n_tweets
```

.big[
No — there are different numbers of tweets in each period. Raw counts would be unfair, exactly like comparing two groups of very different sizes anywhere else in the course.
]

---

# Fixing the Denominator: Proportions

.big[
Instead of "how many times," ask: **"what share of all words in this period was this word?"**
]

```{r, eval=FALSE}
tweet_words %>%
  count(word, PostPresident) %>%
  group_by(PostPresident) %>%
  mutate(prop = n / sum(n))
```

This is the same fix as turning a raw count into a rate.

---

# From Proportions to an Odds Ratio

.big[
To compare one word's usage across two periods in a single number:
]

.huge[
$$\text{odds ratio} = \frac{\text{proportion post-presidency}}{\text{proportion pre-presidency}}$$
]

A ratio above 1 means the word became more common; below 1, less common.

---

# Why We Need Logs Again

```{r}
odds %>%
  ggplot(aes(x = odds_ratio)) +
  geom_histogram(bins = 30)
```

.big[
Just like budgets and gross revenue earlier this semester: heavily skewed. A word that doubles in use and a word that is cut in half should look like mirror images — they don't, on this scale.
]

---

# Logging Fixes the Asymmetry

```{r}
odds <- odds %>% mutate(log_odds = log(odds_ratio))

odds %>%
  ggplot(aes(x = log_odds)) +
  geom_histogram(bins = 30)
```

Now doubling ($\log \approx 0.69$) and halving ($\log \approx -0.69$) are symmetric around zero.

---

# Which Words Distinguish the Two Periods?

```{r}
odds %>%
  group_by(log_odds > 0) %>%
  slice_max(abs(log_odds), n = 10) %>%
  ungroup() %>%
  mutate(word = reorder(word, log_odds)) %>%
  ggplot(aes(x = log_odds, y = word, fill = log_odds < 0)) +
  geom_col() +
  labs(x = "Post- / pre-presidency log-odds", y = NULL) +
  theme(legend.position = "none")
```

---

# What This Chart Does and Doesn't Show

.big[
It shows: which specific words became relatively more or less common after a known date, adjusting for the different number of tweets in each period.
]

It does **not** show:

- what the person **meant** by using those words;
- whether the shift reflects the event, or just time passing;
- anything about **tone** — only **topic**.

<!-- Bridge to Lecture 18: sentiment vs. topic. -->

---

# Topic Is Not the Same as Tone

.big[
A word like "fake" appearing more often tells us a **topic** became more prominent.
]

It doesn't tell us whether the person's overall tone became more positive or negative — that requires a different kind of measurement, which is next class's question.

---

# Audit an AI Claim

Suppose an AI assistant is given only the raw word counts (not proportions) and says:

> "This person clearly talks about the border constantly now — look how many more times 'border' appears."

.big[
**Does a higher raw count account for the number of tweets in each period?**

**What would a fair comparison require?**

**Is a bigger raw count the same as a bigger share of what was said?**
]

A raw count comparison across unequal group sizes is the same mistake this course has flagged all semester — just applied to words instead of numbers.

---

# A Calibrated Conclusion

Instead of:

> "This shows what this person really thinks."

Try:

.big[
> **Relative to the number of tweets in each period, certain words became substantially more or less common after this date — a measurable shift in topic, not a direct measurement of intent, tone, or meaning.**
]

---

# Before You Leave

When you see a claim based on text data, ask:

1. Was the comparison based on raw counts, or adjusted for how much text there was in each group?
2. Does a log transformation make an asymmetric ratio easier to compare?
3. Does the measure capture *topic* (what's discussed) or *tone* (how it's discussed)?
4. What did tokenizing the text throw away, and does that matter here?
5. What would this measure not tell you, even if the numbers are calculated correctly?

---

# The Rule to Remember

.huge[
**Text becomes data the moment you start counting it.**

**Counting is not the same as understanding.**
]

---

# Homework 13

You will practice:

- tokenizing your intuition about what "bag of words" throws away;
- calculating and interpreting a log-odds ratio for word usage across groups;
- auditing an unfair raw-count comparison in text data;
- distinguishing what a word-frequency measure captures from what it misses.

Next:

.big[
**NLP / Sentiment**
]
