---
title: "How Confident Should We Be?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

nba <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/nba_players_2018.Rds")

# Bootstrap distribution used later in the deck.
set.seed(123)
bs_tov <- NULL
for (i in 1:200) {
  tmp <- nba %>%
    sample_n(size = nrow(nba), replace = TRUE) %>%
    group_by(isRookie) %>%
    summarise(mean_tov = mean(tov, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_tov <- bind_rows(bs_tov, tmp)
}

bs_diff <- bs_tov %>%
  pivot_wider(names_from = isRookie, values_from = mean_tov, names_prefix = "rookie_") %>%
  mutate(diff = rookie_FALSE - rookie_TRUE)

conf_tov <- mean(bs_diff$diff > 0)

# Turnovers-per-minute version, used to audit the raw-count claim.
nba <- nba %>% mutate(tov_rate = tov * 60 / minutes)

set.seed(123)
bs_rate <- NULL
for (i in 1:200) {
  tmp <- nba %>%
    sample_n(size = nrow(nba), replace = TRUE) %>%
    group_by(isRookie) %>%
    summarise(mean_rate = mean(tov_rate, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_rate <- bind_rows(bs_rate, tmp)
}

bs_rate_diff <- bs_rate %>%
  pivot_wider(names_from = isRookie, values_from = mean_rate, names_prefix = "rookie_") %>%
  mutate(diff = rookie_FALSE - rookie_TRUE)

conf_rate <- mean(bs_rate_diff$diff > 0)
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 9 opens Unit 3. It introduces sampling variability and bootstrap
resampling as a way to express how much a single sample should make us
uncertain about a conclusion, without requiring formal sampling-distribution
theory. Lecture 10 (Uncertainty II) builds on this to ask what a confidence
statement does and does not justify.

INSTRUCTOR NOTE:
The running example is NBA player data (nba_players_2018.Rds), reused from
the ungraded Test 1 practice homework. Keep this deck's reasoning connected
back to the "association vs. explanation" habits built in Unit 2.
-->

# How Confident Should We Be?

.big[
So far:

> **Given this sample, what relationship do I see, and what else could explain it?**

Today:

> **If I had drawn a different sample, would I still see it?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → interrogate explanations → quantify uncertainty → communicate**
]

Today:

**Quantify uncertainty**

---

# One Sample, One Story

Every analysis so far has used **one sample**:

- one set of colleges;
- one set of Michigan voters;
- one set of movies.

.big[
What if we had happened to sample differently?
]

---

# A New Question

.huge[
> **Do NBA rookies turn the ball over more than non-rookies?**
]

We have the data to check this directly.

---

# Look First

```{r}
nba %>%
  select(namePlayer, isRookie, tov, minutes) %>%
  glimpse()
```

---

# The Comparison

```{r}
nba %>%
  group_by(isRookie) %>%
  summarise(mean_tov = mean(tov, na.rm = TRUE))
```

.big[
Rookies average more turnovers in **this sample**.
]

---

# A Skeptic's Question

.big[
Is this difference **real**, or could it easily have come out the other way with a slightly different sample?
]

This is a new kind of question.

It is not "what else could explain it?" (Unit 2).

It is: **how much should this one sample make us believe anything at all?**

<!-- LO1: Distinguish confounding (Unit 2) from sampling uncertainty (Unit 3). -->

---

# Populations and Samples

.big[
**Population:** every player-season we would want to generalize to.

**Sample:** the players who happen to be in our dataset.
]

Even if the population pattern is real, any one sample can look a little different just by chance.

---

# We Cannot Re-Run the Season

To directly measure how much samples vary, we would want to:

.big[
Watch the 2018–19 NBA season play out **many times** and recompute the comparison each time.
]

That is impossible. There was only one season.

---

# The Next Best Thing

.big[
**Bootstrap resampling:** simulate "redoing the sample" by resampling our own data, with replacement.
]

We cannot rerun history. We can mimic the process of sampling.

---

# What "With Replacement" Means

```{r}
set.seed(123)

nba %>%
  sample_n(size = 5, replace = TRUE) %>%
  select(namePlayer, isRookie, tov)
```

Some players can appear more than once. Others may not appear at all.

---

# One Resampled "Season"

```{r}
set.seed(123)

nba %>%
  sample_n(size = nrow(nba), replace = TRUE) %>%
  group_by(isRookie) %>%
  summarise(mean_tov = mean(tov, na.rm = TRUE))
```

A full resample: same size as our data, drawn with replacement.

---

# A Second Resampled "Season"

```{r}
set.seed(456)

nba %>%
  sample_n(size = nrow(nba), replace = TRUE) %>%
  group_by(isRookie) %>%
  summarise(mean_tov = mean(tov, na.rm = TRUE))
```

.big[
A slightly different answer, just from resampling.
]

---

# Doing This Many Times

Repeating this by hand would take forever.

A `for()` loop repeats the same block of code automatically:

```{r, eval=FALSE}
bs_tov <- NULL
for (i in 1:200) {
  tmp <- nba %>%
    sample_n(size = nrow(nba), replace = TRUE) %>%
    group_by(isRookie) %>%
    summarise(mean_tov = mean(tov, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_tov <- bind_rows(bs_tov, tmp)
}
```

.big[
200 simulated "seasons," each with its own comparison.
]

---

# 200 Simulated Comparisons

```{r}
bs_tov %>%
  head(6)
```

Each `sim` is one resampled dataset's version of the comparison.

---

# Visualize the Spread

```{r, echo=FALSE}
bs_tov %>%
  ggplot(aes(x = mean_tov, fill = isRookie)) +
  geom_density(alpha = .4) +
  labs(x = "Mean turnovers", y = "Density across 200 resamples",
       title = "How much does the rookie/non-rookie comparison move around?")
```

---

# From Two Numbers to One Question

For each resample, ask a yes/no question:

.big[
**Did rookies average more turnovers than non-rookies in this resample?**
]

```{r}
bs_diff %>%
  select(sim, rookie_FALSE, rookie_TRUE, diff) %>%
  head(4)
```

`diff` = non-rookie mean − rookie mean. Positive means rookies had more turnovers.

---

# Turning This Into "Confidence"

```{r}
bs_diff %>%
  summarise(confidence = mean(diff > 0))
```

.big[
Rookies had more turnovers in `r round(conf_tov*100,1)`% of the 200 resamples.
]

That percentage **is** our confidence measure.

---

# What Does That Number Mean?

.huge[
`r round(conf_tov*100,1)`% confidence
]

**Not:** "there is a `r round(conf_tov*100,1)`% chance rookies turn the ball over more."

**Is:** "in `r round(conf_tov*100,1)`% of plausible resamples of this data, the pattern held up."

<!-- LO6: Keep the language tied to what resampling actually measures. -->

---

# Calibrating Against Extremes

.big[
A coin flip would give about **50%**.

Confidence near 50% → the pattern is fragile; a different sample could easily flip it.

Confidence near 100% → the pattern shows up almost no matter how we resample.
]

`r round(conf_tov*100,1)`% is closer to one end than the other. Which one?

---

# Before You Celebrate

.big[
High resampling confidence does **not** mean the explanation is right.
]

It only means: *this pattern in this sample is not fragile to resampling.*

It says nothing about **why** the pattern exists.

<!-- Bridge back to Unit 2 reasoning. -->

---

# Is the Comparison Even Fair?

Rookies and non-rookies do not play the same number of minutes.

```{r}
nba %>%
  group_by(isRookie) %>%
  summarise(mean_minutes = mean(minutes, na.rm = TRUE))
```

.big[
Fewer minutes → fewer *opportunities* to turn the ball over.
]

---

# Re-Measuring the Claim

```{r}
nba <- nba %>%
  mutate(tov_rate = tov * 60 / minutes)

nba %>%
  group_by(isRookie) %>%
  summarise(mean_rate = mean(tov_rate, na.rm = TRUE))
```

Turnovers **per 60 minutes**, instead of raw counts.

---

# Bootstrap the Fairer Comparison

```{r, eval=FALSE}
bs_rate <- NULL
for (i in 1:200) {
  tmp <- nba %>%
    sample_n(size = nrow(nba), replace = TRUE) %>%
    group_by(isRookie) %>%
    summarise(mean_rate = mean(tov_rate, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_rate <- bind_rows(bs_rate, tmp)
}
```

```{r}
bs_rate_diff %>%
  summarise(confidence = mean(diff > 0))
```

---

# Two Confidence Numbers

.big[
Raw turnovers: `r round(conf_tov*100,1)`% confidence rookies turn it over more.

Turnovers per 60 minutes: `r round(conf_rate*100,1)`% confidence rookies turn it over more.
]

Adjusting for playing time changed how confident we should be.

<!-- LO1: Confounding and uncertainty compound; this is not a new bug, it's the same habit from Unit 2 applied to a resampled quantity. -->

---

# Audit an AI Claim

Suppose an AI assistant is given only the raw-turnover bootstrap and says:

> "This proves rookies are worse ball-handlers than veterans."

.big[
**What evidence supports this?**

**What did the claim ignore?**

**Would the AI's confidence survive adjusting for minutes played?**
]

A resampling percentage is not proof, and it does not audit itself.

---

# What Bootstrap Confidence Is and Isn't

.big[
**Is:** a measure of how much a specific comparison, in this specific dataset, moves around under resampling.

**Is not:** a claim about the underlying causal explanation.

**Is not:** a guarantee about next season, a different league, or a different set of players.
]

---

# A Calibrated Conclusion

Instead of:

> "Rookies turn the ball over more."

Try:

.big[
> **In these 2018–19 data, rookies average more turnovers per 60 minutes than non-rookies in about `r round(conf_rate*100,1)`% of resamples of this season — a pattern that is not simply an artifact of playing time, though it does not tell us whether it would hold in other seasons.**
]

---

# Before You Leave

When you see a comparison in a sample, ask:

1. What comparison am I making, and in what sample?
2. How much would that comparison move around if I resampled the data?
3. What percentage of resamples support the comparison?
4. Is that percentage close to a coin flip, or close to certainty?
5. Have I adjusted for an obvious confound before trusting the number?
6. What can this confidence number **not** tell me?

---

# The Rule to Remember

.huge[
**A sample gives one story.**

**Resampling shows how fragile that story is.**
]

---

# Homework 7

You will practice:

- writing and interpreting a bootstrap resampling loop;
- turning resampled estimates into a confidence percentage;
- calibrating what a confidence percentage does and does not mean;
- checking whether a resampling result survives an obvious adjustment;
- auditing an AI's overclaimed confidence.

Next:

.big[
**Uncertainty II**
]
