---
title: "Prediction vs. Explanation"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

gms <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/game_summary.Rds")

# Linear probability model: predicting a binary outcome with lm().
model_lpm <- lm(isWin ~ tov, data = gms)
gms$pred_lpm <- predict(model_lpm)

n_below0 <- sum(gms$pred_lpm < 0, na.rm = TRUE)
n_above1 <- sum(gms$pred_lpm > 1, na.rm = TRUE)

# A "kitchen sink" prediction-oriented model, evaluated by held-out accuracy.
set.seed(123)
inds <- sample(1:nrow(gms), size = round(nrow(gms) / 2), replace = FALSE)
train <- gms %>% slice(inds)
test <- gms %>% slice(-inds)

model_predict <- lm(isWin ~ tov + pts + treb + pctFG, data = train)
test$pred <- predict(model_predict, newdata = test)
test$pred_class <- ifelse(test$pred >= 0.5, TRUE, FALSE)
accuracy <- mean(test$pred_class == test$isWin, na.rm = TRUE)
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 13 opens Unit 4. It does not teach a new model type -- it reframes
the goal. Units 1-3 built habits for explanation (confounders, calibrated
causal language, model summaries). This lecture argues that when the goal
shifts to prediction, the evaluation criterion shifts too (from "is this
variable justified?" to "does this improve held-out accuracy?"), and uses
a linear probability model on a binary outcome (isWin) to motivate why
Classification I needs new tools.

INSTRUCTOR NOTE:
No new lecture-adjacent homework technique is introduced here; the demo is
deliberately a "broken" use of lm() on a 0/1 outcome, to make the case for
classification-specific methods next class.
-->

# Prediction vs. Explanation

.big[
So far:

> **What causes this relationship, and how confident should I be in that explanation?**

Today:

> **What if I don't need to explain anything — I just need to guess right?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → interrogate explanations → quantify uncertainty → summarize with a model → evaluate the model → choose the right goal**
]

Today:

**Choose the right goal**

---

# Two Different Jobs

.big[
**Explanation:** Why does this relationship exist? Which variables matter, and why?

**Prediction:** Given what I know, what is my best guess for the outcome?
]

Both use models. They are not the same job.

---

# A Familiar Example, Revisited

Units 1–3 asked:

.huge[
> **Does selectivity explain graduate earnings?**
]

We worried about confounders, calibrated our language, and avoided claiming causation from an association.

That is an **explanation** question.

---

# A Different Kind of Question

.huge[
> **Given a team's turnovers, points, rebounds, and shooting, can I guess who wins the game?**
]

Nobody asked *why* turnovers relate to winning. The only thing that matters is: **did the guess turn out to be right?**

That is a **prediction** question.

---

# What Changes Between the Two Goals

.big[
**For explanation:** every variable needs a justification. A confounder is a threat to the conclusion. Interpretability matters.

**For prediction:** a variable earns its place by improving accuracy on new data. It doesn't need a causal story. Interpretability is nice, not required.
]

---

# The Same Habit Still Applies

.big[
Whichever goal you have, you still have to **evaluate the model on data it wasn't fit to** — that's Lecture 12's cross-validation lesson, and it applies just as much to prediction as to explanation.
]

What changes is: evaluate it **for what**?

---

# Today's Data

```{r}
gms %>%
  select(idGame, isWin, tov, pts, treb, pctFG) %>%
  glimpse()
```

`isWin`: did the team win this game? `TRUE` / `FALSE`.

---

# A New Kind of Outcome

.big[
Every outcome we have modeled so far has been **continuous**: earnings, turnovers.

`isWin` is different: it only takes two values.
]

What happens if we just plug it into `lm()` like we always have?

---

# Trying the Familiar Tool

```{r}
model_lpm <- lm(isWin ~ tov, data = gms)
summary(model_lpm)$coefficients
```

.big[
R doesn't stop us. It fits a line through 0s and 1s.
]

---

# What Does the Line Predict?

```{r}
gms %>%
  ggplot(aes(x = tov, y = isWin)) +
  geom_jitter(height = .05, alpha = .2) +
  geom_smooth(method = "lm", se = FALSE)
```

---

# A Real Problem

```{r}
gms$pred_lpm <- predict(model_lpm)
range(gms$pred_lpm, na.rm = TRUE)
```

.big[
`r n_below0` predicted values are below 0. `r n_above1` are above 1.
]

A "predicted probability" of winning that is negative, or over 100%, doesn't mean anything.

---

# Why This Happens

.big[
`lm()` assumes $Y$ can take any value and tries to minimize squared error across the whole range.
]

A binary outcome isn't built that way — it only lives at 0 and 1. The straight-line tool from Unit 3 wasn't built for this job.

<!-- Bridge slide: sets up Classification I's need for a bounded-probability model. -->

---

# The Goal Was Never "Fit a Line"

.big[
Step back. What did we actually want?
]

Not a coefficient to interpret. We wanted: **given a new game's stats, guess win or lose, and be right as often as possible.**

That reframing is what Classification I will build tools for.

---

# Redefining "Good," for Prediction

.big[
For a continuous outcome, we evaluated with RMSE: how far off, on average?

For a win/lose outcome, "how far off" isn't quite the right question.
]

A more natural question: **out of all the games where I made a guess, how often was I right?**

---

# A Rough First Pass: Accuracy

Using a 50% cutoff on our (flawed) linear model's predictions, split into train and test:

```{r, eval=FALSE}
model_predict <- lm(isWin ~ tov + pts + treb + pctFG, data = train)
test$pred <- predict(model_predict, newdata = test)
test$pred_class <- ifelse(test$pred >= 0.5, TRUE, FALSE)
accuracy <- mean(test$pred_class == test$isWin, na.rm = TRUE)
```

```{r}
round(accuracy * 100, 1)
```

---

# Prediction Still Needs Held-Out Evaluation

.big[
Notice: we still split into train and test, exactly like Lecture 12.
]

The goal changed (accuracy instead of RMSE). The discipline of testing on data the model hasn't seen did not change.

---

# Kitchen-Sink Predictors

.big[
Because the goal is prediction, not explanation, we didn't ask whether `treb` (rebounds) or `pctFG` "should" belong in the model theoretically.
]

We asked only: **does including it help the held-out guess?**

That would be a very different move if our goal were explanation.

---

# A Genuine Tension

.big[
A model built purely to predict well can end up **uninterpretable** — dozens of predictors, no clean story.

A model built purely to explain can end up a **poor predictor** — clean and justified, but not very accurate on new data.
]

Data scientists have to know which job they were hired to do.

---

# Consequences Are Not Symmetric

.big[
Prediction models are often used to make real decisions about people: who gets a loan, who gets flagged, who gets hired, who gets released.
]

When a model's predictions carry consequences, "it's accurate" is not automatically the same as "it's fair to use this way." We will return to this directly in Unit 5.

---

# Audit an AI Claim

Suppose an AI assistant builds a model to predict wins from turnovers, points, rebounds, and shooting percentage, achieves 78% accuracy, and says:

> "This proves that these four statistics are what cause a team to win."

.big[
**What was the model actually built to do?**

**Does high predictive accuracy establish a causal mechanism?**

**What would you need in order to make an explanatory claim instead?**
]

A model tuned for prediction was never asked to identify causes.

---

# A Calibrated Conclusion

Instead of:

> "These stats are what cause winning."

Try:

.big[
> **A model using turnovers, points, rebounds, and shooting percentage correctly predicted the outcome of about `r round(accuracy*100,1)`% of held-out games — a measure of how useful these variables are for guessing the outcome, not a claim about what causes it.**
]

---

# Before You Leave

Before building any model, ask:

1. Is the goal to explain *why*, or to predict *what*?
2. If explanation: is each variable theoretically justified? What could confound it?
3. If prediction: does each variable improve held-out performance?
4. Does the outcome variable's type (continuous vs. categorical) match the tool I'm about to use?
5. If the prediction will be used to make a real decision, what are the consequences of being wrong?

---

# The Rule to Remember

.huge[
**Explanation asks why. Prediction asks what.**

**Know which question you are actually answering.**
]

---

# Homework 10

You will practice:

- distinguishing an explanation question from a prediction question;
- recognizing when a continuous-outcome tool has been misapplied to a categorical outcome;
- evaluating a predictive claim using held-out performance;
- auditing an AI claim that treats predictive accuracy as causal proof.

Next:

.big[
**Classification I**
]
