---
title: "What Do Text Measures Capture, and What Do They Miss?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

tweet_words <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/Trump_tweet_words.Rds") %>%
  mutate(PostPresident = Tweeting.date > as.Date("2016-11-06"))

nrc <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/nrc.Rds")

n_total_words <- tweet_words %>% distinct(word) %>% nrow()
n_matched_words <- tweet_words %>% distinct(word) %>% inner_join(nrc, by = "word") %>% distinct(word) %>% nrow()
pct_matched <- n_matched_words / n_total_words

tweet_sentiment <- tweet_words %>%
  inner_join(nrc, by = "word") %>%
  filter(sentiment %in% c("positive", "negative"))

tweet_sentiment_summary <- tweet_sentiment %>%
  count(document, PostPresident, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(net_sentiment = positive - negative)

sentiment_means <- tweet_sentiment_summary %>%
  group_by(PostPresident) %>%
  summarise(mean_sentiment = mean(net_sentiment), .groups = "drop")
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 18 completes the text-as-data pair. Lecture 17 measured TOPIC
(which words became more/less common). This lecture measures TONE via a
sentiment dictionary (nrc.Rds), but the pedagogical center of gravity is
the construct-validity audit: dictionary coverage is incomplete, negation
and sarcasm break word-level scoring, and a crude sentiment count is not
the same as knowing what someone felt. This is the same "what does this
variable actually measure" habit from Unit 1, applied to text.

INSTRUCTOR NOTE:
Keep the bootstrap-confidence callback to Unit 3 light (one slide, no new
loop needs to be built) -- the point is measurement validity, not another
uncertainty exercise.
-->

# What Do Text Measures Capture, and What Do They Miss?

.big[
Last class:

> **Which words became more common — a measure of topic.**

Today:

> **Can we measure tone — and what happens when we try?**
]

---

# Where We Are

.big[
**Question → ... → treat text as evidence → measure meaning, not just topic**
]

Today:

**Measure meaning, not just topic**

---

# A New Kind of Question

.huge[
> **Did this person's tone become more positive or negative after a specific date?**
]

Lecture 17's log-odds chart told us **what** shifted. It said nothing about whether the shift felt positive or negative.

---

# Measuring Tone: Sentiment Dictionaries

.big[
A **sentiment dictionary** is a prepared list of words, each tagged with an emotion — built by researchers, not by us.
]

```{r}
nrc %>% distinct(sentiment)
```

---

# What's Inside

```{r}
nrc %>% filter(sentiment %in% c("positive", "negative")) %>% sample_n(8)
```

.big[
Every word in this list was hand-labeled, in advance, by people who were not looking at our tweets.
]

---

# Attaching Sentiment to Our Words

```{r}
tweet_sentiment <- tweet_words %>%
  inner_join(nrc, by = "word") %>%
  filter(sentiment %in% c("positive", "negative"))

tweet_sentiment %>% select(document, word, sentiment) %>% head(5)
```

`inner_join()` keeps only the words that appear in **both** our data and the dictionary.

---

# The First Problem, Immediately

.big[
What happened to every word that **isn't** in the dictionary?
]

```{r}
round(pct_matched * 100, 1)
```

.big[
Only about `r round(pct_matched*100,1)`% of the distinct words in our data even appear in the sentiment dictionary at all.
]

<!-- LO1: Coverage limitation surfaces before any scoring happens. -->

---

# Most of What Was Said Is Invisible to This Measure

.big[
Names, places, slang, numbers, most ordinary words — none of them carry a sentiment tag.
]

Whatever score we calculate next is built entirely from the small slice of words the dictionary happens to recognize.

---

# Scoring a Tweet

.big[
For each tweet, count positive words and negative words, then take the difference.
]

```{r}
tweet_sentiment_summary <- tweet_sentiment %>%
  count(document, PostPresident, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(net_sentiment = positive - negative)

tweet_sentiment_summary %>% head(5)
```

---

# Comparing Before and After

```{r}
tweet_sentiment_summary %>%
  ggplot(aes(x = PostPresident, y = net_sentiment)) +
  geom_boxplot()
```

---

# The Averages

```{r}
sentiment_means
```

.big[
Is this a meaningful difference, or could it show up just from resampling noise? That's exactly the Unit 3 question — the tools from Lectures 9–10 apply here too, unchanged.
]

---

# Before We Trust the Score: What Does It Miss?

.big[
A dictionary score counts word matches. It does not read the sentence.
]

Three specific failure modes are worth knowing by name.

---

# Failure 1: Negation

```{r}
tibble(text = c("This is great.", "This is not great.")) %>%
  mutate(naive_sentiment = str_detect(text, "great"))
```

.big[
Both sentences contain "great." A word-count method scores them **the same** — even though negation flips the meaning entirely.
]

---

# Failure 2: Sarcasm and Context

.big[
"Oh, great. Another delay."
]

.big[
Every word-matching method scores "great" as positive here, with no way to detect that the sentence is sarcastic.
]

Context, tone of voice, and shared knowledge are exactly what a bag-of-words approach throws away.

---

# Failure 3: Domain Mismatch

.big[
Dictionaries are built from **general** language use, not from the specific domain you're studying.
]

A word that reads as neutral or even positive in everyday speech (e.g., "aggressive marketing") might carry a different connotation in politics, sports commentary, or finance.

---

# This Is a Construct-Validity Question

.big[
Recall from Unit 1: before trusting a variable, ask **what it actually measures**.
]

"Net sentiment score" sounds like it measures **feeling**. What it actually measures is: **the count of dictionary-matched positive words minus dictionary-matched negative words, for the roughly `r round(pct_matched*100,1)`% of words the dictionary recognizes.**

Those are not the same thing.

---

# So Is the Measure Useless?

.big[
No — but it is a **proxy**, not a direct read of emotion.
]

It can still be useful for large-scale comparisons, as long as you communicate what it is: an imperfect, dictionary-based approximation, not a direct measurement of what someone felt.

---

# Audit an AI Claim

Suppose an AI assistant computes the net sentiment scores and says:

> "This proves the person's true emotional state became significantly more negative."

.big[
**What does the dictionary actually measure?**

**What share of the original words were even scored?**

**What would negation or sarcasm do to this score, and did the AI check for it?**
]

A sentiment score is a measurement, with all the limitations any measurement has — it is not direct access to what someone felt.

---

# A Calibrated Conclusion

Instead of:

> "This proves the person felt worse."

Try:

.big[
> **Using a dictionary that matched about `r round(pct_matched*100,1)`% of the words in these tweets, average net sentiment was lower after this date than before — a dictionary-based proxy for tone that cannot account for negation, sarcasm, or the words it fails to recognize at all.**
]

---

# Before You Leave

When you see a sentiment score, ask:

1. What dictionary was used, and what share of the text did it actually match?
2. Could negation or sarcasm be flipping some of these scores?
3. Was the dictionary built for this domain, or a general one?
4. Is the score being reported as a proxy, or oversold as direct access to feeling?
5. Does a difference between groups look real, or could it be resampling noise (Unit 3)?

---

# The Rule to Remember

.huge[
**A sentiment score measures word matches.**

**It does not measure feelings.**
]

---

# Homework 14

You will practice:

- joining text data to a sentiment dictionary and measuring coverage;
- calculating and comparing net sentiment across groups;
- identifying negation or sarcasm failures in specific examples;
- auditing an AI claim that treats a sentiment score as direct access to emotion.

Next:

.big[
**Bias & Misinformation**
]
