---
title: "What Can We Learn When Groups Aren't Given?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

h97 <- read_csv("https://raw.githubusercontent.com/jbisbee1/DS1000_F2026/main/data/H097_members.csv") %>%
  filter(chamber %in% c("House", "Senate")) %>%
  mutate(party = case_when(
    party_code == 100 ~ "D",
    party_code == 200 ~ "R",
    TRUE ~ "I"
  )) %>%
  select(bioname, party, nominate_dim1, nominate_dim2) %>%
  drop_na()

h117 <- read_csv("https://raw.githubusercontent.com/jbisbee1/DS1000_F2026/main/data/H117_members.csv") %>%
  filter(chamber %in% c("House", "Senate")) %>%
  mutate(party = case_when(
    party_code == 100 ~ "D",
    party_code == 200 ~ "R",
    TRUE ~ "I"
  )) %>%
  select(bioname, party, nominate_dim1, nominate_dim2) %>%
  drop_na()

set.seed(123)
km2_97 <- kmeans(h97 %>% select(nominate_dim1, nominate_dim2), centers = 2, nstart = 25)

wss_97 <- NULL
for (k in 1:8) {
  m_k <- kmeans(h97 %>% select(nominate_dim1, nominate_dim2), centers = k, nstart = 25)
  wss_97 <- bind_rows(wss_97, tibble(k = k, tot_wss = m_k$tot.withinss))
}

set.seed(123)
km2_117 <- kmeans(h117 %>% select(nominate_dim1, nominate_dim2), centers = 2, nstart = 25)
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 16 closes Unit 4. It introduces k-means clustering as a third "camp"
distinct from explanation (Units 1-3) and prediction (Lecture 13-15): no
outcome variable, no labeled groups, just structure. Uses DW-NOMINATE
ideology scores (H097/H117_members.csv) so students can compare Congresses
83 years apart on a question the course cares about (polarization).

INSTRUCTOR NOTE:
This is the last instructional lecture before the Unit 4 Assessment Day
(Evidence Debate 4 + Unit Test 4). Per the established course pattern, it
closes with a Daily Quiz callout, not a "Homework N" slide -- the next
homework (HW13) belongs to Lecture 17 (Text as Data, the Unit 5 opener).
-->

# What Can We Learn When Groups Aren't Given?

.big[
So far this unit:

> **Can I predict an outcome I already have labels for?**

Today:

> **What if nobody handed me any labels at all?**
]

---

# Where We Are

.big[
**Question → ... → build a classifier → evaluate the tradeoffs → find structure without labels**
]

Today:

**Find structure without labels**

---

# A Third Camp

This semester has had two goals so far:

.big[
**Explanation** (Units 1–3): is X associated with Y, and why?

**Prediction** (Lectures 13–15): given X, what is Y?
]

Both need a $Y$ — an outcome to explain or predict.

---

# What If There's No $Y$?

.huge[
> **Are there natural groups in this data — and what are they?**
]

No claim to test. No outcome to guess. Just: **does this data have structure we didn't put there ourselves?**

This is sometimes called **unsupervised learning** — "unsupervised" because there's no known right answer to check against.

---

# Today's Data: Congress

```{r}
h97 %>% head(5)
```

`nominate_dim1` / `nominate_dim2`: a measure of how a member of Congress's voting record compares to everyone else's, compressed into two numbers. Higher `dim1` generally means more conservative voting.

---

# Look First

```{r}
h97 %>%
  ggplot(aes(x = nominate_dim1, y = nominate_dim2)) +
  geom_point(alpha = .5)
```

.big[
Do you see groups? How many?
]

---

# Letting an Algorithm Find Groups

.big[
**$k$-means clustering:** an algorithm that sorts observations into $k$ groups based on how close together they are.
]

"Close" means **Euclidean distance** — the straight-line distance between two points, using every variable you give it.

---

# Distance in Two Dimensions

```{r}
pt1 <- c(0, 0)
pt2 <- c(3, 4)

sqrt(sum((pt1 - pt2)^2))
```

.big[
Same idea as the distance formula from geometry: $\sqrt{(x_2-x_1)^2 + (y_2-y_1)^2}$ — just applied to `nominate_dim1` and `nominate_dim2` instead of $x$ and $y$.
]

---

# How $k$-means Works

.big[
1. Randomly place $k$ **centroids** (candidate group centers).
2. Assign every observation to its **nearest** centroid.
3. Move each centroid to the **average** position of the points assigned to it.
4. Repeat steps 2–3 until nothing changes.
]

---

# Running It

```{r}
km2_97 <- kmeans(h97 %>% select(nominate_dim1, nominate_dim2),
                  centers = 2, nstart = 25)

km2_97$centers
```

`centers = 2` asks for two groups. `nstart = 25` reruns the random start 25 times and keeps the best result — the starting points are random, so one run alone can get unlucky.

---

# Visualizing the Clusters

```{r}
h97 %>%
  mutate(cluster = factor(km2_97$cluster)) %>%
  ggplot(aes(x = nominate_dim1, y = nominate_dim2, color = cluster)) +
  geom_point(alpha = .6)
```

---

# Does This Match Something We Know?

```{r}
h97 %>%
  mutate(cluster = factor(km2_97$cluster)) %>%
  count(cluster, party)
```

.big[
The algorithm was never told anyone's party. Does the structure it found line up with party anyway?
]

---

# How Many Clusters Should We Ask For?

.big[
Nothing stops us from asking for 3, 5, or 20 groups. How do we choose?
]

Just like regression, we can measure **error**: how far each point sits from its own cluster's center.

---

# Error in Clustering: WSS

.big[
**Within Sum of Squares (WSS):** total squared distance from each point to its cluster's centroid, added up.
]

More clusters will always reduce WSS — in the extreme, every point could be its own cluster with zero error. That's not useful. We need a tradeoff.

---

# Looking for the "Elbow"

```{r, eval=FALSE}
wss_97 <- NULL
for (k in 1:8) {
  m_k <- kmeans(h97 %>% select(nominate_dim1, nominate_dim2),
                centers = k, nstart = 25)
  wss_97 <- bind_rows(wss_97, tibble(k = k, tot_wss = m_k$tot.withinss))
}
```

```{r}
wss_97 %>%
  ggplot(aes(x = k, y = tot_wss)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:8)
```

---

# Reading the Elbow Plot

.big[
WSS keeps falling as $k$ grows — but the *rate* of improvement slows down.
]

Look for the point where adding another cluster stops buying you much: the "elbow." Past that point, you're mostly just fitting noise.

---

# A Judgment Call, Not a Formula

.big[
The elbow method is a guide, not a rule. Two people can look at the same plot and reasonably pick different values of $k$.
]

The right choice also depends on what you're trying to learn — two broad groups tell a different story than five narrow ones.

---

# A Question the Course Cares About

.huge[
> **Has political polarization increased?**
]

Compare the 97th Congress (1981–83) to the 117th Congress (2021–23), 40 years later.

---

# 97th Congress, Two Clusters

```{r,echo=FALSE}
h97 %>%
  mutate(cluster = factor(km2_97$cluster)) %>%
  ggplot(aes(x = nominate_dim1, y = nominate_dim2, color = cluster)) +
  geom_point(alpha = .6) +
  labs(title = "97th Congress (1981-1983)")
```

---

# 117th Congress, Two Clusters

```{r,echo=FALSE}
h117 %>%
  mutate(cluster = factor(km2_117$cluster)) %>%
  ggplot(aes(x = nominate_dim1, y = nominate_dim2, color = cluster)) +
  geom_point(alpha = .6) +
  labs(title = "117th Congress (2021-2023)")
```

---

# What Changed?

.big[
Look at the two plots: are the clusters closer together, farther apart, more separated, or more overlapping?
]

Clustering didn't test the polarization claim for us. It gave us a structured way to **look**.

---

# What Clustering Does Not Do

.big[
It does not tell us *why* two clusters exist.

It does not tell us whether the grouping is the "correct" one — a different $k$, or different variables, could reveal a different structure.

It does not give us a $p$-value or a confidence percentage the way earlier units did.
]

<!-- LO1: Learning is a genuinely different epistemic mode from explanation and prediction. -->

---

# Audit an AI Claim

Suppose an AI assistant runs $k$-means with $k = 2$ on the 117th Congress and says:

> "This proves there are exactly two political ideologies in America."

.big[
**Did the algorithm test that claim, or assume it by choosing $k=2$?**

**What would change if we had asked for $k=3$ instead?**

**What is the difference between "the algorithm found two clusters" and "there are only two ideologies"?**
]

---

# A Calibrated Conclusion

Instead of:

> "This proves there are two political camps."

Try:

.big[
> **Using two-dimensional ideology scores, a $k$-means clustering with $k=2$ groups members of the 117th Congress in a way that closely tracks party — but this reflects one modeling choice among several reasonable ones, not proof that exactly two ideological camps exist.**
]

---

# Before You Leave

When someone shows you a clustering result, ask:

1. What variables define "closeness" here, and why those?
2. How was $k$ chosen — an elbow plot, a theory, or a guess?
3. Would a different $k$ tell a meaningfully different story?
4. Does the clustering line up with something you already understand, as a sanity check?
5. Is this being used to *describe* structure, or overinterpreted as an *explanation*?

---

# The Rule to Remember

.huge[
**Clustering finds structure. It does not explain it, and it does not predict anything.**

**Know which of the three jobs you're actually doing.**
]

---

# The Unit 4 Workflow, Complete

.big[
**Ask: explain, predict, or learn?**

↓

**Explain: summarize with a model, evaluate it out of sample**

↓

**Predict: classify, evaluate with a confusion matrix, ROC, and AUC**

↓

**Learn: find structure with no labels, choose $k$ deliberately, don't overclaim what it means**
]

---

# Daily Quiz 16

.huge[

Now test yourself independently.

]

.big[

Open **Daily Quiz 16** in Brightspace.

]

.small[

No AI or internet access during the quiz.

]
