---
title: "Lecture 5 Notes"
output: html_document
date: "2026-09-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Putting variables together
- always always look at the variable *TYPE*

- continuous by continuous -> scatterplot

```{r}
require(tidyverse)

df <- read_rds('https://github.com/jbisbee1/DS1000_F2026/raw/refs/heads/main/data/sc_debt.Rds')
```

# Reminding ourselves
- what types of data are these variables (`adm_rate` and `md_earn_wne_p6`)?

```{r}
df %>%
  select(adm_rate,md_earn_wne_p6)
```

# Multivariate visualization

```{r}
df %>%
  filter(preddeg == "Bachelor's") %>%
  ggplot(aes(x = adm_rate,
             y = md_earn_wne_p6)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

# Observation 1
- looks like the relationship is different, depending on whether the adm_rate > 0.5

```{r}
df %>%
  filter(preddeg == "Bachelor's",
         adm_rate <= 0.5) %>%
  ggplot(aes(x = adm_rate,
             y = md_earn_wne_p6)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  labs(x = 'Admissions rate',
       y = 'Future earnings',
       subtitle = 'Subsetting to more selective schools')

df %>%
  filter(preddeg == "Bachelor's",
         adm_rate > 0.5) %>%
  ggplot(aes(x = adm_rate,
             y = md_earn_wne_p6)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  labs(x = 'Admissions rate',
       y = 'Future earnings',
       subtitle = 'Subsetting to less selective schools')
```

# Looking at the outliers

- big earnings outliers

```{r}
df %>%
  filter(md_earn_wne_p6 > 100000) %>%
  select(instnm) %>%
  pull(instnm)
```

- small earnings outliers

```{r}
df %>%
  filter(preddeg == "Bachelor's",
         adm_rate < .025) %>%
  select(instnm,stabbr)
```

# Categorical by continuous

- reintroduce histograms / boxplots with colors

```{r}
df %>%
  ggplot(aes(x = adm_rate,
             fill = preddeg)) + 
  geom_density(alpha = .4)
```

# What if there are many categories?

- using states (`stabbr`) as an example

```{r}
df %>%
  ggplot(aes(x = adm_rate,
             fill = stabbr)) + 
  geom_density(alpha = .4)
```

# better approach uses boxplots

```{r}
df %>%
  ggplot(aes(x = adm_rate,
             y = stabbr)) + 
  geom_boxplot()
```

# NBA example

```{r}
nba <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/refs/heads/main/data/nba_players_2018.Rds")

nba %>%
  select(matches('Rookie|conf')) %>%
  count(isRookie,idConference) %>%
  spread(isRookie,n)
```





