---
title: "How Can a Model Summarize a Relationship?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

debt <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/sc_debt.Rds") %>%
  filter(adm_rate > 0, md_earn_wne_p6 > 0, sat_avg > 0)

model_adm <- lm(md_earn_wne_p6 ~ adm_rate, data = debt)
model_sat <- lm(md_earn_wne_p6 ~ sat_avg, data = debt)

# A deliberately bad guess vs. the actual line of best fit, for the RMSE demo.
guess_a <- 0; guess_b <- 0
rmse_guess <- sqrt(mean((debt$md_earn_wne_p6 - (guess_a + guess_b * debt$adm_rate))^2, na.rm = TRUE))
rmse_lm <- sqrt(mean(residuals(model_adm)^2))
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 11 opens the Regression sub-unit of Unit 3. It returns to the
college-selectivity running claim from Units 1-2 (sc_debt.Rds, adm_rate and
sat_avg predicting md_earn_wne_p6) and asks how a line can summarize a
relationship more compactly than conditional means. RMSE is introduced only
as the criterion the line of best fit minimizes; formal model evaluation
(RMSE for assessing fit, logs for skew, cross-validation) is reserved for
Lecture 12 (Regression II).

INSTRUCTOR NOTE:
Keep the p-value/confidence connection brief -- one bridge slide back to
Unit 3's bootstrap material, not a full treatment of inferential statistics.
-->

# How Can a Model Summarize a Relationship?

.big[
So far:

> **How confident should I be in a comparison, given my sample?**

Today:

> **How can I summarize an entire relationship in just a few numbers?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → interrogate explanations → quantify uncertainty → summarize with a model → communicate**
]

Today:

**Summarize with a model**

---

# Back to Our Running Claim

.huge[
> **More selective colleges produce graduates who earn more.**
]

We have described this relationship. We have not yet **summarized** it.

---

# Conditional Means, Revisited

In Unit 2, we compared groups using **conditional means**:

.big[
average earnings for **less selective** colleges

vs.

average earnings for **more selective** colleges
]

That required chopping a continuous variable into groups.

---

# The Problem With Chopping

```{r}
debt %>%
  mutate(selectivity_group = ntile(adm_rate, 4)) %>%
  group_by(selectivity_group) %>%
  summarise(mean_earn = mean(md_earn_wne_p6, na.rm = TRUE))
```

.big[
Useful, but coarse. Where exactly do the group boundaries go? What happens *within* a group?
]

---

# A Smoother Summary

.big[
Instead of a few group averages, what if we fit a single **line** through the whole relationship?
]

That line is a **model**: a compact, continuous summary of how $Y$ changes as $X$ changes.

---

# The Scatterplot We Already Know

```{r}
debt %>%
  ggplot(aes(x = adm_rate, y = md_earn_wne_p6)) +
  geom_point(alpha = .4)
```

---

# Adding a Line

```{r}
debt %>%
  ggplot(aes(x = adm_rate, y = md_earn_wne_p6)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm", se = FALSE)
```

.big[
This line is the **line of best fit**.
]

---

# What Makes a Line "Best"?

Any straight line can be written:

.huge[
$Y = \alpha + \beta X$
]

- $\alpha$ (the **intercept**): predicted $Y$ when $X = 0$
- $\beta$ (the **slope**): how much $Y$ changes for each one-unit increase in $X$

"Best" means: the line that makes the **smallest total mistakes**.

---

# Errors, or Residuals

For every observation, the line makes a prediction $\hat{Y}$. The **residual** is:

.huge[
$\text{residual} = Y - \hat{Y}$
]

The true value minus what the line predicted.

---

# Measuring Total Error: RMSE

.big[
**Root Mean Squared Error (RMSE):**

1. Square every residual (so positive and negative errors don't cancel out).
2. Average the squared residuals.
3. Take the square root, to get back to the original units.
]

The line of best fit is the one that **minimizes RMSE**.

---

# A Bad Guess vs. the Best Line

Suppose we guessed the flattest possible line: $\alpha = 0$, $\beta = 0$ (predict everyone earns \$0).

```{r}
round(rmse_guess, 0)
```

The actual best-fit line:

```{r}
round(rmse_lm, 0)
```

.big[
The best-fit line makes far smaller mistakes, on average.
]

---

# R Finds the Best Line For You

The `lm()` function ("linear model") finds $\alpha$ and $\beta$ automatically.

```{r}
model_adm <- lm(formula = md_earn_wne_p6 ~ adm_rate, data = debt)
```

.big[
`md_earn_wne_p6 ~ adm_rate` reads as **"Y explained by X."**

The `~` is R's way of writing an equation without an `=`.
]

---

# Reading the Output

```{r}
summary(model_adm)
```

---

# Interpreting the Intercept

```{r,echo=FALSE}
coef_adm <- coef(model_adm)
```

.big[
`(Intercept)` = `r round(coef_adm[1],0)`
]

Predicted earnings for a college with an admission rate of **0** (admits no one).

.small[
This is a mathematical extrapolation, not a claim about a real college — no school has `adm_rate = 0`.
]

---

# Interpreting the Slope

.big[
`adm_rate` = `r round(coef_adm[2],0)`
]

For each one-unit (i.e., 100 percentage point) increase in admission rate, predicted earnings change by about \$`r round(coef_adm[2],0)`.

.big[
The sign is negative: **higher admission rate (less selective) → lower predicted earnings.**
]

This matches our running claim's direction.

---

# A Bridge Back to Uncertainty

Look again at the last column of the summary table: `Pr(>|t|)`, the **p-value**.

.big[
Recall Unit 3: we built confidence by resampling.
]

A p-value answers a related question using a formula instead of a bootstrap loop: **how surprising would this estimate be if there were truly no relationship?**

.small[
A small p-value plays a similar role to a high bootstrap confidence. We will not derive the formula in this course — the habit of interpreting the *estimate* is what matters here.
]

---

# Do It Together: A Second Predictor

.blue[Claim]: **Average SAT scores also predict graduate earnings.**

Follow the same steps:

1. Visualize $X$ and $Y$.
2. Add the line of best fit.
3. Fit `lm()`.
4. Interpret $\alpha$ and $\beta$.

---

# Step 1–2: Visualize

```{r}
debt %>%
  ggplot(aes(x = sat_avg, y = md_earn_wne_p6)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm", se = FALSE)
```

---

# Step 3: Fit the Model

```{r}
model_sat <- lm(md_earn_wne_p6 ~ sat_avg, data = debt)
summary(model_sat)
```

---

# Step 4: Interpret

```{r,echo=FALSE}
coef_sat <- coef(model_sat)
```

.big[
Intercept: `r round(coef_sat[1],0)` — predicted earnings when average SAT is 0 (not a real college).

Slope: `r round(coef_sat[2],0)` — for each one-point increase in average SAT, predicted earnings rise by about \$`r round(coef_sat[2],0)`.
]

The sign is positive this time. Does that make sense given the claim?

---

# A Model Is Still Just a Summary

.big[
A regression line describes the **association** as compactly as possible.
]

It does not, by itself, tell us:

- whether selectivity **causes** higher earnings;
- whether a confounder (Unit 2!) explains part of the slope;
- whether this line fits equally well everywhere.

---

# Audit an AI Claim

Suppose an AI assistant sees only the `adm_rate` regression output and says:

> "The negative coefficient proves that admitting fewer students causes graduates to earn more."

.big[
**What does the slope actually describe?**

**What would need to be true for this to be a causal claim?**

**What alternative explanations from Unit 2 still apply here?**
]

A regression coefficient is not immune to confounding.

---

# A Calibrated Conclusion

Instead of:

> "Selectivity causes higher earnings."

Try:

.big[
> **In these data, a one-unit increase in admission rate is associated with about \$`r abs(round(coef_adm[2],0))` lower predicted graduate earnings — a compact summary of the association, not evidence that selectivity itself causes the difference.**
]

---

# Before You Leave

When you see a fitted regression line, ask:

1. What does the intercept represent, and is it a meaningful value of $X$?
2. What does the slope represent, in the original units of $Y$ and $X$?
3. Does the sign of the slope match the claim being tested?
4. What is this line minimizing, and what does that criterion assume?
5. What would it take to move from "summarizes the association" to "identifies the cause"?

---

# The Rule to Remember

.huge[
**A regression line is a summary, not a verdict.**
]

It compresses a relationship into two numbers. It does not, on its own, resolve *why* the relationship exists.

---

# Homework 9

You will practice:

- fitting a linear model with `lm()`;
- interpreting an intercept and a slope in context;
- checking whether a slope's sign matches a claim;
- auditing an AI claim that treats a coefficient as causal;
- writing a calibrated conclusion about a regression result.

Next:

.big[
**Regression II**
]
