---
title: "How Should Classification Performance Be Evaluated?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)
library(scales)

gms <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/game_summary.Rds")

set.seed(123)
inds <- sample(1:nrow(gms), size = round(nrow(gms) / 2), replace = FALSE)
train <- gms %>% slice(inds)
test <- gms %>% slice(-inds)

model_full <- lm(isWin ~ tov + pts + treb + pctFG, data = train)
test$pred_prob <- predict(model_full, newdata = test)
test$pred_class <- ifelse(test$pred_prob >= 0.5, TRUE, FALSE)

tp <- sum(test$pred_class == TRUE & test$isWin == TRUE)
fn_ <- sum(test$pred_class == FALSE & test$isWin == TRUE)
tn <- sum(test$pred_class == FALSE & test$isWin == FALSE)
fp <- sum(test$pred_class == TRUE & test$isWin == FALSE)

sensitivity_50 <- tp / (tp + fn_)
specificity_50 <- tn / (tn + fp)

# Threshold sweep, to build sensitivity/specificity curves and an ROC curve.
sweep_thresholds <- function(probs, truth, thresholds = seq(0, 1, by = 0.02)) {
  out <- NULL
  for (t in thresholds) {
    pred_t <- ifelse(probs >= t, TRUE, FALSE)
    tp_t <- sum(pred_t == TRUE & truth == TRUE)
    fn_t <- sum(pred_t == FALSE & truth == TRUE)
    tn_t <- sum(pred_t == FALSE & truth == FALSE)
    fp_t <- sum(pred_t == TRUE & truth == FALSE)
    out <- bind_rows(out, tibble(
      threshold = t,
      sensitivity = tp_t / (tp_t + fn_t),
      specificity = tn_t / (tn_t + fp_t)
    ))
  }
  out
}

roc_full <- sweep_thresholds(test$pred_prob, test$isWin)

# Trapezoidal AUC.
compute_auc <- function(roc_dat) {
  roc_sorted <- roc_dat %>%
    mutate(fpr = 1 - specificity) %>%
    drop_na(sensitivity, fpr) %>%
    arrange(fpr)
  sum(diff(roc_sorted$fpr) * (head(roc_sorted$sensitivity, -1) + tail(roc_sorted$sensitivity, -1)) / 2)
}

auc_full <- compute_auc(roc_full)

# Repeated cross-validated AUC, comparing the full model to a turnovers-only model.
set.seed(123)
cv_auc <- NULL
for (i in 1:100) {
  inds_i <- sample(1:nrow(gms), size = round(nrow(gms) / 2), replace = FALSE)
  train_i <- gms %>% slice(inds_i)
  test_i <- gms %>% slice(-inds_i)

  m_full_i <- lm(isWin ~ tov + pts + treb + pctFG, data = train_i)
  m_simple_i <- lm(isWin ~ tov, data = train_i)

  probs_full <- predict(m_full_i, newdata = test_i)
  probs_simple <- predict(m_simple_i, newdata = test_i)

  auc_full_i <- compute_auc(sweep_thresholds(probs_full, test_i$isWin))
  auc_simple_i <- compute_auc(sweep_thresholds(probs_simple, test_i$isWin))

  cv_auc <- bind_rows(
    cv_auc,
    tibble(sim = i, model = "Full (tov + pts + treb + pctFG)", auc = auc_full_i),
    tibble(sim = i, model = "Simple (tov only)", auc = auc_simple_i)
  )
}

cv_auc_summary <- cv_auc %>% group_by(model) %>% summarise(mean_auc = mean(auc, na.rm = TRUE))
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 15 completes Unit 4's Classification material. It develops the
sensitivity/specificity vocabulary Lecture 14 only named, builds an ROC
curve and AUC by hand (a manual threshold sweep + trapezoidal integration,
deliberately avoiding a tidymodels/yardstick dependency the course hasn't
introduced), and closes by cross-validating AUC to compare two model
specifications -- mirroring Lecture 12's cross-validated RMSE comparison,
now applied to a classification metric.

INSTRUCTOR NOTE:
Logistic regression / GLMs are intentionally NOT introduced here. The course
stays with the linear-probability-model classifier from Lecture 14; the
point of this lecture is evaluation, not a new model class.
-->

# How Should Classification Performance Be Evaluated?

.big[
Last class:

> **How do I build a classifier and check its overall accuracy?**

Today:

> **What is my classifier good and bad at, specifically — and how do I compare classifiers fairly?**
]

---

# Where We Are

.big[
**Question → ... → summarize with a model → evaluate the model → choose the right goal → build a classifier → evaluate the tradeoffs**
]

Today:

**Evaluate the tradeoffs**

---

# Picking Up the Confusion Matrix

Same classifier as last class, same 0.5 threshold:

```{r}
test %>% count(isWin, pred_class)
```

Last time we asked: what's the overall accuracy? Today: what's happening **inside** each row?

---

# Two Questions, Not One

.big[
**Sensitivity:** of the games actually **won**, what share did the model correctly flag as wins?

**Specificity:** of the games actually **lost**, what share did the model correctly flag as losses?
]

---

# Calculating Both

```{r}
sensitivity_50 <- tp / (tp + fn_)
specificity_50 <- tn / (tn + fp)

round(c(sensitivity = sensitivity_50, specificity = specificity_50), 3)
```

.big[
Are these close to each other, or far apart?
]

---

# Why Both Matter

.big[
A model that predicts "loss" for almost everything can have very high specificity and terrible sensitivity — and still post a respectable overall accuracy, if losses are common.
]

Sensitivity and specificity separate "good at catching wins" from "good at catching losses."

---

# The Threshold Controls Both

.big[
Lecture 14 showed: change the 0.5 cutoff, and the confusion matrix changes.
]

Raising the threshold makes the model more cautious about predicting "win" — which shifts sensitivity and specificity in **opposite directions**.

---

# Sweeping the Threshold

```{r, eval=FALSE}
sweep_thresholds <- function(probs, truth, thresholds = seq(0, 1, by = 0.02)) {
  out <- NULL
  for (t in thresholds) {
    pred_t <- ifelse(probs >= t, TRUE, FALSE)
    tp_t <- sum(pred_t == TRUE & truth == TRUE)
    fn_t <- sum(pred_t == FALSE & truth == TRUE)
    tn_t <- sum(pred_t == FALSE & truth == FALSE)
    fp_t <- sum(pred_t == TRUE & truth == FALSE)
    out <- bind_rows(out, tibble(threshold = t,
                                  sensitivity = tp_t / (tp_t + fn_t),
                                  specificity = tn_t / (tn_t + fp_t)))
  }
  out
}
```

---

# Sensitivity and Specificity vs. Threshold

```{r}
roc_full %>%
  pivot_longer(c(sensitivity, specificity), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = threshold, y = value, color = metric)) +
  geom_line()
```

---

# The Tradeoff, Visualized

.big[
As the threshold rises: specificity tends to rise, sensitivity tends to fall.

There is no threshold that maximizes both at once.
]

Which one matters more depends on the **cost of being wrong** in each direction — a judgment call, not a statistical one.

---

# From a Tradeoff to a Curve

.big[
Instead of picking one threshold, plot the whole tradeoff:
]

**x-axis:** 1 − specificity (how often we wrongly call a loss a win)

**y-axis:** sensitivity (how often we correctly catch a win)

This is the **ROC curve** (Receiver Operating Characteristic).

---

# Plotting the ROC Curve

```{r}
roc_full %>%
  mutate(fpr = 1 - specificity) %>%
  ggplot(aes(x = fpr, y = sensitivity)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "1 − specificity", y = "sensitivity")
```

---

# Reading the ROC Curve

.big[
The dashed diagonal line is what a **coin flip** would produce — no ability to distinguish wins from losses at any threshold.

A curve that bows **up and to the left**, away from the diagonal, means the classifier separates the two classes well across thresholds.
]

---

# One Number for the Whole Curve

.big[
**AUC (Area Under the Curve):** the entire ROC curve's performance, summarized as one number.
]

- AUC of 0.5 → no better than guessing.
- AUC of 1.0 → perfect separation at every threshold.
- Between 0.5 and 1.0 → some real ability to distinguish the classes, evaluated across **every** threshold at once.

---

# Calculating AUC

```{r, eval=FALSE}
compute_auc <- function(roc_dat) {
  roc_sorted <- roc_dat %>%
    mutate(fpr = 1 - specificity) %>%
    arrange(fpr)
  sum(diff(roc_sorted$fpr) * (head(roc_sorted$sensitivity, -1) +
                                 tail(roc_sorted$sensitivity, -1)) / 2)
}
```

```{r}
round(auc_full, 3)
```

---

# Why AUC Is Useful for Comparison

.big[
Accuracy depends on a threshold you have to pick.

AUC does not — it summarizes performance **across every possible threshold**, which makes it a fairer way to compare two different classifiers.
]

---

# Comparing Models, the Lecture 12 Way

Recall: to compare regression models, we cross-validated RMSE across 100 train/test splits.

.big[
Same idea, new metric: cross-validate **AUC** to compare two classifiers.
]

- Full model: `isWin ~ tov + pts + treb + pctFG`
- Simple model: `isWin ~ tov`

---

# Repeated Cross-Validated AUC

```{r, eval=FALSE}
for (i in 1:100) {
  inds_i <- sample(1:nrow(gms), size = round(nrow(gms) / 2), replace = FALSE)
  train_i <- gms %>% slice(inds_i)
  test_i <- gms %>% slice(-inds_i)

  m_full_i <- lm(isWin ~ tov + pts + treb + pctFG, data = train_i)
  m_simple_i <- lm(isWin ~ tov, data = train_i)
  # ...compute AUC for each on test_i, store results...
}
```

---

# The Results

```{r}
cv_auc_summary
```

```{r}
cv_auc %>%
  ggplot(aes(x = auc, fill = model)) +
  geom_density(alpha = .4)
```

---

# What This Comparison Tells Us

.big[
The model with the higher average cross-validated AUC separates wins from losses more reliably, across thresholds, on data it wasn't fit to.
]

It does **not** tell us which predictor *causes* winning — same caution as every model comparison this semester.

---

# Audit an AI Claim

Suppose an AI assistant is shown only the model's accuracy at the default 0.5 threshold and says:

> "This model performs at 74% and that fully describes how good it is."

.big[
**What does a single-threshold accuracy number leave out?**

**Why might AUC give a more complete picture?**

**What judgment call does picking a threshold still require, even with a good AUC?**
]

---

# A Calibrated Conclusion

Instead of:

> "The model is 74% accurate."

Try:

.big[
> **Across repeated train/test splits, the full model's average AUC was `r round(cv_auc_summary$mean_auc[cv_auc_summary$model == "Full (tov + pts + treb + pctFG)"],3)`, compared to `r round(cv_auc_summary$mean_auc[cv_auc_summary$model == "Simple (tov only)"],3)` for the turnovers-only model — a more complete comparison than a single accuracy number, though neither establishes what causes winning, and any real use would still require choosing a threshold based on the cost of each kind of mistake.**
]

---

# Before You Leave

When you see a classifier's performance reported, ask:

1. Was accuracy reported at one threshold, or across the tradeoff?
2. What are sensitivity and specificity, separately?
3. Does the ROC curve bow away from the diagonal?
4. Was AUC calculated with held-out data, ideally cross-validated?
5. What threshold would actually get used, and what does that choice cost in each direction?

---

# The Rule to Remember

.huge[
**One accuracy number is a snapshot.**

**A full evaluation looks across every threshold.**
]

---

# Homework 12

You will practice:

- calculating sensitivity and specificity from a confusion matrix;
- sweeping a threshold to see the tradeoff between them;
- building and reading an ROC curve;
- comparing two classifiers using cross-validated AUC;
- auditing an AI claim that treats one threshold's accuracy as the full picture.

Next:

.big[
**Clustering / Unsupervised Learning**
]
