---
title: "Prepare the Evidence"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
library(tidyverse)
df <- read_rds("https://github.com/jbisbee1/DS1000_F2024/raw/main/data/sc_debt.Rds")
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES
LO1: Interrogate empirical claims.
LO2: Translate questions into data problems.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 4 primarily reinforces LO1, LO2, LO4, and LO5.
-->

# Prepare the Evidence

.big[
We have a question.

We have variables.

We have inspected their distributions.

Now:

> **What decisions are required before these data become trustworthy evidence?**
]

---

# Where We Are

.big[
**Question → Concepts → Variables → Unit of analysis → Understand each variable → Prepare the evidence → Put variables together → Evaluate → Communicate**
]

Today:

**Prepare the evidence**

<!-- LO2: Adds data preparation explicitly to the decomposition workflow. -->

---

# Our Running Question

.big[
> **More selective colleges produce graduates who earn more.**
]

We identified:

- `adm_rate`
- `md_earn_wne_p6`

But the raw College Scorecard contains many institutions and many variables.

We need to decide what actually belongs in our analysis.

---

# Wrangling Is Judgment

.big[
**Data wrangling is the documented process of turning raw observations into the evidence required by a particular question.**
]

R executes the decisions.

**We make the decisions.**

<!-- LO4: Reframes wrangling as analytical judgment rather than syntax. -->

---

# Four Decisions

.big[
Before analysis, ask:

1. **Who belongs?**
2. **What is missing?**
3. **What needs to change?**
4. **Can someone reproduce what I did?**
]

These questions matter more than remembering function names.

---

# Who Belongs?

Our claim sounds like it concerns "colleges."

But the Scorecard contains different kinds of institutions.

.big[
**What population do we actually want to make claims about?**
]

Suppose we decide:

> Bachelor's-degree institutions.

That is a substantive decision.

<!-- LO1: Population definition determines scope of inference. -->

---

# Filtering Changes the Claim

If we remove institutions from the data, we are changing the population represented by the analysis.

.big[
`filter()` is not just a coding operation.

It is an answer to:

> **Which observations belong in my analysis?**
]

---

# Tell R the Decision

Suppose `preddeg == 3` identifies institutions where the predominant degree is a bachelor's degree.

```{r}
college <- df %>%
  filter(preddeg == 3)
```

The important part is not the syntax.

The important part is knowing **why** we made this restriction.

<!-- LO4: Code implements a substantive population decision. -->

---

# Verify the Decision

Never assume the code did what you intended.

```{r}
nrow(df)
nrow(college)

table(college$preddeg, useNA = "ifany")
```

.big[
**Specification → execution → verification**
]

That workflow will recur throughout the course.

<!-- LO5: Same verification workflow applies to AI-generated code. -->

---

# What Is Missing?

Lecture 3 introduced `NA`.

Now ask a harder question:

.big[
> **Which observations are missing the evidence we need?**
]

For our question, we need both:

- admissions rate
- graduate earnings

---

# Count Missingness

```{r}
college %>%
  summarise(
    missing_adm = sum(is.na(adm_rate)),
    missing_earn = sum(is.na(md_earn_wne_p6))
  )
```

You do not need to memorize this syntax.

You do need to understand what is being counted.

---

# Dropping Is a Decision

We could write:

```{r}
analysis_df <- college %>%
  drop_na(adm_rate, md_earn_wne_p6)
```

But before doing so:

.big[
> **Who disappears?**
]

And:

> **Could the missing institutions be systematically different?**

<!-- LO1: Missing-data handling affects evidentiary claims. -->

---

# Compare Before and After

```{r}
nrow(college)
nrow(analysis_df)
```

How many observations did we lose?

That number is part of the analysis.

.big[
Deleting missing data should never be an invisible step.
]

---

# Missingness Can Be Evidence

Why might admissions rate be missing?

Possibilities include:

- open-admission institutions
- non-reporting
- measure not applicable
- data collection problems

.big[
Missingness is sometimes telling us something about how the data were generated.
]

---

# What Needs to Change?

Sometimes the stored variable is correct but inconvenient.

`adm_rate` is stored as a proportion:

`0.25`

Humans often communicate admissions rates as percentages:

`25%`

Same information.

Different representation.

---

# Transform for Meaning

```{r}
analysis_df <- analysis_df %>%
  mutate(adm_pct = adm_rate * 100)
```

.big[
A transformation should have a reason.

Here the reason is **interpretability**.
]

<!-- LO4: Transformation motivated by substantive communication. -->

---

# Transformations Can Mislead

Imagine someone accidentally writes:

```r
mutate(adm_pct = adm_rate / 100)
```

The code runs.

The result is wrong.

.big[
**Computational validity is not analytical validity.**
]

<!-- LO5: Students must audit transformations even if AI/code executes successfully. -->

---

# Keep What We Need

Our question does not require every Scorecard variable.

Suppose we retain:

- institution name
- state
- admissions percentage
- graduate earnings

```{r}
analysis_df <- analysis_df %>%
  select(instnm, stabbr, adm_pct, md_earn_wne_p6)
```

---

# Why Select Variables?

Not because extra columns are inherently bad.

.big[
A focused analytical dataset makes our reasoning easier to inspect.
]

We should be able to explain why every retained variable matters.

---

# The Pipeline

.big[
**Raw data**

↓

Define the population

↓

Choose relevant variables

↓

Investigate missingness

↓

Make justified exclusions

↓

Transform when needed

↓

Verify

↓

**Analytical data**
]

---

# Every Arrow Is a Choice

.big[
There is no neutral "cleaning" step.
]

Choices about:

- inclusion
- exclusion
- missingness
- coding
- transformation

can change the evidence we eventually see.

<!-- LO1: Data preparation itself can shape factual claims. -->

---

# Reproducibility

Imagine two workflows.

**Workflow A**

Open spreadsheet → delete rows → edit cells → save `FINAL_v2_REAL.csv`

**Workflow B**

Raw data → documented code → analytical data

.big[
Which one can another analyst inspect and reproduce?
]

---

# Why We Use Code

The advantage of code is not merely speed.

.big[
Code creates a record of the decisions that transformed the data.
]

Someone else can:

- inspect them
- question them
- rerun them
- change them

That is part of trustworthy analysis.

<!-- LO4 + LO6: Reproducibility supports evaluation and communication. -->

---

# AI Can Write the Pipeline

Suppose we tell an AI assistant:

> Keep bachelor's-degree institutions with observed admissions rates and earnings. Convert admissions rates to percentages. Retain institution name, state, admissions percentage, and earnings.

AI can translate that specification into R.

That is useful.

---

# But AI Does Not Own the Decisions

The AI should not decide for us:

- what population the claim concerns
- which variable represents the concept
- whether dropping missing observations is justified
- whether a transformation preserves meaning

.big[
**AI can implement a specification. We remain responsible for the specification.**
]

<!-- LO5: Explicit division of labor between AI and student. -->

---

# Audit AI-Generated Code

When AI gives you a wrangling pipeline, ask:

1. Did it keep the intended population?
2. Did it retain the intended variables?
3. Did it handle missingness as instructed?
4. Are transformations mathematically correct?
5. Did it make an unstated decision?

.big[
Do not ask only: **Does it run?**
]

Ask: **Does it do what we intended?**

---

# A Plausible Mistake

An AI assistant proposes:

```r
analysis_df <- df %>%
  filter(preddeg == 3,
         adm_rate > 0,
         md_earn_wne_p6 > 0) %>%
  mutate(adm_pct = adm_rate / 100) %>%
  select(instnm, adm_pct, md_earn_wne_p6)
```

.big[
The code runs.

Is it correct?
]

---

# Audit the Pipeline

What should we inspect?

- `adm_rate > 0` is not the same instruction as "observed admissions rate"
- `md_earn_wne_p6 > 0` introduces an additional restriction
- percentage conversion is backwards
- state was omitted

.big[
Plausible code can silently change the analysis.
]

<!-- LO5: Concrete critique of AI-style plausible code. -->

---

# Explain the Analysis

A competent analyst should be able to say:

.big[
> "I restricted the data to bachelor's-degree institutions, required observed admissions-rate and earnings data, converted admissions rates to percentages for interpretability, and retained the variables needed for the analysis."
]

That explanation is more important than reciting the R syntax.

<!-- LO6: Communicate data preparation decisions clearly. -->

---

# A New Claim

Consider:

.big[
> **Students at more expensive colleges graduate with more debt.**
]

Before analyzing it, specify:

- population
- concepts
- variables
- unit of analysis
- missing-data requirements
- transformations, if any

Then write the computational instructions.

<!-- LO2: Transfer to unfamiliar substantive claim. -->

---

# From Question to Evidence

Our first unit now looks like this:

.big[
**What is the claim?**

↓

**What concepts does it involve?**

↓

**How are they measured?**

↓

**What does one row represent?**

↓

**What does each variable look like?**

↓

**What preparation decisions are required?**
]

---

# What Quiz 1 Will Ask

You will not need to reproduce long tidyverse pipelines from memory.

You should be able to:

- identify the unit of analysis
- distinguish concepts from variables
- evaluate an operationalization
- choose an appropriate univariate visualization
- interpret a distribution
- reason about missingness
- determine whether simple R code implements an analytical instruction

<!-- Instructor note: This slide makes the assessment target transparent without giving exact questions. -->

---

# The Rule to Remember

.huge[
**Raw data are not evidence yet.**
]

.big[
They become evidence through a series of choices.

Those choices should be:

**justified, visible, and reproducible.**
]

---

# Next

Homework 4 will ask you to rehearse the full workflow on a new problem.

Then:

.big[
**Lab Quiz 1**
]

You will perform the reasoning independently.
