---
title: "Lecture 3: Know Your Variables"
subtitle: "Looking at a single variable"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis-fonts", "css/ds1000.css"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
.hero-image {
  display: flex;
  align-items: center;
  justify-content: center;
  height: 74vh;
}
.hero-image img {
  max-width: 92%;
  max-height: 74vh;
  object-fit: contain;
}

</style>

```{css,echo=FALSE}
.small .remark-code { font-size: 85% !important; }
.tiny .remark-code { font-size: 60% !important; }
.big { font-size: 150%; }
.huge { font-size: 200%; }
```

```{r setup, include=FALSE}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
library(tidyverse)
df <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/sc_debt.Rds")
```

<!--
LEARNING OBJECTIVES
LO1: Interrogate empirical claims.
LO2: Translate questions into data problems.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 3 primarily introduces LO1, LO3, and LO4, while reinforcing LO2 and LO5.
-->


# Know Your Variables

**Before asking how variables relate, understand each variable on its own.**

Today:

- What does a variable actually measure?
- What kind of variable is it?
- What does its distribution look like?
- Does the distribution make sense?
- What might the data be hiding?

<!--
LO2: This is the next step after Lecture 2's question -> concepts -> variables -> unit of analysis workflow.
-->


---

# Where We Are

**Question → Concepts → Variables → Unit of analysis → Understand each variable → Put variables together → Evaluate → Communicate**

We are here:

**Understand each variable**

<!--
LO2: Reinforces decomposition.
LO4: Sets up univariate analysis as a core analytical step.
-->


---

# Our Running Question

> **More selective colleges produce graduates who earn more.**

In Lecture 2, we identified possible variables:

- `adm_rate`
- `md_earn_wne_p6`

Now we have to decide whether these variables are good evidence.

<!--
LO1: Returns to empirical claim.
-->


---

# Concept or Variable?

**Concept:** selectivity

**Variable:** admissions rate

A concept is an idea we care about.

A variable is how we represent that idea in data.

<!--
LO2: Distinguish substantive concept from observed variable.
-->


---

# What Does `adm_rate` Measure?

`adm_rate` is the proportion of applications that a college accepts.

That is not identical to:

- "How selective is this college?"
- "How difficult is it to get into this college?"
- "How good is this college?"

Those are related concepts, but they are not the same measurement.

<!--
LO1: Encourages skepticism about operationalizations.
LO2: Connects concepts to variables.
-->


---

# Construct Validity

**Construct validity:** How well does a variable capture the concept we actually care about?

For admissions rate:

**Concept:** selectivity

**Measurement:** proportion of applicants admitted

Potential mismatch:

A college's admissions rate depends partly on **who applies**.

<!--
LO1: First explicit introduction of construct validity as a skeptical habit.
-->


---

# What About Earnings?

**Concept:** economic outcomes after college

**Variable:** `md_earn_wne_p6`

Ask:

> What exactly does this variable measure?

It is not simply "how much a graduate earns."

Before interpreting a number, understand what generated it.

<!--
LO1: Measurement skepticism.
LO4: Interpretation begins with understanding the variable.
-->


---

# Variable Types

For today's purposes, we need two broad types.

**Continuous:** values represent quantities.

Examples:

`adm_rate`  
`md_earn_wne_p6`

**Categorical:** values identify groups.

Examples:

`stabbr`  
`preddeg`

<!--
LO3: Variable type determines appropriate visualization.
-->


---

# Variable Type Matters

Different variables require different ways of looking at them.

**One continuous variable → histogram**

**One categorical variable → bar chart**

Today we will focus only on these two.

Scatterplots and boxplots come later, when we put variables together.

<!--
LO3: Explicit method-selection rule.
-->


---

# Start With a Distribution

Suppose we want to understand `adm_rate`.

Before asking whether selectivity is related to earnings, ask:

> **What does admissions rate look like across colleges?**

A histogram gives us a first answer.

<!--
LO4: Univariate analysis as first inspection.
-->


---

# A Histogram

```{r, echo=TRUE, fig.height=4.5, fig.width=8}
library(tidyverse)

df %>%
  ggplot(aes(x = adm_rate)) +
  geom_histogram()
```

<!--
LO4: Basic visualization.
-->


---

# What Did We Learn?

The histogram can tell us about:

- typical values
- variation
- shape
- extreme observations
- possible problems

It cannot, by itself, tell us whether selective colleges have higher earnings.

<!--
LO4: Distinguish description from relationship.
LO6: Sets up limits of an individual visualization.
-->


---

# Read the Distribution

When you see a continuous variable, ask:

**Center:** What values are typical?

**Spread:** How much do values vary?

**Shape:** Symmetric? Skewed? Multiple peaks?

**Extremes:** Are there unusual observations?

<!--
LO4: Concrete interpretation framework.
-->


---

# Read the Distribution

Also ask:

**Bounds:** Are any values impossible?

**Missingness:** Which observations have no value?

**Meaning:** Does the distribution make sense given what this variable measures?

A graph is not just a picture.

It is an **audit of our evidence**.

<!--
LO1: Data skepticism.
LO4: Visualization as analysis and audit.
-->


---

# Data That Make No Sense

Suppose admissions rate contains:

- `-0.2`
- `1.4`

Those values might be perfectly valid R numbers.

They are not valid admissions rates.

**Code can run even when the data are wrong.**

<!--
LO4: Computational correctness is not substantive correctness.
LO5: Same distinction applied to AI-generated code/data handling.
-->


---

# Missing Is Not Zero

An `NA` does not mean:

- zero
- none
- below average

It means we do not have an observed value.

But the important question is:

> **Why is it missing?**

Who is missing?

Why?

Could missingness change what we conclude?

<!--
LO1: Missingness as evidence problem.
LO4: Data quality and interpretation.
-->


---

# Categorical Variables

Now consider `preddeg`.

It tells us the predominant degree offered by an institution.

For a categorical variable, ask:

- What categories exist?
- How common is each?
- Are categories missing?
- Do the categories capture the concept we care about?

<!--
LO3: Categorical-variable reasoning.
-->


---

# A Bar Chart

```{r, echo=TRUE, fig.height=4.5, fig.width=8}
df %>%
  ggplot(aes(x = preddeg)) +
  geom_bar()
```

<!--
LO4: Appropriate univariate visualization.
-->


---

# Counts Can Mislead Too

A bar chart shows how common categories are in the dataset.

But ask:

> What population does this dataset contain?

If the data contain mostly one type of institution, the graph describes the **dataset**.

It does not automatically describe **all colleges**.

<!--
LO1: Generalization and evidence.
LO4: Interpretation of descriptive data.
-->


---

# AI Can Make the Plot

Suppose we ask an AI assistant:

> "I have a numeric variable called `adm_rate` containing college admissions rates. Write ggplot2 code to visualize its distribution and explain what the plot shows."

AI can produce the code quickly.

But that is not the end of the task.

<!--
LO5: Introduces AI as implementation tool.
-->


---

# Our Job Is to Evaluate It

Check:

1. Did it choose an appropriate plot?
2. Is the variable mapped correctly?
3. Are the labels and scales sensible?
4. Does the interpretation match the graph?
5. Did it ignore anything important about the data?

**A plausible visualization is not necessarily a good visualization.**

<!--
LO5: Direct evaluative judgment.
LO3: Method selection.
LO4: Interpretation.
-->


---

# The Human Skill

AI can generate:

- code
- plots
- summaries
- interpretations

You still need to decide:

> **Is this actually an appropriate analysis of the question and the data?**

That judgment is part of being a data scientist.

<!--
LO5: Core AI literacy outcome.
-->


---

# Now Inspect Earnings

Repeat the process for:

`md_earn_wne_p6`

Before plotting, predict:

- What type of variable is it?
- What would a sensible distribution look like?
- What would surprise you?
- What would make you question the measurement?

<!--
LO3 + LO4: Transfer the workflow to a second variable.
-->


---

# An Earnings Distribution

```{r, echo=TRUE, fig.height=4.5, fig.width=8}
df %>%
  ggplot(aes(x = md_earn_wne_p6)) +
  geom_histogram()
```

<!--
LO4: Interpretation of a second continuous variable.
-->


---

# What Should We Notice?

Do not just describe the shape.

Ask:

- What does a typical value represent?
- Why might the distribution be skewed?
- Who is included?
- Who is missing?
- What does an extreme observation represent?
- Is this a sensible measure of "economic outcomes"?

<!--
LO1: Measurement and sample skepticism.
LO4: Distributional interpretation.
-->


---

# We Have Learned Two Variables

We now understand:

**Admissions rate**

and

**Graduate earnings**

better than we did before.

But we still have not answered our original question.

<!--
LO6: Separates description from substantive conclusion.
-->


---

# The Temptation

> **So, do more selective colleges have higher-earning graduates?**

We are tempted to put the two variables together immediately.

But first we needed to understand what each variable was telling us.

<!--
LO2: Complete the decomposition step.
LO3: Prepares transition to multivariate methods.
-->


---

# Next Step

**Univariate analysis**

> What is each variable like?

**Multivariate analysis**

> How do variables relate?

Next we move from:

**understanding the evidence**

to

**using the evidence to answer the question.**

<!--
LO3: Transition to multivariate method selection.
LO4: Analytical progression.
-->


---

# Before You Leave

For any important variable, you should be able to answer:

1. What concept is this variable supposed to measure?
2. What does it actually measure?
3. What is its type?
4. What does its distribution look like?
5. What unusual or missing observations should I investigate?
6. Does the distribution make sense?

<!--
LO1 + LO4: Reusable skepticism checklist.
-->


---

# The Rule to Remember

> **Before asking how variables relate, ask what each variable is.**

And:

> **Before asking what a variable does, ask what it measures.**

<!--
LO1: Core course habit.
LO4: Core analytical habit.
-->


---

# Homework 3

Use the same workflow on `adm_rate` and `md_earn_wne_p6`.

You will:

- identify the concept and operationalization;
- choose an appropriate univariate visualization;
- use AI to help generate R code;
- verify that the visualization is appropriate;
- interpret the distributions;
- investigate missingness or unusual values.

Then we will be ready to put the variables together.

<!--
LO5: AI-assisted analysis with evaluation.
LO6: Communication of descriptive evidence.
-->
