---
title: "What Does Confidence Actually Justify?"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)

library(tidyverse)

gms <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/game_summary.Rds")

# Pooled bootstrap: winning vs. losing teams' turnovers, all seasons together.
set.seed(123)
bs_pooled <- NULL
for (i in 1:200) {
  tmp <- gms %>%
    sample_n(size = nrow(gms), replace = TRUE) %>%
    group_by(isWin) %>%
    summarise(mean_tov = mean(tov, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_pooled <- bind_rows(bs_pooled, tmp)
}

pooled_diff <- bs_pooled %>%
  pivot_wider(names_from = isWin, values_from = mean_tov, names_prefix = "win_") %>%
  mutate(diff = win_FALSE - win_TRUE)

conf_pooled <- mean(pooled_diff$diff > 0)

# Season-by-season bootstrap.
set.seed(123)
bs_season <- NULL
for (i in 1:200) {
  tmp <- gms %>%
    group_by(yearSeason) %>%
    sample_n(size = n(), replace = TRUE) %>%
    group_by(yearSeason, isWin) %>%
    summarise(mean_tov = mean(tov, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_season <- bind_rows(bs_season, tmp)
}

season_diff <- bs_season %>%
  pivot_wider(names_from = isWin, values_from = mean_tov, names_prefix = "win_") %>%
  mutate(diff = win_FALSE - win_TRUE)

conf_by_season <- season_diff %>%
  group_by(yearSeason) %>%
  summarise(confidence = mean(diff > 0, na.rm = TRUE), .groups = "drop")
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 10 completes Unit 3's core instructional content. It assumes students
already know how to build a bootstrap loop and a confidence percentage
(Lecture 9). This lecture's job is not to reteach the mechanics but to press
on what that percentage is entitled to claim: it describes THIS sample's
resampling behavior, not the population, and not a new sample.

INSTRUCTOR NOTE:
The running example moves from player-level data (Lecture 9) to game-level
data (game_summary.Rds), reused from the archived Uncertainty Part 2 deck.
Keep circling back to the phrase "what does this number actually justify?"
-->

# What Does Confidence Actually Justify?

.big[
Last class:

> **How do I turn one sample into a confidence percentage?**

Today:

> **What is that percentage actually entitled to tell me?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → interrogate explanations → quantify uncertainty → communicate**
]

Today, still:

**Quantify uncertainty** — but pushed further.

---

# A New Comparison

.huge[
> **Do winning teams commit fewer turnovers than losing teams?**
]

New data: every NBA game played between 2016 and 2019.

---

# The Data

```{r}
gms %>%
  select(idGame, yearSeason, nameTeam, isWin, tov) %>%
  glimpse()
```

Each row is one **team's performance in one game**.

---

# The Observed Comparison

```{r}
gms %>%
  group_by(isWin) %>%
  summarise(mean_tov = mean(tov, na.rm = TRUE))
```

.big[
Winning teams average about one fewer turnover per game.
]

---

# You Already Know the Next Step

From Lecture 9:

.big[
1. Resample the data, with replacement.
2. Recompute the comparison.
3. Repeat many times.
4. Report the share of resamples where the pattern holds.
]

---

# The Pooled Bootstrap

```{r, eval=FALSE}
bs_pooled <- NULL
for (i in 1:200) {
  tmp <- gms %>%
    sample_n(size = nrow(gms), replace = TRUE) %>%
    group_by(isWin) %>%
    summarise(mean_tov = mean(tov, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_pooled <- bind_rows(bs_pooled, tmp)
}
```

```{r}
pooled_diff %>%
  summarise(confidence = mean(diff > 0))
```

---

# A Comfortable Number

.huge[
`r round(conf_pooled*100,1)`% confidence
]

Winning teams had fewer turnovers in `r round(conf_pooled*100,1)`% of 200 resamples of the pooled 2016–2019 data.

That sounds like a lot. What are we actually entitled to conclude from it?

---

# Two Different Questions

.big[
**Question A:** If I resampled *this same pooled dataset* again, would I likely see the same pattern?

**Question B:** Would I see the same pattern in *a season this dataset doesn't contain*?
]

Bootstrap resampling directly answers only Question A.

<!-- LO1: This is the central distinction the lecture is built around. -->

---

# Why the Difference Matters

.big[
Resampling the pooled data mixes games from 2016–17, 2017–18, and 2018–19 together.
]

A high pooled confidence could reflect:

- a pattern that holds steadily in **every** season, **or**
- a pattern driven mostly by **one big season**, diluted across the rest.

The pooled number alone cannot tell these apart.

---

# Check Season by Season

```{r}
gms %>%
  group_by(yearSeason, isWin) %>%
  summarise(mean_tov = mean(tov, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = isWin, values_from = mean_tov, names_prefix = "win_")
```

The raw gap is not identical across seasons.

---

# Bootstrap Within Each Season

```{r, eval=FALSE}
bs_season <- NULL
for (i in 1:200) {
  tmp <- gms %>%
    group_by(yearSeason) %>%
    sample_n(size = n(), replace = TRUE) %>% #<<
    group_by(yearSeason, isWin) %>%
    summarise(mean_tov = mean(tov, na.rm = TRUE), .groups = "drop") %>%
    mutate(sim = i)
  bs_season <- bind_rows(bs_season, tmp)
}
```

.big[
Resampling **within** each season keeps the seasons from blending together.
]

---

# Confidence, By Season

```{r}
conf_by_season
```

.big[
Compare these to the pooled `r round(conf_pooled*100,1)`%.
]

---

# What This Shows

.big[
The pooled confidence number can look strong even when the pattern is uneven across seasons.
]

A single confidence percentage can hide exactly the kind of heterogeneity Unit 2 taught you to look for.

<!-- Bridge back to "condition on Z" reasoning from Lecture 7. -->

---

# So What Does 78% Actually Justify?

Suppose the pooled confidence is 78%. It justifies:

.big[
**"In 78% of resamples of these 2016–2019 games, winning teams had fewer turnovers."**
]

It does **not** justify:

.big[
**"There is a 78% chance winning causes fewer turnovers."**

**"78% of future games will follow this pattern."**

**"This pattern will look the same in a season we haven't observed."**
]

---

# A Useful Habit

.big[
Before trusting a single confidence number, ask:

**Resampled from what?**
]

Confidence describes resampling of a specific dataset. If that dataset pools together things that differ (seasons, groups, populations), the number can average over real differences.

---

# Sample Size Matters Too

```{r}
gms %>%
  count(yearSeason)
```

.big[
Some seasons contribute far more games than others to the pooled bootstrap.
]

A season with fewer games swings around more under resampling — and contributes less to the pooled average.

---

# Audit an AI Claim

Suppose an AI assistant is given only the pooled 78% figure and says:

> "This confirms that reducing turnovers will improve any team's chances of winning in every season, including next year's."

.big[
**What does the 78% actually describe?**

**What did the claim assume that the number doesn't establish?**

**What would you want to see before trusting the claim about next season?**
]

---

# What a Confidence Number Is Good For

.big[
It is a disciplined way to say **how fragile a specific comparison, in a specific sample, is to resampling noise.**
]

That is genuinely useful. It is also strictly limited.

---

# What It Cannot Do On Its Own

.big[
It cannot tell you whether the comparison reflects a real underlying process.

It cannot tell you whether the pattern would replicate in a new sample.

It cannot substitute for checking subgroups, as we just did with seasons.
]

---

# A Calibrated Conclusion

Instead of:

> "We are 78% confident winning causes fewer turnovers."

Try:

.big[
> **In these 2016–2019 games, winning teams had fewer turnovers in `r round(conf_pooled*100,1)`% of resamples of the pooled data, but the pattern's strength differs by season, and a single confidence number does not establish that it would hold in a season we have not observed.**
]

---

# Before You Leave

When someone hands you a confidence percentage, ask:

1. Resampled from what data, exactly?
2. Does that data pool together groups that might differ?
3. Does the pattern hold up when I check within those groups separately?
4. Am I treating this as a description of this sample, or smuggling in a claim about the population?
5. What would I need to check before trusting this beyond the sample I have?

---

# The Rule to Remember

.huge[
**A confidence percentage describes this sample's resampling behavior.**

**It is not a verdict about the world.**
]

---

# The Unit 3 Workflow

.big[
**See a comparison**

↓

**Resample it**

↓

**Turn resamples into a confidence number**

↓

**Ask what that number does and does not justify**

↓

**Check whether it holds across relevant subgroups**
]

---

# Homework 8

You will practice:

- bootstrapping within subgroups instead of a pooled sample;
- comparing confidence numbers across those subgroups;
- distinguishing what a confidence percentage justifies from what it doesn't;
- auditing an AI claim that treats confidence as a population-level guarantee;
- writing a calibrated conclusion about a confidence result.

Next:

.big[
**Regression I**
]
