---
title: "Describe the Relationship"
subtitle: "DS 1000"
author: "Prof. Bisbee"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
library(tidyverse)
df <- read_rds("https://github.com/jbisbee1/DS1000_F2024/raw/main/data/sc_debt.Rds")

# Small synthetic datasets used only to illustrate relationship shapes.
n <- 120
x_demo <- seq(-3, 3, length.out = n)

demo_positive <- tibble(
  x = x_demo,
  y = 1.2 * x_demo + rnorm(n, 0, 1.1)
)

demo_negative <- tibble(
  x = x_demo,
  y = -1.2 * x_demo + rnorm(n, 0, 1.1)
)

demo_curved <- tibble(
  x = x_demo,
  y = x_demo^2 + rnorm(n, 0, 0.7)
)

demo_outlier <- demo_positive
demo_outlier$y[n] <- demo_outlier$y[n] - 10

demo_weak <- tibble(
  x = x_demo,
  y = 0.4 * x_demo + rnorm(n, 0, 2.5)
)

demo_strong <- tibble(
  x = x_demo,
  y = 1.2 * x_demo + rnorm(n, 0, 0.45)
)
```

<style>
.big { font-size: 130%; }
.huge { font-size: 180%; }
.small { font-size: 80%; }
</style>

<!--
LEARNING OBJECTIVES

LO1: Interrogate empirical claims.
LO3: Select and justify analytical approaches.
LO4: Conduct and evaluate data analyses.
LO5: Critically use AI as an analytical tool.
LO6: Communicate evidence responsibly.

Lecture 6 advances Test 2 competencies:
- characterize an observed relationship;
- interpret correlation;
- recognize limitations of correlation;
- describe categorical-continuous group differences.

INSTRUCTOR NOTE:
Do not introduce formal sampling uncertainty here.
Do not move into confounding except as a teaser at the end.
Lecture 7 owns alternative explanations and subgroup structure.
-->

# Describe the Relationship

.big[
Last class:

> **How should I put X and Y together?**

Today:

> **Once I can see the relationship, what should I say about it?**
]

---

# Where We Are

.big[
**Question → trustworthy variables → appropriate comparison → describe the relationship → explain/evaluate → communicate**
]

Today:

**Describe the relationship**

---

# Back to Our Claim

.big[
> **More selective colleges produce graduates who earn more.**
]

We decided that admissions rate and earnings are both continuous.

So we started with a scatterplot.

---

# See the Relationship

```{r}
df %>%
  ggplot(aes(x = adm_rate, y = md_earn_wne_p6)) +
  geom_point(alpha = .5)
```

.big[
Now what?
]

<!-- LO4: Interpretation follows appropriate representation. -->

---

# Four Questions

For a relationship between two continuous variables, ask:

.big[
**Direction:** Which way does it go?

**Form:** What shape does it take?

**Strength:** How closely do points follow the pattern?

**Unusual observations:** What does not fit?
]

This is our descriptive checklist.

---

# Direction

A relationship can be:

.big[
**Positive:** larger X tends to accompany larger Y.

**Negative:** larger X tends to accompany smaller Y.

**No clear direction:** neither pattern is apparent.
]

---

# Positive Direction

```{r, echo=FALSE}
demo_positive %>%
  ggplot(aes(x, y)) +
  geom_point()
```

.big[
As X increases, Y tends to increase.
]

---

# Negative Direction

```{r, echo=FALSE}
demo_negative %>%
  ggplot(aes(x, y)) +
  geom_point()
```

.big[
As X increases, Y tends to decrease.
]

---

# Statistical vs. Substantive Direction

For admissions rate:

> Higher `adm_rate` = **less selective**

Suppose admissions rate and earnings have a negative relationship.

Statistically:

> Higher admissions rates accompany lower earnings.

Substantively:

.big[
> **More selective colleges tend to have higher earnings.**
]

<!-- LO6: Always translate statistical direction back to substantive meaning. -->

---

# Form

Not every relationship is a straight line.

Ask:

.big[
> **What shape does the relationship take?**
]

Common possibilities:

- approximately linear
- curved
- clustered
- no obvious form

---

# A Curved Relationship

```{r, echo=FALSE}
demo_curved %>%
  ggplot(aes(x, y)) +
  geom_point()
```

.big[
There is a clear relationship.

But it is not linear.
]

---

# Why Form Matters

Suppose someone says:

> "There is no linear relationship."

That does **not** necessarily mean:

> "There is no relationship."

.big[
The shape of the data matters.
]

---

# Strength

Compare two relationships.

.big[
Which one follows its overall pattern more closely?
]

---

# A Weaker Pattern

```{r, echo=FALSE}
demo_weak %>%
  ggplot(aes(x, y)) +
  geom_point()
```

---

# A Stronger Pattern

```{r, echo=FALSE}
demo_strong %>%
  ggplot(aes(x, y)) +
  geom_point()
```

.big[
Strength is about how tightly observations follow the pattern.
]

---

# Unusual Observations

```{r, echo=FALSE}
demo_outlier %>%
  ggplot(aes(x, y)) +
  geom_point()
```

.big[
One observation can matter a lot.

Do not automatically delete it.
]

Ask what it represents.

<!-- LO1: Unusual observations are analytical questions, not automatic cleaning targets. -->

---

# Describe Before Explaining

A good first description might be:

> "X and Y have a moderately strong negative, approximately linear relationship, with one unusual observation."

Notice what is missing:

.big[
**No causal explanation.**
]

First describe what the evidence shows.

Then ask why.

---

# Can One Number Help?

A scatterplot contains a lot of information.

Sometimes we want a compact numerical summary.

For two continuous variables, one common summary is:

.huge[
**correlation**
]

---

# Correlation

The correlation coefficient is usually written:

.huge[
**r**
]

It ranges from:

.big[
**-1 to +1**
]

Its sign summarizes **direction**.

Its magnitude summarizes the **strength of a linear relationship**.

---

# Reading Correlation

.big[
**r > 0** → positive linear relationship

**r < 0** → negative linear relationship

**r ≈ 0** → little **linear** relationship

`|r|` closer to 1 → stronger linear relationship
]

The word **linear** matters.

---

# Calculate It

```{r}
df %>%
  summarise(
    correlation = cor(
      adm_rate,
      md_earn_wne_p6,
      use = "complete.obs"
    )
  )
```

.big[
The number summarizes part of what we saw in the scatterplot.
]

<!-- Instructor note: Students need to interpret cor(), not memorize its syntax. -->

---

# What Correlation Does Well

Correlation gives us a compact answer to:

.big[
> **How strongly and in what direction do two continuous variables move together linearly?**
]

Useful.

But incomplete.

---

# What Correlation Throws Away

A single number cannot show:

- the shape of the relationship
- unusual observations
- clusters
- subgroup structure
- the substantive meaning of X and Y

.big[
**Summaries compress information.**
]

Compression always loses something.

<!-- LO1 + LO4: Numerical summaries should not replace inspection of evidence. -->

---

# A Relationship With Low Correlation

```{r, echo=FALSE}
demo_curved %>%
  ggplot(aes(x, y)) +
  geom_point()
```

```{r, echo=FALSE}
cor(demo_curved$x, demo_curved$y)
```

.big[
A low correlation does not imply "nothing is happening."
]

The relationship may simply not be linear.

---

# Outliers Can Move r

Compare:

```{r, echo=FALSE}
cor(demo_positive$x, demo_positive$y)
```

with:

```{r, echo=FALSE}
cor(demo_outlier$x, demo_outlier$y)
```

.big[
One unusual observation can substantially change a numerical summary.
]

---

# Plot First

.huge[
**Never interpret correlation without looking at the data.**
]

The graph shows the structure.

The correlation summarizes one aspect of it.

Use both.

---

# AI Loves Summaries

Ask an AI assistant:

> "The correlation between X and Y is -0.62. Interpret it."

It can produce fluent prose instantly.

But without seeing the data, neither you nor the AI knows whether:

- the relationship is curved
- one outlier drives it
- important groups are hidden

<!-- LO5: AI can verbalize a statistic but cannot recover information omitted from the prompt. -->

---

# Audit the Interpretation

If AI says:

> "There is a strong negative relationship between X and Y."

Ask:

.big[
**Linear?**

**What does the scatterplot show?**

**Are there unusual observations?**

**What do X and Y actually measure?**
]

A number is not a complete analysis.

---

# What About Groups?

Not every relationship is continuous + continuous.

Suppose we ask:

> **Do public and private colleges differ in earnings?**

Now:

**X:** categorical

**Y:** continuous

---

# Compare Distributions

For categorical + continuous relationships, describe:

.big[
**Typical values**

**Spread**

**Overlap**

**Unusual observations**
]

The question is:

> How does the distribution of Y differ across groups of X?

---

# Group Differences

Suppose one group has a higher median earnings level.

That supports:

> "Typical earnings are higher in this group in these data."

It does **not** automatically support:

> "Being in this group causes higher earnings."

.big[
Description first. Explanation later.
]

---

# Overlap Matters

Imagine two groups with different typical values but substantial overlap.

.big[
A difference in groups does not mean every member of one group exceeds every member of the other.
]

Good communication describes both:

- the difference
- the variation

<!-- LO6: Avoid deterministic interpretation of group differences. -->

---

# Averages Can Hide Variation

Suppose:

> Group A average = $60,000

> Group B average = $50,000

That tells us something.

But not:

- how spread out each group is
- whether distributions overlap
- whether outliers drive the means

.big[
**A summary is not the distribution.**
]

---

# The Same Principle

For continuous + continuous:

> correlation is not the scatterplot.

For categorical + continuous:

> the group mean is not the group distribution.

.huge[
**Never let the summary replace the evidence.**
]

---

# Practice

Suppose a scatterplot shows:

- Y generally decreases as X increases
- the points follow an approximately straight-line pattern
- observations are fairly tightly clustered around that pattern
- one point sits far from the others

A good description is:

.big[
**strong, negative, approximately linear relationship with one unusual observation**
]

---

# Practice With r

Suppose:

**r = 0.08**

What can we safely say?

.big[
There is little **linear** association between X and Y.
]

What can we not safely say?

> There is no relationship of any kind.

---

# Practice With Groups

Suppose median earnings are higher at private colleges than public colleges, but the distributions overlap substantially.

A calibrated description:

.big[
> "Private colleges have higher typical earnings in these data, although there is substantial variation and overlap between the groups."
]

---

# Return to the Claim

We began with:

> **More selective colleges produce graduates who earn more.**

After today's analysis, we might be able to say:

.big[
> "Selectivity and graduate earnings are associated in the College Scorecard data."
]

We are still not entitled to the word:

.huge[
**produce**
]

---

# Why Not?

A relationship tells us:

> X and Y move together.

It does not yet tell us:

> **why** they move together.

Maybe X affects Y.

Maybe Y affects X.

Maybe something else affects both.

That is our next problem.

---

# Next Class

.huge[
**What Else Could Explain It?**
]

We move from:

> **What relationship do I see?**

to:

> **Why might I see that relationship?**

---

# Before You Leave

For a continuous–continuous relationship, ask:

1. What is the direction?
2. What is the form?
3. How strong is the pattern?
4. Are there unusual observations?
5. What does correlation add?
6. What might correlation hide?

For group comparisons, ask:

1. How do typical values differ?
2. How much variation is there?
3. How much overlap is there?
4. Are unusual observations important?

---

# The Rule to Remember

.huge[
**Describe before you explain.**
]

.big[
And never let a numerical summary replace the data it summarizes.
]

---

# Homework 6

You will practice:

- describing direction, form, strength, and unusual observations;
- calculating and interpreting correlation;
- comparing correlations with scatterplots;
- identifying information a numerical summary hides;
- describing group differences without overclaiming.

Next:

.big[
**What Else Could Explain It?**
]
