---
title: "Lecture 3"
output: html_document
date: "2026-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Univariate visualization

- after fighting over the *CONCEPT* of a variable, we now need to **LOOK** at it

# Loading the data

```{r}
require(tidyverse)
library(tidyverse)

df <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/refs/heads/main/data/sc_debt.Rds")

```

# Start by just LOOKING at the variables

```{r}
df %>%
  select(adm_rate,md_earn_wne_p6)
```

# Now visualize them with a plot

- New `R` function called: `ggplot()`
- `ggplot()` uses plus signs (`+`) instead of pipes (`%>%`)

```{r}
df %>%
  ggplot(aes(x = adm_rate)) +
  geom_histogram() + 
  labs(x = "College admissions rate",
       y = "Number of colleges",
       title = "Distribution of admissions rate",
       subtitle = "US colleges surveyed in 2021")
```

```{r}
df %>%
  ggplot(aes(x = md_earn_wne_p6)) +
  geom_histogram() +
  labs(x = "Graduate Earnings",
       y= "Frequency",
       title = "Distribution of Graduate Earnings")
```

# Categorical visualization vs continuous

- For continuous (numeric) variables, use `geom_histogram()` (or `geom_density()`)
- For categorical variables, use `geom_bar()`

```{r}
df %>%
  ggplot(aes(y = stabbr)) + 
  geom_bar()
```




