---
title: "DS-1000 Homework 2"
subtitle: "From Questions to Data"
author: "DS-1000: How Data Shape Our World"
date: "Due before Lecture 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
```

# What is this homework for?

In Homework 1, you started with a factual claim and asked what you would need to observe to evaluate it.

Now we have data.

The goal of Homework 2 is to practice translating a substantive question into a dataset:

> **What does one row represent? Which variables measure the concepts we care about? Which observations and variables belong in our analysis?**

You do **not** need to memorize `R` syntax. You may use your notes, the internet, classmates, and AI. When AI writes code for you, your job is to decide whether the code actually does what you intended.

# Part I: Return to our claim

We will continue with the claim from Homework 1:

> **More selective colleges produce graduates who earn more money.**

Load the tidyverse and College Scorecard data:

```{r load-data}
library(tidyverse)

df <- read_rds("../data/sc_debt.Rds")
```

Take a quick look at the data:

```{r inspect-data}
glimpse(df)
```

Some relevant variables are:

| Variable | Meaning |
|---|---|
| `instnm` | Institution name |
| `stabbr` | State abbreviation |
| `preddeg` | Predominant degree offered |
| `adm_rate` | Proportion of applications accepted |
| `sat_avg` | Average SAT score |
| `md_earn_wne_p6` | Average earnings of recent graduates |
| `grad_debt_mdn` | Median debt of graduates |
| `ugds` | Number of undergraduates |

## 1. What does one row represent?

Complete the sentence:

> **One row in this dataset represents ________________________________.**

**Your answer:**



## 2. Concepts versus variables

Our claim contains two important concepts: **college selectivity** and **graduate earnings**.

Which variable would you use to operationalize each concept?

**Selectivity:**



**Graduate earnings:**



For each choice, explain briefly what the variable *actually measures*. Is that exactly the same thing as the broader concept?

**Your answer:**



# Part II: Build the data we need

Suppose our immediate goal is:

> Keep bachelor's-degree institutions with non-missing admissions-rate and earnings data. Retain only the institution name, state, admissions rate, and earnings variables.

This is an analytical instruction written for a human. We now need to translate it into instructions R can execute.

## 3. Ask AI to translate the instructions

Use ChatGPT, Claude, or another AI assistant. Give it the analytical instruction above and ask it to write tidyverse code using `filter()` and `select()`.

Paste the code it gives you into the chunk below and run it.

```{r ai-code}
# Paste the AI-generated code here.

```

If the code produces an error, use AI, your notes, or classmates to debug it until it runs.

## 4. Explain what the code does

In plain English, explain what each major step of the code is doing.

Do not merely translate syntax word-for-word. Explain the analytical decision.

**Your answer:**



## 5. Verify the output

Do not assume that code is correct because it runs.

Use the following commands, or other checks you think are useful, to inspect the dataset produced by the AI code.

```{r verify-output}
# Replace college_analysis below if your AI gave the new dataset a different name.

# glimpse(college_analysis)
# nrow(college_analysis)
# names(college_analysis)
```

Based on your inspection:

**Does the resulting dataset contain the variables you intended to keep? How do you know?**



**Does it appear to contain the observations you intended to keep? How do you know?**



# Part III: Audit some plausible code

An AI assistant proposes the following solution:

```{r flawed-code, eval=FALSE}
college_analysis <- df %>%
  filter(adm_rate > 0) %>%
  select(instnm, adm_rate, md_earn_wne_p6)
```

The code is valid R. It will run.

## 6. Does running mean correct?

Compare this code to the original instruction:

> Keep bachelor's-degree institutions with non-missing admissions-rate and earnings data. Retain only the institution name, state, admissions rate, and earnings variables.

Identify **at least two ways** the proposed code fails to implement the instruction correctly.

**Your answer:**



This distinction will recur throughout DS-1000:

> **Code that runs is not necessarily code that answers the question you asked.**

# Part IV: Transfer to a new claim

Consider a second claim:

> **Students at more expensive colleges graduate with more debt.**

Do not analyze the relationship yet.

## 7. Decompose the claim

**What are the two main concepts?**



**Which variables in the College Scorecard data could operationalize those concepts?**



**What does each chosen variable actually measure?**



**What is the unit of analysis?**



## 8. Specify the analytical dataset

In plain English, describe which observations and variables you would retain to begin investigating this claim.

Do **not** write R code. Write instructions precise enough that another person—or an AI assistant—could translate them into code without guessing what you mean.

**Your answer:**



# Looking ahead

At this point, we have learned how to move from:

> **Question → concepts → variables → unit of analysis → analytical dataset**

But we are still not ready to put our two variables together.

Before asking whether two variables are related, we should understand each variable on its own.

## 9. What should we inspect first?

Suppose you are about to analyze admissions rates and graduate earnings.

Before examining their relationship, what would you want to know about **each variable individually**?

List two or three things you would want to inspect.

**Your answer:**



Lecture 3 will introduce tools for doing exactly this.

# Before you submit

- [ ] I can explain what one row in the College Scorecard data represents.
- [ ] I can distinguish a concept from a variable used to measure it.
- [ ] I can specify which observations and variables an analysis requires.
- [ ] I used AI to translate an analytical instruction into R code.
- [ ] I checked the resulting data rather than assuming the code was correct.
- [ ] I can explain why code that runs may still be analytically wrong.

## Connection to Lab Quiz 1

You will not be expected to memorize complicated R syntax. You should be able to read simple tidyverse code and determine whether it implements a stated analytical task.

You should also be prepared to independently identify:

1. the unit of analysis;
2. concepts and their operationalizations;
3. the observations and variables needed for a question; and
4. whether a proposed piece of code actually creates the intended analytical dataset.
