---
title: "What Does Confidence Actually Justify?"
subtitle: "Homework 8"
author: "DS-1000: How Data Shape Our World"
date: "Due before Lecture 11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)

gms <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/game_summary.Rds")
```

# What is this homework for?

In Lecture 10, you saw that a single confidence percentage can hide real differences between subgroups — a pooled bootstrap looked strong even though the pattern's strength varied by season.

Now you will run that same check yourself, on a new comparison:

> **A confidence percentage describes resampling of one dataset. Before trusting it, check whether that dataset pools together groups that might behave differently.**

You may use AI, your notes, classmates, and the internet. AI can help you adapt the `for()` loop. Your job is to decide what the resulting numbers do and do not justify.

# The Claim

> **Teams that shoot well from the free-throw line win more often.**

The relevant variables:

| Variable | Meaning |
|---|---|
| `ft_80` | Team scored more than 80% of its free throws in that game |
| `isWin` | Whether the team won the game |
| `yearSeason` | Which season the game was played in |

# Part I: The Pooled Comparison

## 1. Look First

Calculate the win rate for `ft_80 == TRUE` versus `ft_80 == FALSE`, pooling all seasons together.

```{r observed-comparison}
# Calculate the win rate by ft_80 here.

```

**What do you observe?**



# Part II: Pooled Bootstrap

## 2. Build the Loop

Write a `for()` loop that resamples the full (pooled) dataset with replacement at least 200 times, and for each resample calculates the win rate for each `ft_80` group.

```{r pooled-bootstrap}
set.seed(123)

# bs_pooled <- NULL
# for (i in 1:200) {
#   ...
# }

```

## 3. Turn It Into a Confidence Number

What fraction of resamples show `ft_80 == TRUE` teams winning more often than `ft_80 == FALSE` teams?

```{r pooled-confidence}
# Calculate the pooled confidence percentage here.

```

**Your pooled confidence percentage:**



# Part III: Does It Hold Within Every Season?

## 4. Resample Within Seasons

Adapt your loop so that it resamples **within each season** (`group_by(yearSeason)` before `sample_n()`), the way Lecture 10 did for turnovers. Then calculate a separate confidence percentage for each season.

```{r season-bootstrap}
set.seed(123)

# bs_season <- NULL
# for (i in 1:200) {
#   ...
# }

```

```{r season-confidence}
# Summarise confidence by yearSeason here.

```

**What do you find?**



## 5. Compare

Is the pattern equally strong in every season, or does it vary? Does any season look meaningfully different from the pooled result?

**Your answer:**



# Part IV: What Does the Number Justify?

## 6. Write Two Sentences

Write one sentence describing what your pooled confidence percentage **does** justify.

**Your answer:**



Write one sentence describing something it does **not** justify.

**Your answer:**



# Part V: Audit an AI Explanation

Suppose an AI assistant is shown only your pooled confidence number and says:

> **"This proves that any team can guarantee more wins by improving free-throw shooting, in any season, including next year's."**

## 7. Audit the Claim

Identify two problems with this statement, using what you found in Part III.

**Problem 1:**



**Problem 2:**



# Part VI: Calibrated Communication

## 8. Write the Conclusion

Write two or three sentences for a non-statistical audience. Your conclusion should:

- report the pooled confidence percentage and what comparison it describes;
- mention whether the pattern held up consistently across seasons;
- avoid claiming that the result guarantees anything about a future or unobserved season.

**Your answer:**



# Before Lecture 11

Complete these statements:

> **A confidence percentage only describes ________________________________.**



> **Before trusting a pooled confidence number, I should check ________________________________.**



> **If a pattern's strength varies a lot across subgroups, that tells me ________________________________.**



> **A confidence percentage never justifies a claim about ________________________________.**

