---
title: "How Should Classification Performance Be Evaluated?"
subtitle: "Homework 12"
author: "DS-1000: How Data Shape Our World"
date: "Due before Lecture 16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)

gms <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/game_summary.Rds")

sweep_thresholds <- function(probs, truth, thresholds = seq(0, 1, by = 0.02)) {
  out <- NULL
  for (t in thresholds) {
    pred_t <- ifelse(probs >= t, TRUE, FALSE)
    tp_t <- sum(pred_t == TRUE & truth == TRUE)
    fn_t <- sum(pred_t == FALSE & truth == TRUE)
    tn_t <- sum(pred_t == FALSE & truth == FALSE)
    fp_t <- sum(pred_t == TRUE & truth == FALSE)
    out <- bind_rows(out, tibble(
      threshold = t,
      sensitivity = tp_t / (tp_t + fn_t),
      specificity = tn_t / (tn_t + fp_t)
    ))
  }
  out
}

compute_auc <- function(roc_dat) {
  roc_sorted <- roc_dat %>%
    mutate(fpr = 1 - specificity) %>%
    drop_na(sensitivity, fpr) %>%
    arrange(fpr)
  sum(diff(roc_sorted$fpr) * (head(roc_sorted$sensitivity, -1) +
                                 tail(roc_sorted$sensitivity, -1)) / 2)
}
```

# What is this homework for?

In Lecture 15, you saw that one accuracy number depends on a threshold you have to choose, and that sensitivity, specificity, ROC curves, and AUC give a fuller picture. The `sweep_thresholds()` and `compute_auc()` functions above are provided so you can focus on using them correctly rather than re-deriving them.

Now you will use these tools to compare two classifiers:

> **A model comparison should hold up across every threshold, not just one. AUC lets you check that.**

You may use AI, your notes, classmates, and the internet. Your job is to interpret the sensitivity/specificity tradeoff and the AUC comparison correctly.

# The Claim

> **Adding free-throw percentage to a shooting-based classifier improves its ability to predict wins.**

The relevant variables:

| Variable | Meaning |
|---|---|
| `isWin` | Whether the team won the game |
| `pctFG` | Field-goal percentage |
| `pctFT` | Free-throw percentage |

# Part I: Build the Two Models

## 1. Split the Data

```{r train-test-split}
set.seed(123)

# Create train and test sets here, roughly half and half.

```

## 2. Fit Both Models

Fit `isWin ~ pctFG` (simple) and `isWin ~ pctFG + pctFT` (augmented) on the training set.

```{r fit-models}
# Fit both models here.

```

# Part II: Sensitivity and Specificity at One Threshold

## 3. Classify at 0.5

For the **augmented** model, predict on the test set and classify at a 0.5 threshold.

```{r classify}
# Predict and classify here.

```

## 4. Calculate Sensitivity and Specificity

```{r sens-spec}
# Build the confusion matrix and calculate sensitivity and specificity.

```

**Sensitivity:**

**Specificity:**

**Are they close together, or far apart? What does that tell you?**



# Part III: Sweep the Threshold

## 5. Use `sweep_thresholds()`

Apply the provided `sweep_thresholds()` function to the augmented model's predicted probabilities.

```{r sweep}
# Call sweep_thresholds() on your augmented model's predictions.

```

## 6. Plot the Tradeoff

Plot sensitivity and specificity against threshold, the way Lecture 15 did.

```{r plot-tradeoff}
# Create the plot here.

```

**Describe what happens to each metric as the threshold rises.**



# Part IV: ROC and AUC

## 7. Plot the ROC Curve

```{r roc-curve}
# Plot 1 - specificity (x) against sensitivity (y).

```

## 8. Calculate AUC for Both Models

Use `compute_auc()` on the threshold-swept results for **both** the simple and augmented models.

```{r auc-both}
# Calculate AUC for the simple model and the augmented model.

```

**Simple model AUC:**

**Augmented model AUC:**

# Part V: Cross-Validate the Comparison

## 9. Repeat Across Many Splits

Adapt Lecture 15's repeated cross-validation loop to compare the average AUC of the two models across at least 50 train/test splits.

```{r cv-auc}
set.seed(123)

# cv_auc <- NULL
# for (i in 1:50) {
#   ...
# }

```

**Which model has the higher average cross-validated AUC?**



# Part VI: Audit an AI Explanation

Suppose an AI assistant is shown only the augmented model's 0.5-threshold accuracy and says:

> **"This single number proves the augmented model is better."**

## 10. Audit the Claim

Identify two problems with this statement, using what you found in Parts IV and V.

**Problem 1:**



**Problem 2:**



# Part VII: Calibrated Communication

## 11. Write the Conclusion

Write two or three sentences for a non-statistical audience. Your conclusion should:

- report the cross-validated AUC comparison between the two models;
- avoid claiming the better model establishes what causes winning;
- mention that any real use would still require choosing a threshold.

**Your answer:**



# Before Lecture 16

Complete these statements:

> **Sensitivity and specificity separate ________________________________ from ________________________________.**



> **An ROC curve shows ________________________________.**



> **AUC is useful for comparing models because ________________________________.**



> **A higher AUC does not, by itself, establish ________________________________.**

