---
title: "How Can a Model Summarize a Relationship?"
subtitle: "Homework 9"
author: "DS-1000: How Data Shape Our World"
date: "Due before Lecture 12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)

debt <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/sc_debt.Rds") %>%
  filter(ugds > 0, md_earn_wne_p6 > 0)
```

# What is this homework for?

In Lecture 11, you saw how `lm()` finds a line of best fit and how to interpret its intercept and slope.

Now you will fit and interpret a regression model yourself, on a claim the lecture didn't cover:

> **A line is only a summary. Fitting it is the easy part — interpreting what it does and doesn't say is the point.**

You may use AI, your notes, classmates, and the internet. AI can help you write the `lm()` call. Your job is to interpret the output correctly and decide what it does and does not establish.

# The Claim

> **Larger colleges (measured by undergraduate enrollment) have lower graduate earnings.**

The relevant variables:

| Variable | Meaning |
|---|---|
| `ugds` | Number of undergraduate students |
| `md_earn_wne_p6` | Median earnings of graduates |

# Part I: Visualize First

## 1. Look at the Relationship

```{r visualize}
# Create a scatterplot of ugds (x) and md_earn_wne_p6 (y), with a
# line of best fit added.

```

**What does the scatterplot suggest about the direction of the relationship, if any?**



# Part II: Fit the Model

## 2. Run the Regression

```{r fit-model}
# Fit a linear model with md_earn_wne_p6 as the outcome and ugds
# as the predictor. Save it to an object and view it with summary().

```

## 3. Report the Coefficients

**Intercept:**



**Slope:**



# Part III: Interpret in Context

## 4. What Does the Intercept Mean?

In your own words, what does the intercept represent here? Is `ugds = 0` a value that could describe a real college?

**Your answer:**



## 5. What Does the Slope Mean?

In your own words, what does the slope represent? Use the actual units (students, dollars) in your explanation.

**Your answer:**



## 6. Does the Sign Match the Claim?

Does the sign of the slope match the direction of the claim above? Explain.

**Your answer:**



# Part IV: What This Model Does Not Establish

## 7. Plausible Confounders

Name one variable that could be related to both college size and graduate earnings, and that might explain part of this slope without college size *causing* the earnings difference.

**Your answer:**



# Part V: Audit an AI Explanation

Suppose an AI assistant sees only your regression output and says:

> **"This proves that shrinking a college's enrollment would cause its graduates to earn more."**

## 8. Audit the Claim

Identify two problems with this statement.

**Problem 1:**



**Problem 2:**



# Part VI: Calibrated Communication

## 9. Write the Conclusion

Write two or three sentences for a non-statistical audience. Your conclusion should:

- report the slope's direction and approximate size, in context;
- avoid describing the slope as a causal effect;
- mention the confounder you identified in Part IV.

**Your answer:**



# Before Lecture 12

Complete these statements:

> **The intercept of a regression line tells me ________________________________.**



> **The slope of a regression line tells me ________________________________.**



> **A regression line's fit does not, by itself, tell me ________________________________.**



> **Before treating a slope as causal, I should ________________________________.**

