---
title: "Putting Variables Together"
subtitle: "Test 2 Practice (ungraded)"
author: "Prof. Bisbee"
institute: "Vanderbilt University"
date: "Due before Test 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
```

## Purpose

This homework is a rehearsal for **Unit Test 2**.

You may use your notes, classmates, the internet, and AI tools. The point is not to memorize R syntax. The point is to practice the reasoning process that you will need to perform independently on the test:

> **Choose the comparison → See the relationship → Describe the relationship → Ask what else could explain it → Calibrate the conclusion**

For this homework, you will actually use R and AI to help implement your decisions. On Test 2, you will instead be given analytical artifacts (plots, tables, summaries) and asked to evaluate them.

## The Data

We will use a synthetic dataset built to resemble a typical college admissions office's records. This is not real student data — using real admissions records here would raise privacy concerns — but it reflects the kinds of relationships found in real datasets.

```{r}
library(tidyverse)
library(scales)

ad <- read_rds("https://github.com/jbisbee1/DS1000_F2024/raw/main/data/admit_data.rds")
```

Every row is an **admitted** student. Some variables you may find useful:

| Variable | Definition |
|---|---|
| `income` | Family income (AGI) |
| `sat` | SAT/ACT score (ACT converted to SAT scale) |
| `gpa` | High school GPA, four-point scale |
| `visit` | Did the student visit campus? |
| `legacy` | Did a parent of the student attend this college? |
| `net_price` | Tuition minus aid received |
| `yield` | Did the admitted student enroll in the fall? |

Start by inspecting the data.

```{r}
glimpse(ad)
```

## The Claim

An admissions officer says:

> **"Getting admitted students to visit campus increases the chance they enroll."**

Your job is **not** to determine whether this claim is true yet. Your job is to determine what the evidence does and does not support.

## 1. Decompose the Claim

What are the two main concepts in the claim?

**Your answer:**



Which variables operationalize those concepts, and what type is each (categorical or continuous)?

**Your answer:**



## 2. Choose the Right Comparison

Both `visit` and `yield` are binary (yes/no) variables. Which of the following is the appropriate way to visualize their relationship — a scatterplot, a boxplot, or a grouped bar chart / proportion table? Explain why, referencing variable type.

**Your answer:**



## 3. See the Relationship

Create the visualization or table you chose in Question 2.

```{r}
# Visualize or tabulate the relationship between visit and yield here.

```

Describe what you see. Does yield appear to differ by visit status, and by roughly how much?

**Your answer:**



## 4. Correlation — Does It Apply Here?

Calculate the correlation between `visit` and `yield` directly.

```{r}
# Calculate cor(ad$visit, ad$yield) here.

```

Is a raw correlation coefficient the most natural or interpretable way to describe the relationship between two binary variables? What does the group-difference-in-proportions from Question 3 communicate that a bare correlation number does not?

**Your answer:**



## 5. What Else Could Explain It?

Suppose visiting students really do enroll at a higher rate. Before concluding that visiting **causes** higher enrollment, identify a plausible confounder — a third variable that could affect both whether a student visits and whether they enroll.

**Your answer:**



Explain, in one or two sentences, how that variable could produce the association even if visiting itself had no effect.

**Your answer:**



## 6. Test the Confound

Choose one variable from the dataset that could operationalize your proposed confounder (for example, `legacy`). Examine the visit-yield relationship **within** subgroups defined by that variable.

```{r}
# Examine the visit-yield relationship within subgroups of your chosen confounder.

```

Does the relationship between visit and yield look similar across subgroups, or does it change? What does that tell you?

**Your answer:**



## 7. Audit an AI Explanation

Suppose an AI assistant is shown only your Question 3 result and says:

> **"This proves that campus visits cause higher enrollment, so the admissions office should require all admitted students to visit."**

Identify two problems with this statement.

**Problem 1:**



**Problem 2:**



## 8. Calibrated Conclusion

Write two or three sentences a non-statistical audience (e.g., an admissions director) could understand. Your conclusion should:

- describe the observed association between visiting and enrolling;
- mention the confounder you tested and whether it changed your interpretation;
- avoid claiming that visiting causes enrollment.

**Your answer:**



## 9. Are We Ready?

You now have:

- a defined empirical claim and its operationalizations;
- the correct comparison for two binary variables;
- a described relationship;
- a tested confounder;
- a calibrated conclusion.

What is the **next analytical question** you would need to answer to strengthen this analysis further (for example, involving uncertainty, a different confounder, or a different subgroup)?

Do not perform that analysis yet.

**Your answer:**



## Before Test 2

You should now be able to do the following without AI:

- choose an appropriate multivariate visualization based on variable types;
- describe a relationship in terms of direction, form, strength, or group differences;
- interpret a correlation and recognize when it is (or isn't) the most natural summary;
- identify a plausible confounder for an observed association;
- reason about whether a relationship holds up within subgroups;
- audit an AI-generated causal claim;
- communicate a calibrated conclusion that distinguishes association from causation.

The specific admissions context will **not** be on Test 2. The test will ask you to transfer this reasoning to a new empirical problem.
