---
title: "DS-1000 Homework 5"
subtitle: "Variables Together"
author: "DS-1000: How Data Shape Our World"
date: "Due before Lecture 6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)
library(tidyverse)
library(scales)

polls <- read_rds("https://github.com/jbisbee1/DS1000_F2024/raw/main/data/Pres2020_PV.Rds") %>%
  mutate(
    Trump = Trump / 100,
    Biden = Biden / 100
  )
```

# What is this homework for?

In Unit 1, you learned to understand variables **one at a time**.

Lecture 5 added the next step:

> **How should we put X and Y together so that the comparison required by the question is visible?**

The central rule is:

- continuous + continuous → scatterplot;
- categorical + continuous → compare distributions across groups;
- categorical + categorical → compare counts or proportions.

You may use your notes, classmates, the internet, and AI. You are not being asked to memorize `ggplot2` syntax. Your responsibility is to decide **what comparison the question requires**, determine whether the visualization is appropriate, and verify that generated code actually implements that decision.

This homework uses polling data from the 2020 U.S. presidential election. Each row represents a poll.

# Part I: Start With the Question

Consider the claim:

> **Polls conducted using different interview methods produced different estimates of Trump's support.**

Some variables that may be useful are:

| Variable | Meaning |
|---|---|
| `Trump` | Poll estimate of support for Donald Trump |
| `Biden` | Poll estimate of support for Joe Biden |
| `Mode` | Method used to contact/interview respondents |
| `SampleSize` | Number of respondents in the poll |
| `RepCertVote` | Certified Republican vote share |
| `DemCertVote` | Certified Democratic vote share |

## 1. Identify X and Y

For the claim above:

**What is X?**



**What type of variable is X?**



**What is Y?**



**What type of variable is Y?**



## 2. Choose the Comparison

Given those variable types, what kind of visualization should we use first?

Explain why it makes the comparison in the claim visible.

**Your answer:**



# Part II: Ask AI to Implement the Choice

## 3. Generate the Visualization

Ask an AI assistant to write `ggplot2` code that compares the distribution of Trump support across polling modes.

Tell it that:

- `Mode` is categorical;
- `Trump` is continuous;
- you want to compare the distribution of Trump support across polling modes;
- unreadable or clearly invalid mode categories may be removed if necessary, but it should explain any exclusions.

Paste and run the code below.

```{r mode-trump}
# Paste the AI-generated code here.

```

## 4. Audit the Visualization

Do not assume that the visualization is appropriate because the code runs.

Answer each question.

**Does the visualization match the variable types? Why?**



**Does it actually make differences across polling modes visible?**



**Did the AI filter, transform, reorder, or otherwise change the data? If so, was that choice justified?**



**What does one observation represented in this visualization correspond to?**



# Part III: Same Data, Different Question

Now consider:

> **Polls with larger samples were more accurate in estimating Trump's vote share.**

We can define polling error as the difference between the poll's estimate of Trump support and Trump's certified vote share.

```{r create-error}
polls <- polls %>%
  mutate(TrumpError = Trump - RepCertVote / 100)
```

## 5. Identify the Variables

For this new question:

**What is X, and what type of variable is it?**



**What is Y, and what type of variable is it?**



**What visualization follows from those variable types?**



## 6. Generate and Check the Plot

Ask AI to generate an appropriate visualization of the relationship between sample size and Trump polling error.

Paste and run the code below.

```{r sample-error}
# Paste the AI-generated code here.

```

Before interpreting it, check:

**Does the plot use the variables on sensible axes?**



**Does one mark correspond to one poll?**



**Did AI make any filtering or transformation decisions that you did not request?**



# Part IV: A Technically Valid but Unhelpful Plot

Suppose another analyst wants to evaluate:

> **Polls conducted using different interview methods produced different estimates of Trump's support.**

They make this plot:

```{r wrong-question}
polls %>%
  ggplot(aes(x = Trump)) +
  geom_histogram()
```

## 7. Audit the Analysis

The code is valid and the histogram is an appropriate **univariate** visualization of `Trump`.

Why does it nevertheless fail to answer the research question?

**Your answer:**



Describe the comparison that the visualization needs to make visible instead.

**Your answer:**



# Part V: Categorical + Categorical

Suppose we create a simple variable indicating whether each poll overestimated or underestimated Trump's certified vote share.

```{r error-direction}
polls <- polls %>%
  mutate(
    ErrorDirection = ifelse(
      TrumpError > 0,
      "Overestimated Trump",
      "Underestimated Trump"
    )
  )
```

Now ask:

> **Was the direction of polling error different across interview modes?**

## 8. Choose the Representation

What type of variable is `Mode`?



What type of variable is `ErrorDirection`?



What kind of comparison is appropriate for these two variables?

**Your answer:**



Ask AI to generate a visualization that compares **proportions** rather than only raw counts across polling modes.

```{r categorical-categorical}
# Paste the AI-generated code here.

```

Why might proportions be more informative than raw counts in this case?

**Your answer:**



# Part VI: What Can We Say?

## 9. Describe, Don't Explain

Choose **one** of the multivariate visualizations you created above.

In two or three sentences, describe the visible pattern in substantive language.

At this stage, do **not** explain why the relationship exists and do **not** make a causal claim.

**Your answer:**



## 10. The Rule

Complete this statement in your own words:

> **Do not choose the graph first. Instead, ________________________________.**

**Your answer:**



# Before Quiz 4

You should be able to do the following without AI:

- identify X and Y from a substantive question;
- determine whether each variable is continuous or categorical;
- choose a visualization that matches the variable types and the comparison;
- recognize when a technically valid graph fails to answer the research question;
- explain what one observation or mark represents;
- evaluate whether AI-generated visualization code actually implements the intended comparison;
- describe the visible pattern without claiming why it exists.

You do **not** need to memorize long `ggplot2` commands.

Lecture 6 will ask the next question:

> **Once we can see a relationship, how should we describe and summarize it?**
