---
title: "How Confident Should We Be?"
subtitle: "Homework 7"
author: "DS-1000: How Data Shape Our World"
date: "Due before Lecture 10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)

nba <- read_rds("https://github.com/jbisbee1/DS1000_F2026/raw/main/data/nba_players_2018.Rds")
```

# What is this homework for?

In Lecture 9, you saw that any single sample gives one story, and that resampling (the **bootstrap**) shows how fragile that story is.

Now you will run the same procedure on a new comparison, in your own hands:

> **We observe a difference between two groups. How confident should we be that it isn't just an artifact of this one sample?**

You may use AI, your notes, classmates, and the internet. AI can help you write the `for()` loop. Your job is to decide what the resulting confidence number does and does not tell you.

# The Claim

> **Non-rookies shoot free throws more accurately than rookies.**

The relevant variables:

| Variable | Meaning |
|---|---|
| `isRookie` | Whether the player is a rookie |
| `pctFT` | Free-throw percentage |
| `fta` | Free throws attempted |

# Part I: The Observed Comparison

## 1. Look First

Calculate the average free-throw percentage for rookies and non-rookies.

```{r observed-comparison}
# Calculate the group means here.

```

**What do you observe?**



# Part II: What Would "Confidence" Mean Here?

## 2. Before You Resample

In your own words, what would it mean to be "80% confident" that non-rookies shoot free throws more accurately?

**Your answer:**



What would it mean if that confidence were close to 50%?

**Your answer:**



# Part III: Build the Bootstrap

## 3. One Resample

Draw one resample of the full dataset (with replacement) and recompute the group means.

```{r one-resample}
set.seed(123)

# Resample nba with replacement (same size as nba) and calculate
# mean pctFT by isRookie.

```

## 4. Many Resamples

Write a `for()` loop that repeats Step 3 at least 200 times, saving each resample's group means. Ask AI for help if you get stuck, but make sure you can explain every line.

```{r bootstrap-loop}
set.seed(123)

# bs_ft <- NULL
# for (i in 1:200) {
#   ...
# }

```

## 5. Turn It Into a Confidence Number

For each resample, was the non-rookie mean higher than the rookie mean? What fraction of resamples say yes?

```{r confidence-number}
# Calculate the confidence percentage here.

```

**Your confidence percentage:**



# Part IV: Interpret the Number

## 6. What It Does and Doesn't Mean

Write one sentence explaining what your confidence percentage actually measures.

**Your answer:**



Write one sentence describing a wrong way to interpret it (something it does *not* establish).

**Your answer:**



# Part V: Is the Comparison Fair?

## 7. A Plausible Confound

Free-throw percentage can be noisy for players who attempted very few free throws all season.

```{r attempts-check}
# Compare fta (free throws attempted) between rookies and non-rookies here.

```

Does attempt volume differ meaningfully between the two groups? Could that affect how much you trust `pctFT` as a comparison?

**Your answer:**



## 8. A Fairer Version (Optional Challenge)

If you have time, filter to players with a reasonable minimum number of free-throw attempts (for example, `fta >= 20`) and redo the bootstrap loop from Part III.

```{r fair-bootstrap}
# Optional: repeat the bootstrap on a filtered sample.

```

Did your confidence number change? Why might that happen?

**Your answer:**



# Part VI: Audit an AI Explanation

Suppose an AI assistant sees only your Part III confidence number and says:

> **"This proves non-rookies are better free-throw shooters."**

## 9. Audit the Claim

Identify two problems with this statement.

**Problem 1:**



**Problem 2:**



# Part VII: Calibrated Communication

## 10. Write the Conclusion

Write two or three sentences for a non-statistical audience. Your conclusion should:

- report your confidence percentage and what comparison it is about;
- avoid describing the confidence number as a probability that the claim is "true";
- mention anything from Part V that should make a reader more cautious.

**Your answer:**



# Before Lecture 10

Complete these statements:

> **Bootstrap resampling helps me because ________________________________.**



> **A confidence percentage near 50% tells me ________________________________.**



> **A high confidence percentage still does not tell me ________________________________.**



> **Before trusting a bootstrap result, I should check ________________________________.**


