---
title: "From Claim to Evidence"
subtitle: "Homework 4"
author: "Prof. Bisbee"
institute: "Vanderbilt University"
date: "Due before Test 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
```

## Purpose

This homework is a rehearsal for **Unit Test 1**.

You may use your notes, classmates, the internet, and AI tools. The point is not to memorize R syntax. The point is to practice the reasoning process that you will need to perform independently on the test:

> **Claim → concepts → variables → unit of analysis → understand each variable → prepare the evidence → verify**

For this homework, you will actually use R and AI to help implement your decisions. On Test 1, you will instead be given analytical artifacts and asked to evaluate them.

## The Data

We will use data on NBA players who were active during the 2018–19 season. These are adapted from a dataset used in a previous version of DS-1000.

```{r}
library(tidyverse)

nba <- read_rds("https://github.com/jbisbee1/DS1000_F2024/raw/main/data/nba_players_2018.Rds")
```

Some variables that may be useful are:

| Variable | Definition |
|---|---|
| `namePlayer` | Player name |
| `slugTeam` | Team abbreviation |
| `gp` | Games played |
| `gs` | Games started |
| `fgm` | Field goals made |
| `fga` | Field goals attempted |
| `pctFG` | Field-goal percentage |
| `pctFT` | Free-throw percentage |
| `agePlayer` | Player age |
| `minutes` | Minutes played |
| `pts` | Total points |
| `isRookie` | Whether the player is a rookie |
| `numberPlayerSeason` | Which NBA season this is for the player |

Start by inspecting the data.

```{r}
glimpse(nba)
```

## The Claim

An NBA commentator says:

> **"More experienced NBA players are more productive scorers."**

Your job is **not** to determine whether this claim is true yet.

Your job is to determine whether these data can provide trustworthy evidence for evaluating it.

## 1. Decompose the Claim

What are the two main concepts in the claim?

**Your answer:**



Which variables in the dataset could operationalize those concepts?

**Your answer:**



For each variable you chose, explain briefly what it actually measures and one way it may be an imperfect measure of the broader concept.

**Your answer:**



## 2. Identify the Unit of Analysis

Complete the sentence:

> **One row in this dataset represents ________________________________.**

Why does knowing this matter for evaluating the claim?

**Your answer:**



## 3. Understand the Experience Variable

Before putting variables together, inspect your proposed measure of **experience** on its own.

What type of variable is it?

**Your answer:**



Use an appropriate univariate visualization. You may write the code yourself or ask an AI assistant to generate it.

```{r}
# Create an appropriate univariate visualization here.

```

Describe the distribution. Address at least two useful features such as typical values, spread, shape, unusual values, bounds, or missingness.

**Your answer:**



Does anything about the distribution make you reconsider whether this variable is a good operationalization of "experience"?

**Your answer:**



## 4. Understand the Scoring Variable

Now inspect your proposed measure of **scoring productivity** on its own.

What type of variable is it?

**Your answer:**



Create an appropriate univariate visualization.

```{r}
# Create an appropriate univariate visualization here.

```

Describe the distribution and identify anything that might matter before using this variable as evidence.

**Your answer:**



Now reconsider the measurement itself.

For example, would `pts` measure **productivity**, **opportunity to play**, or some combination of both? Would points per game or points per minute represent the concept differently?

You do not need to change your variable. Explain the tradeoff.

**Your answer:**



## 5. Investigate Missingness

Use R to determine whether either of your two proposed variables has missing values.

You may ask AI to help write the code.

```{r}
# Check missingness here.

```

Report what you found.

**Your answer:**



If observations were missing, would you automatically delete them? Explain what you would want to know before making that decision.

**Your answer:**



## 6. Define the Analytical Sample

Suppose we decide that a player should have played at least one game to belong in the analysis.

In plain English, write a precise specification for the analytical dataset you want R to create.

Your specification should state:

- which observations should remain;
- which variables should remain;
- how missingness should be handled;
- whether any transformation is needed.

**Do not write R code yet.**

**Your specification:**



## 7. Let AI Implement Your Specification

Give your specification from Question 6 to ChatGPT, Claude, or another AI assistant.

Ask it to write tidyverse code that creates the analytical dataset.

Paste the code below and run it.

```{r}
# Paste and run the AI-generated code here.

```

## 8. Verify the Result

Do not assume the pipeline is correct because it runs.

Use commands such as `glimpse()`, `nrow()`, `names()`, `summary()`, or other checks that you think are useful.

```{r}
# Verify the resulting analytical dataset here.

```

Answer each question:

**Did the intended observations remain? What evidence tells you that?**



**Did the intended variables remain? What evidence tells you that?**



**Were missing values handled as you specified? What evidence tells you that?**



**Did the AI make any consequential decision that you did not explicitly request?**



## 9. Audit Another Analyst's Pipeline

Another analyst proposes:

```{r, eval=FALSE}
analysis_df <- nba %>%
  filter(numberPlayerSeason > 1,
         pts > 0) %>%
  mutate(pts = pts * 100) %>%
  select(namePlayer, numberPlayerSeason, pts)
```

The code will run.

Identify **at least two analytical problems** with this pipeline.

For each problem, explain why it matters for evaluating the original claim.

**Your answer:**



## 10. Are We Ready?

You now have:

- a defined empirical claim;
- operationalizations of its concepts;
- a known unit of analysis;
- inspected univariate distributions;
- a defined analytical sample;
- and a verified preparation pipeline.

What is the **next analytical question** we would need to answer in order to evaluate the original claim?

Do not perform that analysis yet.

**Your answer:**



## Before Test 1

You should now be able to do the following without AI:

- identify the concepts in an empirical claim;
- distinguish a concept from its operationalization;
- identify the unit of analysis;
- choose an appropriate univariate visualization;
- interpret important features of a distribution;
- reason about missingness and sample restrictions;
- explain why data-preparation choices matter;
- read simple R code and determine whether it implements an analytical instruction;
- distinguish code that **runs** from code that is **analytically appropriate**;
- explain what evidence would be needed next without claiming more than the current analysis supports.

The specific NBA context will **not** be on Test 1. The test will ask you to transfer this reasoning to a new empirical problem.
